### 2025-05-15 draft.tex
new: add tex file

% Created 2025-05-14 Wed 12:29
% Intended LaTeX compiler: pdflatex
\documentclass{article}

\input{style/header.tex}
\usepackage[inkscapelatex=false]{svg}
\date{}
\title{}
\hypersetup{
 pdfauthor={Alan Munoz},
 pdftitle={},
 pdfkeywords={},
 pdfsubject={},
 pdfcreator={Emacs 30.1 (Org mode 9.7.11)}, 
 pdflang={English}}
\usepackage{natbib}
\begin{document}

\twocolumn[
\icmltitle{cp\_measure: Morphological features for bioimaging}

% It is OKAY to include author information, even for blind
% submissions: the style file will automatically remove it for you
% unless you've provided the [accepted] option to the icml2025
% package.

% List of affiliations: The first argument should be a (short)
% identifier you will use later to specify author affiliations
% Academic affiliations should list Department, University, City, Region, Country
% Industry affiliations should list Company, City, Region, Country

% You can specify symbols, otherwise they are numbered in order.
% Ideally, you should not use this facility. Affiliations will be numbered
% in order of appearance and this is the preferred way.
\icmlsetsymbol{equal}{*}

\begin{icmlauthorlist}
\icmlauthor{Alan F. Munoz}{broad}
\icmlauthor{Tim Treis}{hh,broad}
\icmlauthor{Alexandr A. Kalinin}{broad}
\icmlauthor{Shatavisha Dasgupta}{broad}
\icmlauthor{Fabian Theis}{hh}
\icmlauthor{Shantanu Singh}{broad}
\icmlauthor{Anne E. Carpenter}{broad}
\end{icmlauthorlist}

\icmlaffiliation{broad}{Broad Institute of MIT and Harvard, United States}
\icmlaffiliation{hh}{Institute of Computational biology, Helmholtz Zentrum München, Germany}

\icmlcorrespondingauthor{Shantanu Singh}{shantanu@broadinstitute.org}

% You may provide any keywords that you
% find helpful for describing your paper; these are used to populate
% the "keywords" metadata in the PDF but will not be shown in the document
\icmlkeywords{Machine Learning, ICML}

\vskip 0.3in
]

% this must go after the closing bracket ] following \twocolumn[ ...

% This command actually creates the footnote in the first column
% listing the affiliations and the copyright notice.
% The command takes one argument, which is text to display at the start of the footnote.
% The \icmlEqualContribution command is standard text for equal contribution.
% Remove it (just {}) if you do not need this facility.

\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
% \printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.

\begin{abstract}
Quantifying the contents of objects in images is a common challenge in biological imaging. The most widely used software to acomplish require manual adjustments. Here we introduce our library cp\_measure, which provides programmatic access to the most widespread metrics to convert images and objects into features. We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives. Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing both the accessibility and reproducibility of the field.
\end{abstract}
\section{Introduction}
\label{sec:org6430daa}
The field of biological imaging leverages data from cell microscopy images to obtain biological insights. Through microscopy techniques we can obtain information from a multitude of sources, such as illumination, fluorescent dyes or even proteins tagged with other proteins that fluoresce, acting as a beacon and providing information on quantity and/or localization.
The main task at the core bioimage analysis is identifying features that convey information of the state of a cell or its contents. This entails combining identified objects with pixel data to quantify their morphology and other information sources, such as intensity distributions.

Morphological profiling a subfield of bioimaging that focuses on the cells and the shapes and properties of their organelles. It is usually supported by both statistical methods and machine and deep learning approaches. One of its biggest applications is drug discovery, where it is used primarily due to its low acquisition cost and capabilities for scaling to screen many compounds efficiently \citep{sealDecadeSystematicReview2024}. It can also function as a complement to other high throughput techniques, such as genomics.
\subsection{The current state of bioimage analysis}
\label{sec:org8f5b33d}
The most widely-used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}. Its target users are biologists with limited programmatic expertise. The accessibility that manually implemented workflows reduces its suitability for automated analyses and the composition of more complex pipelines. In this context we define accessibility as the lack of friction to use the software in computational pipelines.

The usage of CellProfiler entails a higher degree of friction for composite pipelines that orchestrate multiple tools. Its focus on manually-defined workflows adjusted for specific use-cases makes it less suitable for many high throughput analyses. Additionally, CellProfiler has a significant number of dependencies. It depends on libraries from both the Python and Java languages, thus increasing the number of things that may go wrong when setting up or integrating with other tools. Tough containers partially alleviate this problem, they carry additional complexity and caveats.

A small subset of the features in CellProfiler is provided by scikit-image, but this is greatly limited and cannot be reliably compared to the existing body of work that used CellProfiler to obtain the measurements. The other big problem of CellProfiler is its dependency on human interaction at many steps of the analysis, opening the door to human mistakes and hindering analyses reproducibility.

A CellProfiler build pipeline requires cannot easily integrate with others. Its existing plug-in system is inaccessible for most users and has strong limitations. Building automated pipelines that contain CellProfiler and integrating tools within its components are time and effort-consuming challenges.

In this space there is a vacuum for a software library that covers the generalist and high throughput use-cases for which CellProfiler is inadequate. Distributed CellProfiler would cover this case if not for its cloud-only approach and dependency on preconfigured pipelines, as well as its limited debugging capabilities \citep{mcquinCellProfiler30Nextgeneration2018}.

Other recent software has been published aiming to fill this unoccupied niche, but their solutions are unsuitable for practical usage at scale: First, their features have completely independent implementations from the original ones, thus are not guaranteed to reproduce existing results and the biological interpretations may be different from the current field standard. Additionally, their implementations are not modular nor scale well upon increases to the data size. Some of them, akin to Distributed CellProfiler, are based on main use-case is cloud services.
\subsection{Engineered features in the era of deep learning}
\label{sec:org9dc3dfa}
The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data. Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}.

Unlike with deep learning-based features, whose interpretation requires complex methods such as counterfactuals or attention maps, most of the engineered features have a mathematical definition that is vastly simpler to translate into biological insights.

Engineered features can also outperform deep learning ones under some conditions \citep{tangMorphologicalProfilingDrug2024}. Though deep learning features are more likely to improve in the coming years

It also has a higher barrier of entry, since most models require annotating data for training and testing. Even new foundational models do not generalize well for out-of-distribution data.

The main target users of such a tool would be biologists in need of automation and reproducibility, as well as computer and data scientists that want to dive into the field of biological imaging.

As a whole, there is still a role to be played for engineered features. The community is missing a generalist tool that converts images and masks into vectors in an automated fashion and portable across fields. It serves as complementary approach deep learning methods.
\section{Extraction of interpretable features using cp\_measure}
\label{sec:org5633614}
With cp\_measure we recapitulated all features used in the vast majority of experiments: One object and one imaging channel (e.g., intensity), one object and multiple channels (e.g., Manders correlation) and multiple objects (e.g., number of neighbors).

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
\caption{\label{fig:orgea7e72a}Extraction of measurements based on a collective of objects. cp\_measure generates features from images by using the local information in a mask-defined region of interest. Its produces measurements for every object, and is usually used to acquire information on the pairwise combination of all the available channels.}
\end{figure}

The cp\_measure's code base was based on CellProfiler's, but we removed the sections of the code that are unnecessary for programmatic usage, thus separating the actual functional components from the orchestration logic. It improves reusability, testability and its makes it possible for more people to access and improve its internals in the future. It aims to be a stepping stone to increased extensibility and maintainability.

Our library aims to remain consistent with the current scientific python ecosystem. The main interface matches that of the widely-used scikit-image \citep{waltScikitimageImageProcessing2014}. This greatly reduces the effort needed to integrate it in existing workflows and tools.

By isolating and cleaning the implemented mathematics of CellProfiler we aim to make the measurement methods more accessible to computer and data scientists, while still leveraging the decade and a half of existing features that are present in numerous datasets, maintaining a higher degree of compatibility. To retain this compatibility in the long term requires contributing this changes back into CellProfiler, be it directly or as a dependency.

When comparing the equivalent pipelines, cp\_measure's minimal overhead results in faster processing times than CellProfiler (TODO add benchmark).

We demonstrate the correlation of cp\_measure compared to a comparable CellProfiler pipeline using a representative subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}. Then we demonstrate cases in which cp\_measure can help identify biological phenotypes: first using 3D images of astrocytes and then using spatial transcriptomics dataset. These use-cases demonstrate its widespread applicability. 
\subsection{Our features match analogous CellProfiler measurements}
\label{sec:org37e389d}

Our first aim was to validate the features from cp\_measure compared to the CellProfiler standard ones. We used images from the JUMP dataset, containing more than 130,000 chemical and genetic perturbations. We first curated a set of genetic perturbations that, based on existing analyses, contained cells with distinctive morphologies to all others based on the CellProfiler features. Using this curated dataset, we segmented the cytosol and nuclei of the cells using a CellProfiler pipeline to obtain both segmentation masks and the original morphological features.

Next, we used the segmentation masks to calculate the cp\_measure features. Then we mapped the features from CellProfiler to cp\_measure and compared the values in both. Figure \ref{fig:org7fe862f} shows this comparison. Then we demonstrate that we recapitulate the CellProfiler features.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/astro3d}
\caption{\label{fig:org7fe862f}{[}Placeholder] Comparison of CellProfiler and cp\_measure.}
\end{figure}
\subsection{Results and examples of usage}
\label{sec:org8ef524e}
\subsubsection{Classification of astrocytes and their distinctive features}
\label{sec:org3d97e6f}

As a proof of concept, we showcased that cp\_measure can be used in a workflow to train a classifier for cell populations. We used it to process 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}. We then calculated the median value for every feature in a cell and the number of cells, following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}. Then we trained a Gradient Boosting classifier to identify which day. With this we identified which features distinguish cells on the later samples and distinguish subpopulations. It is worth noting that there will be some redundancy in the information contained in the cp\_measure features, and thus during cases in which multiple features inform on the same groups of cells Shapley values is likely to only capture one of them.

\begin{figure}[htbp]
\centering
\includesvg[width=.9\linewidth]{./figs/shap}
\caption{\label{fig:org4620d4d}Shapley values of most important features to classify the day in which an image was taken (a multi-class classification task). The test data accuracy is shown in bold. Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis.}
\end{figure}
\subsubsection{Applicability on spatial transcriptomics}
\label{sec:org1d0d8b8}
\begin{figure}[htbp]
\centering
\includegraphics[width=.9\linewidth]{./figs/spatial.png}
\caption{\label{fig:org722dcf5}{[}PLACEHOLDER] Spatial omics analysis.}
\end{figure}
\section{Discussion}
\label{sec:orgf37b369}
The usage of image analysis pipelines that require manual setups hinders reproducibility and hinders our ability to compare different datasets. In this work we introduced our new library cp\_measure, which provides widely used engineered features and enables simpler automated analyses of microscopy data in either short scripts and complex pipelines. This also removes the requirement of using graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy data without the need of cloud-based infrastructure.

The biologically interpretable features provided by cp\_measure complement deep learning ones and offer a better mechanistic understanding of the underlying biology. When used in tandem with generalist tools it enables more insightful pipelines that leverage machine and deep learning approaches.

These measurements have already been used in non-biological contexts, such as environmental monitoring \citep{ideharaExploringNileRed2025}, thus these engineered metrics also benefit other scientific fields beyond morphological profiling. In general, we see the decomposition of CellProfiler into modular components as a way to facili
\section{Future work}
\label{sec:org5cdbb12}
There are multiple paths to improve and expand the functionality of cp\_measure. The first and most obvious is to integrate its measurements back to CellProfiler library. This would ensure that the results from pipelines built with either tool will be comparable in the future, while also providing the opportunity of formalizing the programmatic interface --- inputs and outputs --- of measurements.

Developing a comprehensive tests suite would guarantee mathematical correctness under the possible edge cases that may be encountered when dealing with new data. This test suite in turn would in turn open the door to further speed-ups in multiple ways: Firstly, optimizing the measurements that consume the most time, starting with object granularity (\textasciitilde{}80\% of the time). Additionally, it is possible to implement measurements using numba for just-in-time compiling and/or adding GPU support \citep{lamNumbaLLVMbasedPython2015}.

We plan to add other quality-of-life features. First, add a wrapper to call measurements on the fly, analogous to scikit-image's regionprops function. Secondly, a list of essential measurements for use-cases where speed is more important than using all the features. By lowering the barrier of effort required to integrate cp\_measure into existing pipelines these

Long-term, we envision cp\_measure can be the place to develop and distribute new measurements. While CellProfiler's measurements are widely used in bioimaging studies, the existing palette of measurements could be further extended to cover novel use-cases brought upon by novel developments in imaging acquisition devices and methods. Working with the community to further the number of measurements to better match the current questions scientists pose to imaging data.
\section{Methods}
\label{sec:org91ddb3a}
\subsection{Data and software}
\label{sec:orgf44b7cd}
The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}. All code to reproduce the analyses and figures, alongside links to the original data, is available on the Github repository \url{https://github.com/afermg/2025\_cpmeasure/}. The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.

\bibliographystyle{icml2025}
\bibliography{bibliography}
\end{document}

### 2025-05-15 draft.tex
docs: Update draft text

@@
-% Created 2025-05-14 Wed 12:29
+% Created 2025-05-15 Thu 13:57
 % Intended LaTeX compiler: pdflatex
@@
 \icmlauthor{Fabian Theis}{hh}
+\icmlauthor{Anne E. Carpenter}{broad}
 \icmlauthor{Shantanu Singh}{broad}
-\icmlauthor{Anne E. Carpenter}{broad}
 \end{icmlauthorlist}
@@
 Quantifying the contents of objects in images is a common challenge in biological imaging.
-The most widely used software to acomplish require manual adjustments.
+The most widely used software to do so require significant manual intervention.
 Here we introduce our library cp\_measure, which provides programmatic access to the most widespread metrics to convert images and objects into features.
@@
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
-Through microscopy techniques we can obtain information from a multitude of sources, such as illumination, fluorescent dyes or even proteins tagged with other proteins that fluoresce, acting as a beacon and providing information on quantity and/or localization.
+Microscopy techniques allows us to acquire information from a multitude of sources, such as illumination, fluorescent dyes or even proteins tagged with other proteins that fluoresce, acting as a beacon and providing information on quantity and/or localization.
+
 The main task at the core bioimage analysis is identifying features that convey information of the state of a cell or its contents.
@@
 
-A CellProfiler build pipeline requires cannot easily integrate with others.
+A CellProfiler pipeline is does not integrate well with other tools.
 Its existing plug-in system is inaccessible for most users and has strong limitations.
@@
 
-As a proof of concept, we showcased that cp\_measure can be used in a workflow to train a classifier for cell populations.
+As a demonstration of its ease of use, we used cp\_measure for featurization in a cell classification workflow.
 We used it to process 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}.
@@
 With this we identified which features distinguish cells on the later samples and distinguish subpopulations.
-It is worth noting that there will be some redundancy in the information contained in the cp\_measure features, and thus during cases in which multiple features inform on the same groups of cells Shapley values is likely to only capture one of them.
+It is worth noting that there will be some redundancy in the information contained in the cp\_measure features, and thus during cases in which multiple features inform on similar data subsets.
 
@@
 
-We plan to add other quality-of-life features.
-First, add a wrapper to call measurements on the fly, analogous to scikit-image's regionprops function.
+There is further space for improvement.
+First, provide a wrapper for all features that masp to scikit-image's regionprops as close as possible.
 Secondly, a list of essential measurements for use-cases where speed is more important than using all the features.

### 2025-05-15 src/astro3d/get_figures.py
feat(astro3d): add script to produce figs

@@ -1,0 +1,213 @@
+#!/usr/bin/env python
+from itertools import chain
+from pathlib import Path
+
+import matplotlib.pyplot as plt
+import numpy as np
+import polars as pl
+import polars.selectors as cs
+import seaborn as sns
+import xgboost
+from cp_measure.bulk import get_core_measurements
+from joblib import Parallel, delayed
+from pooch import Untar, retrieve
+from skimage.io import imread
+from sklearn.model_selection import train_test_split
+from xgboost import XGBClassifier
+
+MEASUREMENTS = get_core_measurements()
+
+overwrite = False
+profiles_fpath = Path("tmp.parquet")
+figs_path = Path(".") / ".." / ".." / "figs"
+figs_path.parent.mkdir(parents=True, exist_ok=True)
+target_feature = "day"
+
+cell_count_col = "Cell Count"
+seed = 42
+
+
+def fix_non_continuous_labels(labels_redz: np.ndarray) -> np.ndarray:
+    """
+    Enforce continuous labels from 1 to `n_unique_labels`.
+
+    Some measurements assume that labels continuous in 1->n.
+    This function removes any gaps (e.g., [0,1,3,4] -> [0,1,2,3]).
+    """
+    max_val = labels_redz.max()
+    uniq = np.unique(labels_redz)
+    missing_ix = np.setdiff1d(
+        np.arange(max_val),
+        uniq,
+        assume_unique=True,
+    )
+    if len(missing_ix):
+        new_ranges = np.arange(1, len(uniq))
+        for i, ix in enumerate(uniq[1:]):
+            labels_redz[labels_redz == ix] = new_ranges[i]
+
+    return labels_redz
+
+
+def apply_measurements(mask_path: str, img_path: str) -> pl.DataFrame:
+    meta = Path(img_path).parent.name.split("_")[:2]
+    meta = (*meta, Path(img_path).stem)
+
+    labels = imread(mask_path)
+    labels_redz = labels.max(axis=0)
+    # Cover case where the reduction on z removes an entire item
+    # Unlike in other cases, here we assume that the input mask can have gaps
+    labels_redz = fix_non_continuous_labels(labels_redz)
+
+    img = imread(img_path)
+    flat_img = img.max(axis=0)
[…]

### 2025-05-16 draft.tex
fix typo
docs: Update draft with spatial transcriptomics analysis

@@
-% Created 2025-05-15 Thu 13:57
+% Created 2025-05-16 Fri 08:47
 % Intended LaTeX compiler: pdflatex
@@
 We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives.
-Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing both the accessibility and reproducibility of the field.
+Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing developer accessibility and reproducibility of the pipelines.
 \end{abstract}
@@
 One of its biggest applications is drug discovery, where it is used primarily due to its low acquisition cost and capabilities for scaling to screen many compounds efficiently \citep{sealDecadeSystematicReview2024}.
-It can also function as a complement to other high throughput techniques, such as genomics.
+It can also complement to other high throughput methods, such as genomic sequencing.
 \subsection{The current state of bioimage analysis}
@@
 Engineered features can also outperform deep learning ones under some conditions \citep{tangMorphologicalProfilingDrug2024}.
-Though deep learning features are more likely to improve in the coming years
+Even if deep learning features improve over the coming years, they will most likely require more complex interpretability methods, if they are interpretable at all.
 
@@
 These use-cases demonstrate its widespread applicability.
-\subsection{Our features match analogous CellProfiler measurements}
-\label{sec:org37e389d}
+\subsection{Our features match CellProfiler standard measurements}
+\label{sec:org04a138e}
 
@@
 \label{sec:org1d0d8b8}
+A key advantage of providing these CellProfiler-like measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
+The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality.
+However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment.
+In contrast, classical morphological measurements yield explicit, interpretable readouts—for instance, the co-localization of fluorescent markers—facilitating clear biological interpretations.
+
+To demonstrate this utility, we integrated our cp$\backslash$\_measure-based feature extraction into the widely used spatial analysis library Squidpy [CITE].
+Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData [CITE] framework underlying Squidpy.
+Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings—up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts.
+Leveraging the modular design of cp$\backslash$\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores.
+This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.
+
+To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data.
+This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
+We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform [CITE, \url{https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/}].
+Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
+Morphological features were extracted from these 5-channel images for both datasets.
+Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy [CITE] and PyCytoMiner [CITE], respectively.
+We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig.
+XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data.
+Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type.
+Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included.
+Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp$\backslash$\_measure beyond its traditional scope.
+
 \begin{figure}[htbp]

### 2025-05-16 draft.tex
Updates from Overleaf

@@
-% Created 2025-05-16 Fri 08:47
+% Created 2025-05-15 Thu 14:07
 % Intended LaTeX compiler: pdflatex
@@
 We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives.
-Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing developer accessibility and reproducibility of the pipelines.
+Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing both the accessibility and reproducibility of the field.
 \end{abstract}
@@
 One of its biggest applications is drug discovery, where it is used primarily due to its low acquisition cost and capabilities for scaling to screen many compounds efficiently \citep{sealDecadeSystematicReview2024}.
-It can also complement to other high throughput methods, such as genomic sequencing.
+It can also function as a complement to other high throughput techniques, such as genomics.
 \subsection{The current state of bioimage analysis}
@@
 Engineered features can also outperform deep learning ones under some conditions \citep{tangMorphologicalProfilingDrug2024}.
-Even if deep learning features improve over the coming years, they will most likely require more complex interpretability methods, if they are interpretable at all.
+Though deep learning features are more likely to improve in the coming years
 
@@
 
-To demonstrate this utility, we integrated our cp$\backslash$\_measure-based feature extraction into the widely used spatial analysis library Squidpy [CITE].
+To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy [CITE].
 Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData [CITE] framework underlying Squidpy.
 Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings—up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts.
-Leveraging the modular design of cp$\backslash$\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores.
+Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores.
 This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.
@@
 This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
-We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform [CITE, \url{https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/}].
+We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform [CITE, https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/].
 Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
@@
 Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included.
-Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp$\backslash$\_measure beyond its traditional scope.
-
+Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.
 \begin{figure}[htbp]

### 2025-05-16 src/jump/compare_cps.py
[WIP]add measurement script

@@ -19,21 +19,24 @@ from joblib import Parallel, delayed
 from skimage.io import imread
 from tqdm import tqdm
 
-img_dir = Path("/datastore/alan/cp_measure/jump_subset")
-out_dir = Path("/datastore/alan/cp_measure/jump_masks")
-out_dir.mkdir(exist_ok=True, parents=True)
-img_dir.mkdir(exist_ok=True, parents=True)
+output_dir = Path("/datastore/alan/cp_measure/")
+img_dir = output_dir / "jump_subset"
+mask_dir = output_dir / "jump_masks"
+profiles_dir = output_dir / "profiles"
+for folder in (img_dir, mask_dir, results_dir):
+    folder.mkdir(exist_ok=True, parents=True)
 
 
 MEASUREMENTS = get_core_measurements()
 
 
-def get_keys(fpath: Path, n: int = 2) -> tuple[str, str]:
+def get_keys(fpath: Path, n: int = 2) -> tuple[str]:
     return tuple(fpath.stem.split("_")[:n])
 
 
 def segment_save(mask: np.ndarray, name: str) -> str:
-    fpath = out_dir / f"{name}.npz"
+    fpath = out_dir / f"shape_{mask.shape[1]}" / f"{name}.npz"
+    fpath.parent.mkdir(parents=True, exist_ok=True)
     np.savez(fpath, mask)
     return fpath
 
@@ -71,35 +74,29 @@ segmentable_files = [
 ]
 
 
-def apply_measurements(mask_path: Path, img_path: Path) -> pl.DataFrame:
-    # meta = Path(img_path).parent.name.split("_")[:2]
-    # meta = (*meta, Path(img_path).stem)
-
-    if mask_path.endswith("npz"):
+def read_labels(mask_path: Path):
+    if str(mask_path).endswith("npz"):
         labels = np.load(mask_path)["arr_0"]
     else:
         labels = imread(mask_path)
-    # if gaps:
-    #     labels = labels.max(axis=0)
-    # Cover case where the reduction on z removes an entire item
-    # Unlike in other cases, here we assume that the input mask can have gaps
-    # labels = fix_non_continuous_labels(labels_redz)
+    return labels
+
+
+def apply_measurements(mask_path: Path, img_path: Path) -> pl.DataFrame:
+    gene, site, channel = img_path.stem.split("_")[:3]
+    labels = read_labels(mask_path)
 
     img = imread(img_path)
     if img.ndim == 3:
[…]

### 2025-05-16 src/jump/explore_cellprof_data.py
feat: Add CellProfiler data comparison script

@@ -1,0 +1,152 @@
+"""
+Compare Cellprofiler and cp_measure data.
+Plot the comparison.
+"""
+
+from pathlib import Path
+
+import duckdb
+import matplotlib.pyplot as plt
+import polars as pl
+import pooch
+from duckdb.typing import VARCHAR
+from parse_features import get_feature_groups
+from polars import selectors as cs
+from pooch import Unzip
+from skimage.io import imread
+
+figs_dir = Path("../../") / "figs"
+
+
+# Turns out it is easier to trim down the CellProfiler measurements
+# into cp_measure than anything else
+def trim_features(name: str) -> str:
+    """
+    Trim object and channel information from CellProfiler
+    """
+    replacement_obj = ("Cells_", "Image_", "Nuclei_", "Cytoplasm_")
+    replacement_ch = [f"_Orig{x}" for x in ("DNA", "AGP", "ER", "Mito", "RNA")]
+    replacement_mea = ["AreaShape_"]
+
+    for obj in replacement_obj:
+        name = name.replace(obj, "")
+    for ch in replacement_ch:
+        name = name.replace(ch, "")
+    for mea in replacement_mea:
+        name = name.replace(mea, "")
+    return name
+
+
+con = duckdb.connect()
+con.create_function("trim_features", trim_features, [VARCHAR], VARCHAR)
+
+cp_data = pooch.retrieve(
+    "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
+    known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
+    processor=Unzip(),
+)
+
+real_files = [x for x in cp_data if not Path(x).name.startswith(".")]
+
+# %%
+# sample_img = imread(real_files[0])
+# plt.imshow(sample_img)
+# plt.savefig("delme.png", dpi=300)
+
+sql_files = [(Path(x).parent.parent.name, x) for x in cp_data if x.endswith("db")]
+con.sql("INSTALL sqlite;")
+con.sql("LOAD sqlite;")
+for name, sql_file in sql_files:
+    con.sql(f"ATTACH '{sql_file}' AS {name} (TYPE sqlite);")
+
+con.sql("SHOW DATABASES;")
+con.sql("SHOW TABLES;")
+# %%
+original_tables = []
+for name, _ in sql_files:
+    con.sql(f"USE {name}")
[…]

### 2025-05-16 src/jump/get_profiles.py

@@ -19,11 +19,11 @@ from joblib import Parallel, delayed
 from skimage.io import imread
 from tqdm import tqdm
 
-output_dir = Path("/datastore/alan/cp_measure/")
-img_dir = output_dir / "jump_subset"
-mask_dir = output_dir / "jump_masks"
-profiles_dir = output_dir / "profiles"
-for folder in (img_dir, mask_dir, results_dir):
+out_dir = Path("/datastore/alan/cp_measure/")
+img_dir = out_dir / "jump_subset"
+mask_dir = out_dir / "jump_masks"
+profiles_dir = out_dir / "profiles"
+for folder in (img_dir, mask_dir, profiles_dir):
     folder.mkdir(exist_ok=True, parents=True)
 
 
@@ -187,6 +187,17 @@ for mask, img in combs:
     assert (
         np.load(out_dir / img.parent.name / mask)["arr_0"].shape == imread(img).shape
     ), "error"
+measure_results = Parallel(n_jobs=-1)(
+    delayed(apply_measurements)(out_dir / img.parent.name / m, img)
+    for m, img in tqdm(combs[:10])
+)
+first_set = pl.concat(measure_results)
+first_set.write_parquet(profiles_dir / "first_set.parquet")
+# multichannel
+
+MEASUREMENTS_TYPE2 = get_correlation_measurements()
+
+# multimask
 # table = table.append_column(
 #     "object",
 #     pa.array([step_name.split("_")[-1]] * len(table), pa.string()),
@@ -196,14 +207,3 @@ for mask, img in combs:
 type2_pairs = [list(product((k, combinations(v, 2)))) for k, v in pairs]
 type2_pairs = [list(product((k, combinations(v, 2)))) for k, v in pairs]
 type2_pairs = [list(product((k, combinations(v, 2)))) for k, v in pairs]
-measure_results = Parallel(n_jobs=-1)(
-    delayed(apply_measurements)(out_dir / img.parent.name / m, img)
-    for m, img in tqdm(combs)
-)
-first_set = pl.concat(measure_results)
-first_set.write_parquet(profiles_dir / "first_set.parquet")
-# multichannel
-
-MEASUREMENTS_TYPE2 = get_correlation_measurements()
-
-# multimask

### 2025-05-16 src/jump/parse_features.py

@@ -1,0 +1,77 @@
+#!/usr/bin/env jupyter
+"""Parse feature names to divide it into its components."""
+
+import re
+from functools import cache
+
+import polars as pl
+
+
+def get_feature_groups(
+    feature_fullnames: tuple[str],
+    feature_names: tuple[str] = ("Compartment", "Feature", "Channel", "Suffix"),
+) -> pl.DataFrame:
+    """Group features in a consistent manner using a regex.
+
+    Parameters
+    ----------
+    feature_fullnames : tuple[str]
+        Tuple of full names of the features.
+    feature_names : tuple[str]
+        Tuple of names of the features to be extracted.
+
+    Returns
+    -------
+    pl.DataFrame
+        DataFrame containing the grouped feature information.
+
+    Notes
+    -----
+    The function supports two cases: Channel-based and Non-channel based shape.
+    It applies regular expressions, converts to format MASK,FEATURE,CHANNEL(opt),SUFFIX,
+    merging channels where necessary.
+
+    """
+    masks = "|".join(("Cells", "Nuclei", "Cytoplasm", "Image", "Object"))
+    channels = "|".join((
+        "DNA",
+        "AGP",
+        "RNA",
+        "ER",
+        "Mito",
+        "Image",
+    ))
+    chless_feats = "|".join((
+        "AreaShape",
+        "Neighbors",
+        "RadialDistribution",
+        "Location",
+        "Count",
+        "Number",
+        "Parent",
+        "Children",
+        "ObjectSkeleton",
+        "Threshold",
+    ))
+
+    std = re.compile(rf"({masks})_(\S+)_(Orig)?({channels})(_.*)?")
+    chless = re.compile(f"({masks})_?({chless_feats})_?([a-zA-Z]+)?(.*)?")
+
+    results = [(std.findall(x) or chless.findall(x))[0] for x in feature_fullnames]
+    results = [
+        (
+            (x[0], "".join(x[1:3]), "", x[3])
+            if len(x) < 5
+            else (*x[:2], "".join(x[2:4]), x[4])
+        )
+        for x in results
+    ]
+
+    # Select Mask, Feature and Channel features
+    feature_meta = pl.DataFrame(
+        [x for x in results],
+        schema=[(col, str) for col in feature_names],
+        orient="row",
+    )
+
+ee    return feature_meta.with_columns(fullname=pl.Series(feature_fullnames))

### 2025-05-16 draft.tex
docs: Update draft text

--- 2025-05-16 draft.tex
@@
-% Created 2025-05-16 Fri 08:47
+% Created 2025-05-16 Fri 08:56
 % Intended LaTeX compiler: pdflatex

### 2025-05-16 src/jump/explore_cellprof_data.py
feat: add scripts for profiling with provided masks

@@ -1,152 +1,0 @@
-"""
-Compare Cellprofiler and cp_measure data.
-Plot the comparison.
-"""
-
-from pathlib import Path
-
-import duckdb
-import matplotlib.pyplot as plt
-import polars as pl
-import pooch
-from duckdb.typing import VARCHAR
-from parse_features import get_feature_groups
-from polars import selectors as cs
-from pooch import Unzip
-from skimage.io import imread
-
-figs_dir = Path("../../") / "figs"
-
-
-# Turns out it is easier to trim down the CellProfiler measurements
-# into cp_measure than anything else
-def trim_features(name: str) -> str:
-    """
-    Trim object and channel information from CellProfiler
-    """
-    replacement_obj = ("Cells_", "Image_", "Nuclei_", "Cytoplasm_")
-    replacement_ch = [f"_Orig{x}" for x in ("DNA", "AGP", "ER", "Mito", "RNA")]
-    replacement_mea = ["AreaShape_"]
-
-    for obj in replacement_obj:
-        name = name.replace(obj, "")
-    for ch in replacement_ch:
-        name = name.replace(ch, "")
-    for mea in replacement_mea:
-        name = name.replace(mea, "")
-    return name
-
-
-con = duckdb.connect()
-con.create_function("trim_features", trim_features, [VARCHAR], VARCHAR)
-
-cp_data = pooch.retrieve(
-    "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
-    known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
-    processor=Unzip(),
-)
-
-real_files = [x for x in cp_data if not Path(x).name.startswith(".")]
-
-# %%
-# sample_img = imread(real_files[0])
-# plt.imshow(sample_img)
-# plt.savefig("delme.png", dpi=300)
-
-sql_files = [(Path(x).parent.parent.name, x) for x in cp_data if x.endswith("db")]
-con.sql("INSTALL sqlite;")
-con.sql("LOAD sqlite;")
-for name, sql_file in sql_files:
-    con.sql(f"ATTACH '{sql_file}' AS {name} (TYPE sqlite);")
-
-con.sql("SHOW DATABASES;")
-con.sql("SHOW TABLES;")
-# %%
-original_tables = []
-for name, _ in sql_files:
-    con.sql(f"USE {name}")
[…]

### 2025-05-16 src/jump/get_profiles_from_masks.py

@@ -1,0 +1,70 @@
+#!/usr/bin/env jupyter
+"""
+Compare CellProfiler and cp_measure metrics
+"""
+
+from concurrent.futures import ThreadPoolExecutor
+from itertools import chain, combinations, groupby, product
+from pathlib import Path
+
+import polars as pl
+import pooch
+from joblib import Parallel, delayed
+from pooch import Unzip
+from skimage.io import imread
+from tqdm import tqdm
+from utils import apply_measurements, get_keys, read_labels
+
+cp_data = pooch.retrieve(
+    "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
+    known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
+    processor=Unzip(),
+)
+mask_files = [Path(x) for x in cp_data if "masks" in x]
+# %%
+
+out_dir = Path("/datastore/alan/cp_measure/")
+img_dir = out_dir / "jump_subset"
+profiles_dir = out_dir / "profiles_via_masks"
+for folder in (img_dir, profiles_dir):
+    folder.mkdir(exist_ok=True, parents=True)
+
+
+channels = ["DNA", "AGP", "Mito", "RNA", "ER"]
+
+
+def try_imread(x):
+    try:
+        return imread(x)
+    except:
+        return str(x)
+
+
+img_paths = sorted(list(img_dir.rglob("*.tif")))
+
+
+# %%
+# Organize masks
+
+# We pair the mask names with their associated images
+
+sorted_mask_files = sorted(
+    mask_files, key=lambda x: "_".join(x.name.split("_")[:3]) + x.parent.name
+)
+grouped_masks = {k: list(v) for k, v in groupby(sorted_mask_files, get_keys)}
+pairs = [(grouped_masks[k], sorted(v)) for k, v in groupby(sorted(img_paths), get_keys)]
+
+# Check that all image sets are the same size
+assert len(set([len(x[1]) for x in pairs])) == 1, "Heterogeneous number of channels"
+
+# Get the product of all the mask-image combinations
+combs = list(chain(*[list(product(*xy)) for xy in pairs]))
+
+# Check that shapes match
+for mask, img in combs:
+    assert read_labels(mask).shape == imread(img).shape, "error"
+measure_results = Parallel(n_jobs=-1)(
+    delayed(apply_measurements)(m, img) for m, img in tqdm(combs)
+)
[…]

### 2025-05-17 src/astro3d/get_figures.py
refactor: Remove unused file and rename get_profiles

@@ -7,6 +7,7 @@ import numpy as np
 import polars as pl
 import polars.selectors as cs
 import seaborn as sns
+import shap
 import xgboost
 from cp_measure.bulk import get_core_measurements
 from joblib import Parallel, delayed
@@ -197,10 +198,11 @@ sns.violinplot(
 plt.gca().spines[["right", "top"]].set_visible(False)
 plt.savefig(figs_path / "astro3d.svg")
 
-import shap
 
 pred = model.predict(Xd, output_margin=True)
-explainer = shap.TreeExplainer(model, feature_names=data.columns)
+explainer = shap.TreeExplainer(
+    model, feature_names=[x.replace("_", " ") for x in data.columns]
+)
 explanation = explainer(Xd)
 shap_values = explanation.values
 # make sure the SHAP values add up to marginal predictions
@@ -211,3 +213,54 @@ plt.title("Impactful features")
 plt.text(0.4, -0.5, acc, fontweight="bold")
 plt.tight_layout()
 plt.savefig(figs_path / "shap.svg")
+
+# %% example figure
+axd = plt.figure(layout="constrained").subplot_mosaic(
+    """
+    AC
+    BB
+    """
+)
+i = 1
+img = imread(pairs[i][1]).max(axis=0)
+labels = imread(pairs[i][0]).max(axis=0)
+print(labels.max())
+axd["A"].imshow(img)
+axd["A"].axis("off")
+axd["A"].set_title("Z-projected image")
+axd["C"].imshow(labels)
+axd["C"].axis("off")
+axd["C"].set_title("Labels")
+shap.plots.beeswarm(explanation, max_display=6, ax=axd["B"], plot_size=None)
+axd["B"].set_yticklabels(axd["B"].get_yticklabels(), rotation=30)
+plt.text(0.15, -0.5, acc, fontweight="bold")
+# plt.tight_layout()
+plt.savefig(figs_path / "example_shap.svg")
+plt.close()
+# %%
+import matplotlib.pyplot as plt
+import numpy as np
+
+# prepare some coordinates
+x, y, z = np.indices((8, )
+
+# draw cuboids in the top left and bottom right corners, and a link between
+# them
+cube1 = (x < 3) & (y < 3) & (z < 3)
+cube2 = (x >= 5) & (y >= 5) & (z >= 5)
+link = abs(x - y) + abs(y - z) + abs(z - x) <= 2
+
+# combine the objects into a single boolean array
[…]

### 2025-05-17 src/jump/get_profiles_from_masks.py
refactor: Remove unused file and rename get_profiles

@@ -15,6 +15,11 @@ from skimage.io import imread
 from tqdm import tqdm
 from utils import apply_measurements, get_keys, read_labels
 
+
+def get_mask_name(x):
+    return x.parent.name.split("_")[0]
+
+
 cp_data = pooch.retrieve(
     "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
     known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
@@ -64,7 +69,7 @@ combs = list(chain(*[list(product(*xy)) for xy in pairs]))
 for mask, img in combs:
     assert read_labels(mask).shape == imread(img).shape, "error"
 measure_results = Parallel(n_jobs=-1)(
-    delayed(apply_measurements)(m, img) for m, img in tqdm(combs)
+    delayed(apply_measurements)(m, img, get_mask_name(m)) for m, img in tqdm(combs)
 )
 first_set = pl.concat(measure_results)
 first_set.write_parquet(profiles_dir / "first_set.parquet")

### 2025-05-17 src/jump/get_profiles_via_segmentation.py

@@ -19,11 +19,11 @@ from joblib import Parallel, delayed
 from skimage.io import imread
 from tqdm import tqdm
 
-out_dir = Path("/datastore/alan/cp_measure/")
-img_dir = out_dir / "jump_subset"
-mask_dir = out_dir / "jump_masks"
-profiles_dir = out_dir / "profiles"
-for folder in (img_dir, mask_dir, profiles_dir):
+output_dir = Path("/datastore/alan/cp_measure/")
+img_dir = output_dir / "jump_subset"
+mask_dir = output_dir / "jump_masks"
+profiles_dir = output_dir / "profiles"
+for folder in (img_dir, mask_dir, results_dir):
     folder.mkdir(exist_ok=True, parents=True)
 
 
@@ -187,17 +187,6 @@ for mask, img in combs:
     assert (
         np.load(out_dir / img.parent.name / mask)["arr_0"].shape == imread(img).shape
     ), "error"
-measure_results = Parallel(n_jobs=-1)(
-    delayed(apply_measurements)(out_dir / img.parent.name / m, img)
-    for m, img in tqdm(combs[:10])
-)
-first_set = pl.concat(measure_results)
-first_set.write_parquet(profiles_dir / "first_set.parquet")
-# multichannel
-
-MEASUREMENTS_TYPE2 = get_correlation_measurements()
-
-# multimask
 # table = table.append_column(
 #     "object",
 #     pa.array([step_name.split("_")[-1]] * len(table), pa.string()),
@@ -207,3 +196,14 @@ MEASUREMENTS_TYPE2 = get_correlation_measurements()
 type2_pairs = [list(product((k, combinations(v, 2)))) for k, v in pairs]
 type2_pairs = [list(product((k, combinations(v, 2)))) for k, v in pairs]
 type2_pairs = [list(product((k, combinations(v, 2)))) for k, v in pairs]
+measure_results = Parallel(n_jobs=-1)(
+    delayed(apply_measurements)(out_dir / img.parent.name / m, img)
+    for m, img in tqdm(combs)
+)
+first_set = pl.concat(measure_results)
+first_set.write_parquet(profiles_dir / "first_set.parquet")
+# multichannel
+
+MEASUREMENTS_TYPE2 = get_correlation_measurements()
+
+# multimask

### 2025-05-17 src/jump/parse_features.py

@@ -74,4 +74,4 @@ def get_feature_groups(
         orient="row",
     )
 
-ee    return feature_meta.with_columns(fullname=pl.Series(feature_fullnames))
+    return feature_meta.with_columns(fullname=pl.Series(feature_fullnames))

### 2025-05-17 draft.tex
feat: Add validation of cp_measure vs cprof

@@
-% Created 2025-05-16 Fri 08:56
+% Created 2025-05-17 Sat 12:54
 % Intended LaTeX compiler: pdflatex
@@
 \centering
-\includesvg[width=.9\linewidth]{./figs/astro3d}
+\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
 \caption{\label{fig:org7fe862f}{[}Placeholder] Comparison of CellProfiler and cp\_measure.}
@@
 \centering
-\includesvg[width=.9\linewidth]{./figs/shap}
+\includesvg[width=.9\linewidth]{./figs/example_shap}
 \caption{\label{fig:org4620d4d}Shapley values of most important features to classify the day in which an image was taken (a multi-class classification task).
@@
 \label{sec:org1d0d8b8}
-A key advantage of providing these CellProfiler-like measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
+A key advantage of providing these CellProfiler measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
 The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality.

### 2025-05-17 draft.tex
docs: Update draft text

@@
-% Created 2025-05-17 Sat 12:54
-% Intended LaTeX compiler: pdflatex
+% Created 2025-05-17 Sat 16:54
+% Intended LaTeX compiler: xelatex
 \documentclass{article}
@@
 \date{}
-\title{}
+\title{draft}
 \hypersetup{
- pdfauthor={Alan Munoz},
+ pdfauthor={},
  pdftitle={},
@@
  pdfsubject={},
- pdfcreator={Emacs 30.1 (Org mode 9.7.11)}, 
+ pdfcreator={},
  pdflang={English}}
@@
 \section{Introduction}
-\label{sec:org6430daa}
+\label{sec:orgbc2e426}
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
@@
 \subsection{The current state of bioimage analysis}
-\label{sec:org8f5b33d}
+\label{sec:orgc18b83b}
 The most widely-used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
@@
 \subsection{Engineered features in the era of deep learning}
-\label{sec:org9dc3dfa}
+\label{sec:org59080de}
 The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data.
 Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}.
-
+Engineered features can also outperform deep learning ones under some conditions \citep{tangMorphologicalProfilingDrug2024}.
+Even as deep learning features improve over time, they still require complex interpretability methods, if they are interpretable at all.
+Deep Learning also has a higher barrier of entry, since most models require annotating data for training and testing.
+The foundational models that aim to solve this do not generalize well for out-of-distribution data.
 Unlike with deep learning-based features, whose interpretation requires complex methods such as counterfactuals or attention maps, most of the engineered features have a mathematical definition that is vastly simpler to translate into biological insights.
 
-Engineered features can also outperform deep learning ones under some conditions \citep{tangMorphologicalProfilingDrug2024}.
-Even if deep learning features improve over the coming years, they will most likely require more complex interpretability methods, if they are interpretable at all.
-
-It also has a higher barrier of entry, since most models require annotating data for training and testing. Even new foundational models do not generalize well for out-of-distribution data.
-
 The main target users of such a tool would be biologists in need of automation and reproducibility, as well as computer and data scientists that want to dive into the field of biological imaging.
-
 As a whole, there is still a role to be played for engineered features.
 The community is missing a generalist tool that converts images and masks into vectors in an automated fashion and portable across fields.
-It serves as complementary approach deep learning methods.
+It stands as a complement to deep learning methods.
 \section{Extraction of interpretable features using cp\_measure}
-\label{sec:org5633614}
-With cp\_measure we recapitulated all features used in the vast majority of experiments: One object and one imaging channel (e.g., intensity), one object and multiple channels (e.g., Manders correlation) and multiple objects (e.g., number of neighbors).
+\label{sec:orgb8fc762}
+With cp\_measure isolated the now-standard methods for bioimage analysis: One object and one imaging channel (e.g., intensity), one object and multiple channels (e.g., Manders correlation) and multiple objects (e.g., number of neighbors).
 
@@
 \includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
-\caption{\label{fig:orgea7e72a}Extraction of measurements based on a collective of objects.
-cp\_measure generates features from images by using the local information in a mask-defined region of interest.
-Its produces measurements for every object, and is usually used to acquire information on the pairwise combination of all the available channels.}
+\caption{\label{fig:overview}Our library cp\_measure produces features from images by using the local information in every region of interest.
+This can be used to featurize the pairwise combination of all the available channels.}
 \end{figure}
@@
 
-By isolating and cleaning the implemented mathematics of CellProfiler we aim to make the measurement methods more accessible to computer and data scientists, while still leveraging the decade and a half of existing features that are present in numerous datasets, maintaining a higher degree of compatibility.
+By isolating and cleaning the implemented mathematics of CellProfiler we vastly reduce the amount of time and manual effort required to perform data analyses, while also providing the features present in numerous datasets.
 To retain this compatibility in the long term requires contributing this changes back into CellProfiler, be it directly or as a dependency.
 
-When comparing the equivalent pipelines, cp\_measure's minimal overhead results in faster processing times than CellProfiler (TODO add benchmark).
-
-We demonstrate the correlation of cp\_measure compared to a comparable CellProfiler pipeline using a representative subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}.
-Then we demonstrate cases in which cp\_measure can help identify biological phenotypes: first using 3D images of astrocytes and then using spatial transcriptomics dataset.
+First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}.
+Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: first using 3D images of astrocytes and then using spatial transcriptomics dataset.
 These use-cases demonstrate its widespread applicability.
 \subsection{Our features match CellProfiler standard measurements}
-\label{sec:org04a138e}
-
-Our first aim was to validate the features from cp\_measure compared to the CellProfiler standard ones.
-We used images from the JUMP dataset, containing more than 130,000 chemical and genetic perturbations.
-We first curated a set of genetic perturbations that, based on existing analyses, contained cells with distinctive morphologies to all others based on the CellProfiler features.
-Using this curated dataset, we segmented the cytosol and nuclei of the cells using a CellProfiler pipeline to obtain both segmentation masks and the original morphological features.
-
-Next, we used the segmentation masks to calculate the cp\_measure features.
-Then we mapped the features from CellProfiler to cp\_measure and compared the values in both.
-Figure \ref{fig:org7fe862f} shows this comparison.
-Then we demonstrate that we recapitulate the CellProfiler features.
+\label{sec:orga0be6b1}
 
@@
 \includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:org7fe862f}{[}Placeholder] Comparison of CellProfiler and cp\_measure.}
+\caption{\label{fig:cp_vs_cpmeasure}We recapitulate the CellProfiler features.
+\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure, obtained from the same sets of masks and images.
+\textbf{Right panel.} $$R^2$$ value of a linear fit for each individual feature, comparing cp\_measure to cellprofiler.}
 \end{figure}
+
+We first performed a numerical validation of cp\_measure, relative to the original CellProfiler features.
+To do so, used curated a subset of 150 perturbations from the JUMP dataset, selecting the genetic perturbations with the most distinctive features.
+To ensure that we are using the exact same object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, which yielded both morphological profiles and object masks.
+Next, we applied our measurements these masks with the original images and mapped the features from cp\_measure to CelProfiler.
+Figure \ref{fig:cp_vs_cpmeasure} shows some examples of the feature comparison alongside the $$R^2$$ value of a linear fit for all the mapped features.
 \subsection{Results and examples of usage}
-\label{sec:org8ef524e}
+\label{sec:orgf42c78f}
+We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
 \subsubsection{Classification of astrocytes and their distinctive features}
-\label{sec:org3d97e6f}
+\label{sec:org45cf749}
 
@@
 \includesvg[width=.9\linewidth]{./figs/example_shap}
-\caption{\label{fig:org4620d4d}Shapley values of most important features to classify the day in which an image was taken (a multi-class classification task).
+\caption{\label{fig:astrocytes}Shapley values of most important features to classify the day in which an image was taken (a multi-class classification task).
 The test data accuracy is shown in bold.
@@
 \subsubsection{Applicability on spatial transcriptomics}
-\label{sec:org1d0d8b8}
+\label{sec:org1818efd}
 A key advantage of providing these CellProfiler measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
@@
 
-To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy [CITE].
-Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData [CITE] framework underlying Squidpy.
+To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy (CITE).
+Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy.
 Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings—up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts.
@@
 This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
-We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform [CITE, \url{https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/}].
+We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE \url{https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/}).
 Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
 Morphological features were extracted from these 5-channel images for both datasets.
-Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy [CITE] and PyCytoMiner [CITE], respectively.
+Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and PyCytoMiner (CITE) respectively.
 We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig.
@@
 Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included.
-Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp$\backslash$\_measure beyond its traditional scope.
+Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.
 
@@
 \includegraphics[width=.9\linewidth]{./figs/spatial.png}
-\caption{\label{fig:org722dcf5}{[}PLACEHOLDER] Spatial omics analysis.}
+\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
 \end{figure}
 \section{Discussion}
-\label{sec:orgf37b369}
+\label{sec:org9c328df}
 The usage of image analysis pipelines that require manual setups hinders reproducibility and hinders our ability to compare different datasets.
@@
 \section{Future work}
-\label{sec:org5cdbb12}
+\label{sec:org3947c26}
 There are multiple paths to improve and expand the functionality of cp\_measure.
@@
 \section{Methods}
-\label{sec:org91ddb3a}
+\label{sec:org07789c0}
 \subsection{Data and software}
-\label{sec:orgf44b7cd}
+\label{sec:org4dc1288}
 The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
@@
 \bibliographystyle{icml2025}
-\bibliography{bibliography}
+\bibliography{bibliography,~/Documents/broad/bibliography/bibliography}
 \end{document}

### 2025-05-17 draft.tex
draft: minor fixes

@@
-% Created 2025-05-17 Sat 16:54
+% Created 2025-05-17 Sat 17:04
 % Intended LaTeX compiler: xelatex
@@
 \begin{icmlauthorlist}
-\icmlauthor{Alan F. Munoz}{broad}
+\icmlauthor{Al\'an F. Munoz}{broad}
 \icmlauthor{Tim Treis}{hh,broad}
@@
 \section{Introduction}
-\label{sec:orgbc2e426}
+\label{sec:org24b39bd}
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
@@
 \subsection{The current state of bioimage analysis}
-\label{sec:orgc18b83b}
+\label{sec:orgb31be4c}
 The most widely-used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
@@
 \subsection{Engineered features in the era of deep learning}
-\label{sec:org59080de}
+\label{sec:org1df9830}
 The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data.
@@
 \section{Extraction of interpretable features using cp\_measure}
-\label{sec:orgb8fc762}
+\label{sec:orgae39a4b}
 With cp\_measure isolated the now-standard methods for bioimage analysis: One object and one imaging channel (e.g., intensity), one object and multiple channels (e.g., Manders correlation) and multiple objects (e.g., number of neighbors).
@@
 \includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
-\caption{\label{fig:overview}Our library cp\_measure produces features from images by using the local information in every region of interest.
+\caption{\label{fig:overview}Our library $$cp_measure$$ produces features from images by using the local information in every region of interest.
 This can be used to featurize the pairwise combination of all the available channels.}
@@
 \subsection{Our features match CellProfiler standard measurements}
-\label{sec:orga0be6b1}
+\label{sec:org5ae7c2d}
 
@@
 \caption{\label{fig:cp_vs_cpmeasure}We recapitulate the CellProfiler features.
-\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure, obtained from the same sets of masks and images.
-\textbf{Right panel.} $$R^2$$ value of a linear fit for each individual feature, comparing cp\_measure to cellprofiler.}
+\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to $$cp_measure$$, obtained from the same sets of masks and images.
+\textbf{Right panel.} $$R^2$$ value of a linear fit for each individual feature, comparing $$cp_measure$$ to CellProfiler.}
 \end{figure}
@@
 \subsection{Results and examples of usage}
-\label{sec:orgf42c78f}
+\label{sec:org6cadf0d}
 We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
 \subsubsection{Classification of astrocytes and their distinctive features}
-\label{sec:org45cf749}
+\label{sec:org241d95b}
 
@@
 \subsubsection{Applicability on spatial transcriptomics}
-\label{sec:org1818efd}
+\label{sec:org5419370}
 A key advantage of providing these CellProfiler measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
@@
 \section{Discussion}
-\label{sec:org9c328df}
+\label{sec:org2b3c331}
 The usage of image analysis pipelines that require manual setups hinders reproducibility and hinders our ability to compare different datasets.
@@
 \section{Future work}
-\label{sec:org3947c26}
+\label{sec:org76eaf99}
 There are multiple paths to improve and expand the functionality of cp\_measure.
@@
 \section{Methods}
-\label{sec:org07789c0}
+\label{sec:orga949269}
 \subsection{Data and software}
-\label{sec:org4dc1288}
+\label{sec:orgfc7bf0e}
 The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.

### 2025-05-17 draft.tex
fix: replace hyphen with --, minor fixes

--- 2025-05-17 draft.tex
@@
-% Created 2025-05-17 Sat 16:54
+% Created 2025-05-17 Sat 17:19
 % Intended LaTeX compiler: xelatex
@@
 \section{Introduction}
-\label{sec:orgbc2e426}
+\label{sec:org372736e}
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
@@
 \subsection{The current state of bioimage analysis}
-\label{sec:orgc18b83b}
+\label{sec:org1c5c74c}
 The most widely-used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
@@
 \subsection{Engineered features in the era of deep learning}
-\label{sec:org59080de}
+\label{sec:orgc755c87}
 The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data.
@@
 \section{Extraction of interpretable features using cp\_measure}
-\label{sec:orgb8fc762}
+\label{sec:org8ca3631}
 With cp\_measure isolated the now-standard methods for bioimage analysis: One object and one imaging channel (e.g., intensity), one object and multiple channels (e.g., Manders correlation) and multiple objects (e.g., number of neighbors).
@@
 \subsection{Our features match CellProfiler standard measurements}
-\label{sec:orga0be6b1}
+\label{sec:orge7bf29c}
 
@@
 \caption{\label{fig:cp_vs_cpmeasure}We recapitulate the CellProfiler features.
-\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure, obtained from the same sets of masks and images.
-\textbf{Right panel.} $$R^2$$ value of a linear fit for each individual feature, comparing cp\_measure to cellprofiler.}
+\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure's, obtained from the same sets of masks and images.
+\textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
 \end{figure}
@@
 Next, we applied our measurements these masks with the original images and mapped the features from cp\_measure to CelProfiler.
-Figure \ref{fig:cp_vs_cpmeasure} shows some examples of the feature comparison alongside the $$R^2$$ value of a linear fit for all the mapped features.
+Figure \ref{fig:cp_vs_cpmeasure} shows some examples of the feature comparison alongside the \(R^2\) value of a linear fit for all the mapped features.
 \subsection{Results and examples of usage}
-\label{sec:orgf42c78f}
+\label{sec:orgc72999e}
 We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
 \subsubsection{Classification of astrocytes and their distinctive features}
-\label{sec:org45cf749}
+\label{sec:org0622674}
 
@@
 \subsubsection{Applicability on spatial transcriptomics}
-\label{sec:org1818efd}
+\label{sec:org6a1cdc5}
 A key advantage of providing these CellProfiler measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
@@
 However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment.
-In contrast, classical morphological measurements yield explicit, interpretable readouts—for instance, the co-localization of fluorescent markers—facilitating clear biological interpretations.
+In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.
 
@@
 Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy.
-Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings—up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts.
+Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts.
 Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores.
@@
 \section{Discussion}
-\label{sec:org9c328df}
+\label{sec:org3adb4ba}
 The usage of image analysis pipelines that require manual setups hinders reproducibility and hinders our ability to compare different datasets.
@@
 \section{Future work}
-\label{sec:org3947c26}
+\label{sec:orgd895d97}
 There are multiple paths to improve and expand the functionality of cp\_measure.
@@
 \section{Methods}
-\label{sec:org07789c0}
+\label{sec:orgf3a6049}
 \subsection{Data and software}
-\label{sec:org4dc1288}
+\label{sec:orgba4bd5c}
 The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.

### 2025-05-17 draft.tex
Updates from Overleaf

--- 2025-05-17 draft.tex
@@
 \caption{\label{fig:cp_vs_cpmeasure}We recapitulate the CellProfiler features.
-\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to $$cp_measure$$, obtained from the same sets of masks and images.
-\textbf{Right panel.} $$R^2$$ value of a linear fit for each individual feature, comparing $$cp_measure$$ to CellProfiler.}
+\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp_measure, obtained from the same sets of masks and images.
+\textbf{Right panel.} $R^2$ value of a linear fit for each individual feature, comparing cp_measure to CellProfiler.}
 \end{figure}

### 2025-05-18 draft.tex
finish first draft (missing spatial fig)

--- 2025-05-17 draft.tex
@@
-% Created 2025-05-17 Sat 17:19
-% Intended LaTeX compiler: xelatex
+% Created 2025-05-17 Sat 20:18
+% Intended LaTeX compiler: pdflatex
 \documentclass{article}
@@
 \date{}
-\title{draft}
+\title{}
 \hypersetup{
- pdfauthor={},
+ pdfauthor={Alan Munoz},
  pdftitle={},
@@
  pdfsubject={},
- pdfcreator={},
+ pdfcreator={Emacs 30.1 (Org mode 9.7.11)}, 
  pdflang={English}}
@@
 \section{Introduction}
-\label{sec:org372736e}
+\label{sec:org4c9ba67}
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
@@
 \subsection{The current state of bioimage analysis}
-\label{sec:org1c5c74c}
-The most widely-used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
+\label{sec:org8f5b33d}
+The most widely used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
 Its target users are biologists with limited programmatic expertise.
-The accessibility that manually implemented workflows reduces its suitability for automated analyses and the composition of more complex pipelines.
-In this context we define accessibility as the lack of friction to use the software in computational pipelines.
+While this has been of great benefit for biologists, it has a number of issues that hamper its usability in many contexts.
 
-The usage of CellProfiler entails a higher degree of friction for composite pipelines that orchestrate multiple tools.
-Its focus on manually-defined workflows adjusted for specific use-cases makes it less suitable for many high throughput analyses.
-Additionally, CellProfiler has a significant number of dependencies.
-It depends on libraries from both the Python and Java languages, thus increasing the number of things that may go wrong when setting up or integrating with other tools.
-Tough containers partially alleviate this problem, they carry additional complexity and caveats.
+CellProfiler elicits a higher degree of friction for composite pipelines that orchestrate multiple tools.
+Its focus on manually-defined workflows adjusted for specific use-cases makes it less suitable for low-customization high throughput analyses, more so since the development of generalist segmentation methods and foundational models.
+Additionally, its large number of dependencies, which include modules from both the Python and Java, can increase the complexity of setup if one is to integrate CellProfiler with other tools.
+Tough containerisation mitigates this problem, they are not without their own complexity and caveats.
 
-A small subset of the features in CellProfiler is provided by scikit-image, but this is greatly limited and cannot be reliably compared to the existing body of work that used CellProfiler to obtain the measurements.
-The other big problem of CellProfiler is its dependency on human interaction at many steps of the analysis, opening the door to human mistakes and hindering analyses reproducibility.
+Though certain morphological features can be obtained using scikit-image, these are a fraction of the current standard and cannot be directly compared to the existing datasets produced using CellProfiler.
+Another challenge resultant of using CellProfiler is its dependency on human interaction at multiple points, which provides the ability of inspecting and parameterizing data but opens the door to human mistakes and hindering reproducibility.
 
-A CellProfiler pipeline is does not integrate well with other tools.
+A CellProfiler pipeline struggles integrating well with other computational tools.
 Its existing plug-in system is inaccessible for most users and has strong limitations.
-Building automated pipelines that contain CellProfiler and integrating tools within its components are time and effort-consuming challenges.
+Building automated pipelines that contain CellProfiler and integrating tools within its components are time and effort-consuming challenges that require a decent understanding of its interfaces.
+Treating it as a black box is an option, but its adding functionality, such as adding a custom-made image filter is not possible.
 
-In this space there is a vacuum for a software library that covers the generalist and high throughput use-cases for which CellProfiler is inadequate.
+There is the need for a software library that covers the generalist and high throughput use-cases for which CellProfiler is inadequate.
 Distributed CellProfiler would cover this case if not for its cloud-only approach and dependency on preconfigured pipelines, as well as its limited debugging capabilities \citep{mcquinCellProfiler30Nextgeneration2018}.
-
-Other recent software has been published aiming to fill this unoccupied niche, but their solutions are unsuitable for practical usage at scale: First, their features have completely independent implementations from the original ones, thus are not guaranteed to reproduce existing results and the biological interpretations may be different from the current field standard.
-Additionally, their implementations are not modular nor scale well upon increases to the data size.
-Some of them, akin to Distributed CellProfiler, are based on main use-case is cloud services.
+Some tools, such as ScaleFex \citep{comoletHighlyEfficientScalable2024}, attempt to fill this gaping hole in the biological software, but they have limitations: First, their features have completely independent implementations from the original ones, thus are not reproduce existing results and the biological interpretations may be different from the current field standard.
+Additionally, their implementations are usually an off-the-shelf solution, which works well if they match our use-case, but they are not modular nor usually scale well upon increases to the data size.
+Other alternatives are directed at cloud-first and neglect the use-cases for which local compute would suffice.
 \subsection{Engineered features in the era of deep learning}
-\label{sec:orgc755c87}
+\label{sec:org9dc3dfa}
 The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data.
 Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}.
-Engineered features can also outperform deep learning ones under some conditions \citep{tangMorphologicalProfilingDrug2024}.
-Even as deep learning features improve over time, they still require complex interpretability methods, if they are interpretable at all.
+Engineered features can also outperform deep learning ones under certain conditions \citep{tangMorphologicalProfilingDrug2024}.
+Even as deep learning features improve over time, they still require complex interpretability methods, in the rare case where they can be interpreted.
 Deep Learning also has a higher barrier of entry, since most models require annotating data for training and testing.
-The foundational models that aim to solve this do not generalize well for out-of-distribution data.
-Unlike with deep learning-based features, whose interpretation requires complex methods such as counterfactuals or attention maps, most of the engineered features have a mathematical definition that is vastly simpler to translate into biological insights.
+The foundational models whose goal is to bypass this step do not generalize well for out-of-distribution data.
+Unlike deep learning-based features, whose interpretation requires complex methods such as counterfactuals or attention maps, most of the engineered features have a mathematical definition that is vastly simpler to translate into biological insights.
 
-The main target users of such a tool would be biologists in need of automation and reproducibility, as well as computer and data scientists that want to dive into the field of biological imaging.
-As a whole, there is still a role to be played for engineered features.
-The community is missing a generalist tool that converts images and masks into vectors in an automated fashion and portable across fields.
-It stands as a complement to deep learning methods.
+The core target users of such a tool would be biologists and data scientists who need reproducibility and data within a few lines of code.
+As a whole, there is still a role to be played for engineered features to, at the very least, complement deep learning ones.
+The community is missing a generalist tool that converts images and masks into representative vectors without the need of interfaces or cloud compute and that integrates seamlessly with other scientific libraries.
+This is the role that we envision cp\_measure can occupy in the scientific software ecosystem.
 \section{Extraction of interpretable features using cp\_measure}
-\label{sec:org8ca3631}
-With cp\_measure isolated the now-standard methods for bioimage analysis: One object and one imaging channel (e.g., intensity), one object and multiple channels (e.g., Manders correlation) and multiple objects (e.g., number of neighbors).
+\label{sec:org5633614}
+In cp\_measure, we isolated the CellProfiler measurements and categorize them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (such as correlation features) and multiple objects (e.g., number of neighbors).
 
@@
 \includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
-\caption{\label{fig:overview}Our library cp\_measure produces features from images by using the local information in every region of interest.
-This can be used to featurize the pairwise combination of all the available channels.}
+\caption{\label{fig:overview}Our tool produces features from images by using the local information in every region of interest.
+This can be used to featurize the pairwise combination of all the available channels.
+The resultant matrices represent the entirety of the experiment in a much more compressed way, and can be processed with standard machine and deep learning methods.}
 \end{figure}
 
-The cp\_measure's code base was based on CellProfiler's, but we removed the sections of the code that are unnecessary for programmatic usage, thus separating the actual functional components from the orchestration logic.
+The cp\_measure features were based off the CellProfiler implementation, but we separated the actual functional components (i.e., math) from the user-facing and orchestration components.
 It improves reusability, testability and its makes it possible for more people to access and improve its internals in the future.
-It aims to be a stepping stone to increased extensibility and maintainability.
+We positioned it to become a stepping stone in extensibility and maintainability.
+By isolating and cleaning the implemented mathematics of CellProfiler we vastly reduce the amount of time and manual effort required to perform data analyses while retaining the features present in numerous datasets.
 
-Our library aims to remain consistent with the current scientific python ecosystem.
-The main interface matches that of the widely-used scikit-image \citep{waltScikitimageImageProcessing2014}.
+Our aim is to remain consistent with the current scientific python ecosystem.
+Its interface corresponds that of the widely used scikit-image \citep{waltScikitimageImageProcessing2014}.
 This greatly reduces the effort needed to integrate it in existing workflows and tools.
 
-By isolating and cleaning the implemented mathematics of CellProfiler we vastly reduce the amount of time and manual effort required to perform data analyses, while also providing the features present in numerous datasets.
-To retain this compatibility in the long term requires contributing this changes back into CellProfiler, be it directly or as a dependency.
-
+We will now demonstrate the validity and capabilities of cp\_measure in cases that would normally require domain expertise and a considerable amount of time.
 First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}.
-Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: first using 3D images of astrocytes and then using spatial transcriptomics dataset.
-These use-cases demonstrate its widespread applicability.
-\subsection{Our features match CellProfiler standard measurements}
-\label{sec:orge7bf29c}
+Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: First 3D images of astrocytes and then spatial transcriptomics.
+These use-cases demonstrate its widespread suitability for different types of problems.
+\subsection{Recapitulating CellProfiler measurements}
+\label{sec:org09b0cd2}
 
@@
 \includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:cp_vs_cpmeasure}We recapitulate the CellProfiler features.
-\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure's, obtained from the same sets of masks and images.
+\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
+\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure's, generated using matching pairs of masks and images.
 \textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
@@
 
-We first performed a numerical validation of cp\_measure, relative to the original CellProfiler features.
-To do so, used curated a subset of 150 perturbations from the JUMP dataset, selecting the genetic perturbations with the most distinctive features.
-To ensure that we are using the exact same object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, which yielded both morphological profiles and object masks.
-Next, we applied our measurements these masks with the original images and mapped the features from cp\_measure to CelProfiler.
-Figure \ref{fig:cp_vs_cpmeasure} shows some examples of the feature comparison alongside the \(R^2\) value of a linear fit for all the mapped features.
-\subsection{Results and examples of usage}
-\label{sec:orgc72999e}
+We first performed the numerical validation of cp\_measure, relative to the original CellProfiler features.
+In preparation, we collected 300 images corresponding to 150 perturbations from the JUMP dataset.
+the most significant phenotypes for a given measurement each.
+To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features.
+Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CelProfiler.
+Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value.
+
+The validity of cp\_measure is shown on Figure \ref{fig:cp_vs_cpmeasure}.
+We show examples of the features on the panel on the left.
+Mostly straight lines demonstrate the recapitulation of measurements from our implementation.
+A few data points fall outside the diagonals, which may indicate that some edge-cases are treated differently.
+The panel on the right shows the \(R^2\) value of a linear interpolation.
+Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements resulted in a linear relation, regardless of whether the masks were for nuclei or cytosols.
+This result provides reassurance that cp\_measure can be confidently in cases where CellProfiler would be used.
+\subsection{Results}
+\label{sec:orge5b5c6b}
 We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
-\subsubsection{Classification of astrocytes and their distinctive features}
-\label{sec:org0622674}
+\subsubsection{Astrocytes 3D data}
+\label{sec:org447090b}
 
-As a demonstration of its ease of use, we used cp\_measure for featurization in a cell classification workflow.
-We used it to process 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}.
-We then calculated the median value for every feature in a cell and the number of cells, following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-Then we trained a Gradient Boosting classifier to identify which day.
+As a demonstration of its ease of use and the value of interpretable features, we used cp\_measure in a standard classification workflow.
+We processed 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}.
+We preprocessed the data following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
+Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired.
 With this we identified which features distinguish cells on the later samples and distinguish subpopulations.
-It is worth noting that there will be some redundancy in the information contained in the cp\_measure features, and thus during cases in which multiple features inform on similar data subsets.
+Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells [CITE Shapley].
 
@@
 \includesvg[width=.9\linewidth]{./figs/example_shap}
-\caption{\label{fig:astrocytes}Shapley values of most important features to classify the day in which an image was taken (a multi-class classification task).
-The test data accuracy is shown in bold.
-Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis.}
+\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astroctyes image and masks.
+The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
+\textbf{Bottom panel.} Shapley values of the most important features to classify the day in which an image was taken (out of three).
+The test data accuracy is shown in bold.}
 \end{figure}
-\subsubsection{Applicability on spatial transcriptomics}
-\label{sec:org6a1cdc5}
-A key advantage of providing these CellProfiler measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
+
+Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on cp\_measure features.
+Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis.
+While it was posisble to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
+\subsubsection{Spatial transcriptomics}
+\label{sec:org5711d86}
+A key advantage of providing these measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
 The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality.
@@
 This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
-We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE \url{https://nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/}).
+We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/).
 Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
@@
 \section{Discussion}
-\label{sec:org3adb4ba}
+\label{sec:orgf37b369}
 The usage of image analysis pipelines that require manual setups hinders reproducibility and hinders our ability to compare different datasets.
 In this work we introduced our new library cp\_measure, which provides widely used engineered features and enables simpler automated analyses of microscopy data in either short scripts and complex pipelines.
-This also removes the requirement of using graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy data without the need of cloud-based infrastructure.
+This also removes the requirement of using graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy even without cloud infrastructure.
 
@@
 These measurements have already been used in non-biological contexts, such as environmental monitoring \citep{ideharaExploringNileRed2025}, thus these engineered metrics also benefit other scientific fields beyond morphological profiling.
-In general, we see the decomposition of CellProfiler into modular components as a way to facili
 \section{Future work}
-\label{sec:orgd895d97}
-There are multiple paths to improve and expand the functionality of cp\_measure.
-The first and most obvious is to integrate its measurements back to CellProfiler library.
-This would ensure that the results from pipelines built with either tool will be comparable in the future, while also providing the opportunity of formalizing the programmatic interface --- inputs and outputs --- of measurements.
+\label{sec:org5cdbb12}
+The most obvious way to make cp\_measure more useful is to contribute it back to CellProfiler.
+This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity of formalizing the inputs and outputs of all measurements.
 
-Developing a comprehensive tests suite would guarantee mathematical correctness under the possible edge cases that may be encountered when dealing with new data.
-This test suite in turn would in turn open the door to further speed-ups in multiple ways: Firstly, optimizing the measurements that consume the most time, starting with object granularity (\textasciitilde{}80\% of the time).
-Additionally, it is possible to implement measurements using numba for just-in-time compiling and/or adding GPU support \citep{lamNumbaLLVMbasedPython2015}.
-
-There is further space for improvement.
-First, provide a wrapper for all features that masp to scikit-image's regionprops as close as possible.
-Secondly, a list of essential measurements for use-cases where speed is more important than using all the features.
-By lowering the barrier of effort required to integrate cp\_measure into existing pipelines these
+Developing a comprehensive tests suite will guarantee mathematical correctness, which currently not even CellProfiler has.
+This test suite in turn would in turn expedite improvements in multiple ways: Firstly, optimizing the most compute-consuming features, such as granularity.
+Later on, we could add to support just-in-time compiling and GPUs.
 
 Long-term, we envision cp\_measure can be the place to develop and distribute new measurements.
-While CellProfiler's measurements are widely used in bioimaging studies, the existing palette of measurements could be further extended to cover novel use-cases brought upon by novel developments in imaging acquisition devices and methods.
-Working with the community to further the number of measurements to better match the current questions scientists pose to imaging data.
-\section{Methods}
-\label{sec:orgf3a6049}
-\subsection{Data and software}
-\label{sec:orgba4bd5c}
-The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
-All code to reproduce the analyses and figures, alongside links to the original data, is available on the Github repository \url{https://github.com/afermg/2025\_cpmeasure/}.
-The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.
+While CellProfiler's measurements are already ubiquituous in bioimaging studies, the existing palette of measurements could be further extended to cover unexplored use-cases.
+We also see adding community-contributed measurements to better match the current questions scientists pose to imaging data.
 
 \bibliographystyle{icml2025}
-\bibliography{bibliography,~/Documents/broad/bibliography/bibliography}
+\bibliography{bibliography}
 \end{document}

### 2025-05-18 src/astro3d/get_figures.py

@@ -174,12 +174,7 @@ Xd = DMatrix(X_train, label=y_train)
 Xd_test = DMatrix(X_test, label=y_test)
 model = xgboost.train({"eta": 1, "max_depth": 3, "base_score": 0, "lambda": 0}, Xd, 1)
 
-# bst.fit(X_train, y_train)
-# preds = bst.predict(X_test)
-# test_acc = (preds & np.array(y_test)).sum() / len(y_test)
-# print(test_acc)
 print("Model error =", np.linalg.norm(y_train - model.predict(Xd)))
-print(model.get_dump(with_stats=True)[0])
 print(
     f"Training prediction accuracy {np.round((y_train - np.round(model.predict(Xd)) == 0).sum() / len(y_train), 2)}"
 )
@@ -224,7 +219,7 @@ axd = plt.figure(layout="constrained").subplot_mosaic(
 i = 1
 img = imread(pairs[i][1]).max(axis=0)
 labels = imread(pairs[i][0]).max(axis=0)
-print(labels.max())
+
 axd["A"].imshow(img)
 axd["A"].axis("off")
 axd["A"].set_title("Z-projected image")
@@ -232,35 +227,12 @@ axd["C"].imshow(labels)
 axd["C"].axis("off")
 axd["C"].set_title("Labels")
 shap.plots.beeswarm(explanation, max_display=6, ax=axd["B"], plot_size=None)
-axd["B"].set_yticklabels(axd["B"].get_yticklabels(), rotation=30)
+axd["B"].set_yticklabels(
+    axd["B"].get_yticklabels(),
+    rotation=-30,
+    ha="right",
+    rotation_mode="anchor",
+)
 plt.text(0.15, -0.5, acc, fontweight="bold")
-# plt.tight_layout()
 plt.savefig(figs_path / "example_shap.svg")
 plt.close()
-# %%
-import matplotlib.pyplot as plt
-import numpy as np
-
-# prepare some coordinates
-x, y, z = np.indices((8, )
-
-# draw cuboids in the top left and bottom right corners, and a link between
-# them
-cube1 = (x < 3) & (y < 3) & (z < 3)
-cube2 = (x >= 5) & (y >= 5) & (z >= 5)
-link = abs(x - y) + abs(y - z) + abs(z - x) <= 2
-
-# combine the objects into a single boolean array
-voxelarray = cube1 | cube2 | link
-
-# set the colors of each object
[…]

### 2025-05-18 draft.tex
docs: Update feature section titles

@@
-% Created 2025-05-17 Sat 20:18
+% Created 2025-05-17 Sat 20:28
 % Intended LaTeX compiler: pdflatex
@@
 This is the role that we envision cp\_measure can occupy in the scientific software ecosystem.
-\section{Extraction of interpretable features using cp\_measure}
-\label{sec:org5633614}
+\section{Extracting interpretable features}
+\label{sec:org61842b5}
 In cp\_measure, we isolated the CellProfiler measurements and categorize them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (such as correlation features) and multiple objects (e.g., number of neighbors).

### 2025-05-19 draft.tex
docs: revamp first half of introduction

@@
-% Created 2025-05-17 Sat 20:28
+% Created 2025-05-19 Mon 17:52
 % Intended LaTeX compiler: pdflatex
@@
 \twocolumn[
-\icmltitle{cp\_measure: Morphological features for bioimaging}
+\icmltitle{cp\_measure: Morphological profiling for data scientists}
 
@@
 \begin{icmlauthorlist}
-\icmlauthor{Al\'an F. Munoz}{broad}
+\icmlauthor{Al\'an F. Mu\~{n}oz}{broad}
 \icmlauthor{Tim Treis}{hh,broad}
@@
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
-Microscopy techniques allows us to acquire information from a multitude of sources, such as illumination, fluorescent dyes or even proteins tagged with other proteins that fluoresce, acting as a beacon and providing information on quantity and/or localization.
+Modern microscopes can capture a wisde range of information, such as illumination and fluorescence.
+When coupled with fluorescence dyes or proteins biologists are able to observe the location and distribution of cells and organelles.
 
-The main task at the core bioimage analysis is identifying features that convey information of the state of a cell or its contents.
-This entails combining identified objects with pixel data to quantify their morphology and other information sources, such as intensity distributions.
+The main purpose of image analysis in the context of biology is to characterize the cell state or its components.
+Early on, human eyes where the sole judge on this, but a more objective, scalable and reproducible method was required.
+This commonly entailed identifying regions of interest and calculating metrics that represent these regions, such as intensity distributions.
 
-Morphological profiling a subfield of bioimaging that focuses on the cells and the shapes and properties of their organelles.
-It is usually supported by both statistical methods and machine and deep learning approaches.
-One of its biggest applications is drug discovery, where it is used primarily due to its low acquisition cost and capabilities for scaling to screen many compounds efficiently \citep{sealDecadeSystematicReview2024}.
-It can also complement to other high throughput methods, such as genomic sequencing.
+Morphological profiling is a technique where we generate a large number of shape nad intensity features for all objects, cells or otherwise.
+These features are fed into statistical or machine learning methods to identify patterns.
+One of its biggest applications is drug discovery, where scientists leverages microscopy's low acquisition cost and high throughput \citep{sealDecadeSystematicReview2024}.
+Morphological profiling serves not only as validation but also as information orthogonal to other high throughput methods, such as genomic sequencing.
 \subsection{The current state of bioimage analysis}
 \label{sec:org8f5b33d}
-The most widely used software to process biological imaging data is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
-Its target users are biologists with limited programmatic expertise.
-While this has been of great benefit for biologists, it has a number of issues that hamper its usability in many contexts.
-
-CellProfiler elicits a higher degree of friction for composite pipelines that orchestrate multiple tools.
-Its focus on manually-defined workflows adjusted for specific use-cases makes it less suitable for low-customization high throughput analyses, more so since the development of generalist segmentation methods and foundational models.
-Additionally, its large number of dependencies, which include modules from both the Python and Java, can increase the complexity of setup if one is to integrate CellProfiler with other tools.
-Tough containerisation mitigates this problem, they are not without their own complexity and caveats.
-
-Though certain morphological features can be obtained using scikit-image, these are a fraction of the current standard and cannot be directly compared to the existing datasets produced using CellProfiler.
-Another challenge resultant of using CellProfiler is its dependency on human interaction at multiple points, which provides the ability of inspecting and parameterizing data but opens the door to human mistakes and hindering reproducibility.
+The most widely used software to process biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
+Its intended audience is experimental biologists with scant programmatic experience, for whom it has been a boon.
+In contrast, its usefulness greatly diminishes for computer-savvy scientists who fear no command line.
+For them, CellProfiler means friction in pipelines comprised of multiple tools.
+Its catering for manually-defined workflows unbefits low-customization high throughput analyses, which are prevalent in the age of generalist segmentation and foundational models.
 
 A CellProfiler pipeline struggles integrating well with other computational tools.
-Its existing plug-in system is inaccessible for most users and has strong limitations.
-Building automated pipelines that contain CellProfiler and integrating tools within its components are time and effort-consuming challenges that require a decent understanding of its interfaces.
-Treating it as a black box is an option, but its adding functionality, such as adding a custom-made image filter is not possible.
+Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces.
+Treating it as a black box and using its output precludes the addition of custom-made image filters.
+Lastly, CellProfiler depends on many number of packages from both Python and Java, amplifying the risk of dependency conflicts.
+Containers mitigate this problem, but not without increased complexity and caveats.
+To summarize, CellProfiler is an all-or-nothing tool, integrating other tools demands more work and time than most scientists can spare.
 
-There is the need for a software library that covers the generalist and high throughput use-cases for which CellProfiler is inadequate.
-Distributed CellProfiler would cover this case if not for its cloud-only approach and dependency on preconfigured pipelines, as well as its limited debugging capabilities \citep{mcquinCellProfiler30Nextgeneration2018}.
-Some tools, such as ScaleFex \citep{comoletHighlyEfficientScalable2024}, attempt to fill this gaping hole in the biological software, but they have limitations: First, their features have completely independent implementations from the original ones, thus are not reproduce existing results and the biological interpretations may be different from the current field standard.
-Additionally, their implementations are usually an off-the-shelf solution, which works well if they match our use-case, but they are not modular nor usually scale well upon increases to the data size.
-Other alternatives are directed at cloud-first and neglect the use-cases for which local compute would suffice.
+Object featurization is not exclusive of CellProfiler.
+For example, scikit-image can produce some overlapping features.
+Tools like ScaleFex \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations.
+Moreover, it enforces a specific data or metadata structure, which also adds complexity to the setup.
+Lastly, a common pattern profilers akin to ScaleFex is that they enforce a specific shape for the input data and require wrangling with the data before processing a single image.
+Some alternatives, such as ScaleFex, are designed for cloud usage, neglectling a significant number of cases where local compute suffices.
+We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
 \subsection{Engineered features in the era of deep learning}

### 2025-05-19 draft.tex
docs: finish introduction revamp

@@
-% Created 2025-05-19 Mon 17:52
+% Created 2025-05-19 Mon 18:10
 % Intended LaTeX compiler: pdflatex
@@
 
-A CellProfiler pipeline struggles integrating well with other computational tools.
+Integrating a CellProfiler pipelines with other tools is a struggle.
 Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces.
@@
 Tools like ScaleFex \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations.
-Moreover, it enforces a specific data or metadata structure, which also adds complexity to the setup.
-Lastly, a common pattern profilers akin to ScaleFex is that they enforce a specific shape for the input data and require wrangling with the data before processing a single image.
-Some alternatives, such as ScaleFex, are designed for cloud usage, neglectling a significant number of cases where local compute suffices.
+Moreover, a common portability problem is that these tools enforce a specific shape for the input data and require wrangling with the data before processing a single image.
+Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices.
 We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
@@
 Engineered features can also outperform deep learning ones under certain conditions \citep{tangMorphologicalProfilingDrug2024}.
-Even as deep learning features improve over time, they still require complex interpretability methods, in the rare case where they can be interpreted.
-Deep Learning also has a higher barrier of entry, since most models require annotating data for training and testing.
-The foundational models whose goal is to bypass this step do not generalize well for out-of-distribution data.
-Unlike deep learning-based features, whose interpretation requires complex methods such as counterfactuals or attention maps, most of the engineered features have a mathematical definition that is vastly simpler to translate into biological insights.
+Even as deep learning features improve over time, they still require complex interpretability methods, even if they manage to be interpretable at all.
+Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require data annotating.
+Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data.
+Unlike deep learning-based features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and simpler to translate into biological concepts.
 
-The core target users of such a tool would be biologists and data scientists who need reproducibility and data within a few lines of code.
-As a whole, there is still a role to be played for engineered features to, at the very least, complement deep learning ones.
-The community is missing a generalist tool that converts images and masks into representative vectors without the need of interfaces or cloud compute and that integrates seamlessly with other scientific libraries.
-This is the role that we envision cp\_measure can occupy in the scientific software ecosystem.
+The intended audience for such a featurization tool would be, in addition to biologists, data scientists striving to achieve reproducible data analyses within a few lines of code.
+As a whole, engineered features still have the role of, at the minimum, complementing deep learning ones.
+The bioimage software ecosystems lacks a modular tool to featurize images into interpretable features with no compulsory graphical interfaces or cloud compute.
+Ideally it should also harmonize with other scientific libraries.
+With all this in mind we have developed cp\_measure.
 \section{Extracting interpretable features}

### 2025-05-20 draft.tex
docs: Update draft text, convert to Oxford style

@@
-% Created 2025-05-19 Mon 18:10
+% Created 2025-05-19 Mon 21:10
 % Intended LaTeX compiler: pdflatex
@@
 \begin{abstract}
-Quantifying the contents of objects in images is a common challenge in biological imaging.
-The most widely used software to do so require significant manual intervention.
+Quantifying the contents of objects in images is a core challenge in biological imaging.
+The current tools require significant human intervention.
 Here we introduce our library cp\_measure, which provides programmatic access to the most widespread metrics to convert images and objects into features.
 We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives.
-Our tool opens the door to community-driven  development and expansion of bioimage analysis metrics and pipelines, increasing developer accessibility and reproducibility of the pipelines.
+cp\_measure opens the door to community-driven development and improvement of bioimage analysis metrics and pipelines, increasing the scaling capabilities, reproducibility and accessibility for computational and data scientists.
 \end{abstract}
@@
 The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
-Modern microscopes can capture a wisde range of information, such as illumination and fluorescence.
+Modern microscopes can capture a wide range of information, such as illumination and fluorescence.
 When coupled with fluorescence dyes or proteins biologists are able to observe the location and distribution of cells and organelles.
 
-The main purpose of image analysis in the context of biology is to characterize the cell state or its components.
+The main purpose of image analysis in biology is to characterize the cell state or its components.
 Early on, human eyes where the sole judge on this, but a more objective, scalable and reproducible method was required.
@@
 
-Morphological profiling is a technique where we generate a large number of shape nad intensity features for all objects, cells or otherwise.
+Morphological profiling is a technique where we generate an array of shape and intensity features for all objects, cells or otherwise.
 These features are fed into statistical or machine learning methods to identify patterns.
@@
 For them, CellProfiler means friction in pipelines comprised of multiple tools.
-Its catering for manually-defined workflows unbefits low-customization high throughput analyses, which are prevalent in the age of generalist segmentation and foundational models.
+It caters for manually-defined workflows instead of low-customization high throughput analyses, which are becoming prevalent in the age of generalist segmentation and foundational models.
 
@@
 Containers mitigate this problem, but not without increased complexity and caveats.
-To summarize, CellProfiler is an all-or-nothing tool, integrating other tools demands more work and time than most scientists can spare.
+To summarize, CellProfiler is an all-or-nothing tool, whose extensibility and integration with other tools demands more time and effort than most scientists can spare.
 
 Object featurization is not exclusive of CellProfiler.
-For example, scikit-image can produce some overlapping features.
-Tools like ScaleFex \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations.
-Moreover, a common portability problem is that these tools enforce a specific shape for the input data and require wrangling with the data before processing a single image.
+For example, scikit-image can produce some overlapping features \citep{waltScikitimageImageProcessing2014}.
+Tools like ScaleFex or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} can generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations.
+Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image.
 Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices.
@@
 \label{sec:org9dc3dfa}
-The unprecedented evolution of deep learning methods in recent years has made available many different ways to process imaging data.
-Nevertheless, these are limited in their capacity to provide biological interpretation \citep{moenDeepLearningCellular2019}.
-Engineered features can also outperform deep learning ones under certain conditions \citep{tangMorphologicalProfilingDrug2024}.
-Even as deep learning features improve over time, they still require complex interpretability methods, even if they manage to be interpretable at all.
-Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require data annotating.
+The latest deep learning models can streamline image processing.
+Nevertheless, their capacity to provide biological interpretation is limited \citep{moenDeepLearningCellular2019}.
+In some cases, engineered features can even outperform learned ones \citep{tangMorphologicalProfilingDrug2024}.
+Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data.
 Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data.
-Unlike deep learning-based features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and simpler to translate into biological concepts.
+Unless a major change in neural networks occurs, they will still require more complex interpretability methods.
+Unlike learned features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
 
@@
 As a whole, engineered features still have the role of, at the minimum, complementing deep learning ones.
-The bioimage software ecosystems lacks a modular tool to featurize images into interpretable features with no compulsory graphical interfaces or cloud compute.
+The bioimage software ecosystems lacks a modular tool to convert images into interpretable features with no compulsory graphical interfaces or cloud compute.
 Ideally it should also harmonize with other scientific libraries.
@@
 \label{sec:org61842b5}
-In cp\_measure, we isolated the CellProfiler measurements and categorize them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (such as correlation features) and multiple objects (e.g., number of neighbors).
+We build cp\_measure by extracting the CellProfiler measurements from the code base and grouping them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (e.g., Pearson correlation) and multiple objects (e.g., number of neighbours).
 
@@
 \includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
-\caption{\label{fig:overview}Our tool produces features from images by using the local information in every region of interest.
-This can be used to featurize the pairwise combination of all the available channels.
-The resultant matrices represent the entirety of the experiment in a much more compressed way, and can be processed with standard machine and deep learning methods.}
+\caption{\label{fig:overview}cp\_measure generates features from images by using the local information in every region of interest.
+It is commonly used to featurise the pairwise combination of all the available channels and objects.
+The resultant matrices represent the entire experiment and can be studied using machine and deep learning methods.}
 \end{figure}
 
-The cp\_measure features were based off the CellProfiler implementation, but we separated the actual functional components (i.e., math) from the user-facing and orchestration components.
-It improves reusability, testability and its makes it possible for more people to access and improve its internals in the future.
-We positioned it to become a stepping stone in extensibility and maintainability.
-By isolating and cleaning the implemented mathematics of CellProfiler we vastly reduce the amount of time and manual effort required to perform data analyses while retaining the features present in numerous datasets.
+By separating the numerical from the user-facing and orchestration components we improved reusability, testability and extensibility.
+We positioned it to become a stepping stone to make the morphological profiling numerical methods accessible to the data science community.
+Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to perform data analyses, while still retaining the standard features already present in numerous datasets.
 
-Our aim is to remain consistent with the current scientific python ecosystem.
-Its interface corresponds that of the widely used scikit-image \citep{waltScikitimageImageProcessing2014}.
+Our aim is to remain consistent with the current scientific Python ecosystem.
+Its interface corresponds that of the widely used scikit-image.
 This greatly reduces the effort needed to integrate it in existing workflows and tools.
@@
 We first performed the numerical validation of cp\_measure, relative to the original CellProfiler features.
-In preparation, we collected 300 images corresponding to 150 perturbations from the JUMP dataset.
-the most significant phenotypes for a given measurement each.
+For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each.
 To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features.
-Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CelProfiler.
-Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value.
+Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler.
+Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.
 
@@
 Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis.
-While it was posisble to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
+While it was possible to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
 \subsubsection{Spatial transcriptomics}
@@
 Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy.
-Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artifacts.
+Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts.
 Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores.
@@
 This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
-We analyzed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/).
+We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/).
 Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
 Morphological features were extracted from these 5-channel images for both datasets.
-Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and PyCytoMiner (CITE) respectively.
+Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and Pycytominer (CITE) respectively.
 We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig.
@@
 Long-term, we envision cp\_measure can be the place to develop and distribute new measurements.
-While CellProfiler's measurements are already ubiquituous in bioimaging studies, the existing palette of measurements could be further extended to cover unexplored use-cases.
+While CellProfiler's measurements are already ubiquitous in bioimaging studies, the existing palette of measurements could be further extended to cover unexplored use-cases.
 We also see adding community-contributed measurements to better match the current questions scientists pose to imaging data.

### 2025-05-20 draft.tex
docs: add imports

@@
-% Created 2025-05-19 Mon 21:10
+% Created 2025-05-20 Tue 10:41
 % Intended LaTeX compiler: pdflatex
@@
 
-By separating the numerical from the user-facing and orchestration components we improved reusability, testability and extensibility.
+By isolating the numerical from the graphical and orchestration components we improved reusability, testability and extensibility.
 We positioned it to become a stepping stone to make the morphological profiling numerical methods accessible to the data science community.
-Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to perform data analyses, while still retaining the standard features already present in numerous datasets.
+Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already present in numerous datasets.
 
@@
 
-The validity of cp\_measure is shown on Figure \ref{fig:cp_vs_cpmeasure}.
-We show examples of the features on the panel on the left.
-Mostly straight lines demonstrate the recapitulation of measurements from our implementation.
-A few data points fall outside the diagonals, which may indicate that some edge-cases are treated differently.
+The validation of our cp\_measure features is shown on \ref{fig:cp_vs_cpmeasure}.
+The left panel shows examples of the comparison of our features against CellProfiler's.
+The straight lines demonstrate the recapitulation of measurements from our implementation.
+A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool.
 The panel on the right shows the \(R^2\) value of a linear interpolation.
-Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements resulted in a linear relation, regardless of whether the masks were for nuclei or cytosols.
+Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols.
 This result provides reassurance that cp\_measure can be confidently in cases where CellProfiler would be used.
@@
 With this we identified which features distinguish cells on the later samples and distinguish subpopulations.
-Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells [CITE Shapley].
+Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.
 
@@
 
-To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy (CITE).
-Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData (CITE) framework underlying Squidpy.
+To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy \citep{pallaSquidpyScalableFramework2022}.
+Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData \citep{marconatoSpatialDataOpenUniversal2025} framework underlying Squidpy.
 Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts.
@@
 This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
-We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform (CITE nanostring.com/products/cosmx-spatial-molecular-imager/ffpe-dataset/cosmx-smi-mouse-brain-ffpe-dataset/).
+We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform \citep{CosMxSMIMouse2025}.
 Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
 Morphological features were extracted from these 5-channel images for both datasets.
-Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy (CITE) and Pycytominer (CITE) respectively.
+Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy \citep{wolfSCANPYLargescaleSinglecell2018} and Pycytominer \citep{serranoReproducibleImagebasedProfiling2025} respectively.
 We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig.

### 2025-05-20 draft.tex
fix minor typos
docs: Update draft text

@@
-% Created 2025-05-20 Tue 10:41
+% Created 2025-05-20 Tue 11:00
 % Intended LaTeX compiler: pdflatex
@@
 \caption{\label{fig:overview}cp\_measure generates features from images by using the local information in every region of interest.
-It is commonly used to featurise the pairwise combination of all the available channels and objects.
+It is commonly used to featurize the pairwise combination of all the available channels and objects.
 The resultant matrices represent the entire experiment and can be studied using machine and deep learning methods.}
@@
 
-We first performed the numerical validation of cp\_measure, relative to the original CellProfiler features.
+We first performed the numerical validation of cp\_measure relative to the original CellProfiler features.
 For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each.
@@
 Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols.
-This result provides reassurance that cp\_measure can be confidently in cases where CellProfiler would be used.
+This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
 \subsection{Results}
 \label{sec:orge5b5c6b}
-We showcase a couple of use-cases in which cp\_measure makes our machine-learning workflows faster and integrate better with existing tools.
 \subsubsection{Astrocytes 3D data}
@@
 
-Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on cp\_measure features.
-Our results showcased the axis length of the cell to be a major indicator of phenotypic effect, implying that cells became more elongated on their minor axis.
-While it was possible to CellProfiler for this task, cp\_measure made it much faster and simpler to implement, specially since the segmentation masks were already provided, which is quite common in public datasets.
+Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on our features.
+This indicates that the major axis length of the cell to is an indicator of phenotypic effect over the course of the experiment, implying that cells became more elongated on their minor axis.
+Though unsurprising, as we expect astrocytes to extend over time, cp\_measure was able to recover this with only a few lines of code.
 \subsubsection{Spatial transcriptomics}

### 2025-05-20 draft.tex
docs: Update validation text
Add appendix

@@
-% Created 2025-05-20 Tue 11:00
+% Created 2025-05-20 Tue 11:40
 % Intended LaTeX compiler: pdflatex
@@
 Unless a major change in neural networks occurs, they will still require more complex interpretability methods.
-Unlike learned features, whose interpretation requires complex methods like counterfactuals or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
+Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
 
@@
 
-We first performed the numerical validation of cp\_measure relative to the original CellProfiler features.
+We first performed the numerical validation that cp\_measure features, for a set of images and masks representative of the possible values, match the original CellProfiler features.
 For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each.
@@
 
-The validation of our cp\_measure features is shown on \ref{fig:cp_vs_cpmeasure}.
+The initial validation of our cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}.
 The left panel shows examples of the comparison of our features against CellProfiler's.
@@
 \bibliography{bibliography}
+\section{Appendix}
+\label{sec:orgdd18dd8}
+\subsection{Methods}
+\label{sec:orgb3e9382}
+\subsubsection{Data and software}
+\label{sec:orgbda0ae2}
+The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
+All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}.
+The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.
 \end{document}

### 2025-05-20 draft.tex
Add Alan as correspondig author
remove paragraph out of place
improve first results paragraph

@@
-% Created 2025-05-20 Tue 11:40
+% Created 2025-05-20 Tue 13:14
 % Intended LaTeX compiler: pdflatex
@@
 
+\icmlcorrespondingauthor{Al\'an F. Mu\~{n}oz}{amunozgo@broadinstitute.org}
 \icmlcorrespondingauthor{Shantanu Singh}{shantanu@broadinstitute.org}
@@
 \label{sec:org61842b5}
-We build cp\_measure by extracting the CellProfiler measurements from the code base and grouping them based on their inputs: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (e.g., Pearson correlation) and multiple objects (e.g., number of neighbours).
+In our library cp\_measure we extracted the CellProfiler measurements, removed anything that is not essential to solve the image-to-features problem and categorized the measurenents based on their input: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (e.g., Pearson correlation) and multiple objects (e.g., number of neighbours).
 
@@
 Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already present in numerous datasets.
-
-Our aim is to remain consistent with the current scientific Python ecosystem.
-Its interface corresponds that of the widely used scikit-image.
-This greatly reduces the effort needed to integrate it in existing workflows and tools.
 

### 2025-05-21 draft.tex
[WIP] implement Anne's revisions

@@
 \twocolumn[
-\icmltitle{cp\_measure: Morphological profiling for data scientists}
+\icmltitle{cp\_measure: morphological profiling for data scientists}
 
@@
 \icmlaffiliation{broad}{Broad Institute of MIT and Harvard, United States}
-\icmlaffiliation{hh}{Institute of Computational biology, Helmholtz Zentrum München, Germany}
+\icmlaffiliation{hh}{Institute of Computational Biology, Helmholtz Zentrum München, Germany}
 
@@
 \begin{abstract}
-Quantifying the contents of objects in images is a core challenge in biological imaging.
+Biological image analysis focuses on measuring a particular visual property of interest for cells or other entities. It 
+Quantifying object properties in images is a core challenge in biological imaging.
 The current tools require significant human intervention.
-Here we introduce our library cp\_measure, which provides programmatic access to the most widespread metrics to convert images and objects into features.
-We then demonstrate that the features are consistent to the standard ones and showcase tasks for which our tool is more suitable than the alternatives.
-cp\_measure opens the door to community-driven development and improvement of bioimage analysis metrics and pipelines, increasing the scaling capabilities, reproducibility and accessibility for computational and data scientists.
+Here we introduce an efficient computational library, cp\_measure, which provides programmatic access to the most commonly used metrics to convert images and objects into features.
+The features are consistent with those produced by the popular graphical user interface-based software CellProfiler, and we showcase tasks for which cp\_measure is more suitable than alternatives.
+cp\_measure opens the door to community-driven development and improvement of bioimage analysis metrics and pipelines, increasing the scaling capabilities, reproducibility, and accessibility for computational and data scientists.
 \end{abstract}
@@
 \label{sec:org4c9ba67}
-The field of biological imaging leverages data from cell microscopy images to obtain biological insights.
-Modern microscopes can capture a wide range of information, such as illumination and fluorescence.
-When coupled with fluorescence dyes or proteins biologists are able to observe the location and distribution of cells and organelles.
+Modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components by microscopy.
 
-The main purpose of image analysis in biology is to characterize the cell state or its components.
-Early on, human eyes where the sole judge on this, but a more objective, scalable and reproducible method was required.
-This commonly entailed identifying regions of interest and calculating metrics that represent these regions, such as intensity distributions.
+Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility.
+Now, quantification is standard, with software often identifying regions of interest (such as cells) and calculating metrics that represent these regions, such as intensity distributions.
 
-Morphological profiling is a technique where we generate an array of shape and intensity features for all objects, cells or otherwise.
-These features are fed into statistical or machine learning methods to identify patterns.
-One of its biggest applications is drug discovery, where scientists leverages microscopy's low acquisition cost and high throughput \citep{sealDecadeSystematicReview2024}.
-Morphological profiling serves not only as validation but also as information orthogonal to other high throughput methods, such as genomic sequencing.
+Morphological profiling is a technique measuring an array of shape and intensity features for all biological objects, such as cells.
+These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.
+One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.
 \subsection{The current state of bioimage analysis}
 \label{sec:org8f5b33d}
-The most widely used software to process biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
-Its intended audience is experimental biologists with scant programmatic experience, for whom it has been a boon.
-In contrast, its usefulness greatly diminishes for computer-savvy scientists who fear no command line.
-For them, CellProfiler means friction in pipelines comprised of multiple tools.
-It caters for manually-defined workflows instead of low-customization high throughput analyses, which are becoming prevalent in the age of generalist segmentation and foundational models.
+The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
+It can be used by experimental biologists with scant programming experience, for whom it has been a boon.
+By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools.
+CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
 
-Integrating a CellProfiler pipelines with other tools is a struggle.
-Building plug-ins to add functionality is time and effort-consuming challenges that require an understanding of its interfaces.
-Treating it as a black box and using its output precludes the addition of custom-made image filters.
-Lastly, CellProfiler depends on many number of packages from both Python and Java, amplifying the risk of dependency conflicts.
+From the data-science perspective, there are multiple challenges lessening the attractiveness of CellProfiler.
+First, the integration of CellProfiler with other tools can be a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
+Additionally, its usage precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
+Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts.
 Containers mitigate this problem, but not without increased complexity and caveats.
-To summarize, CellProfiler is an all-or-nothing tool, whose extensibility and integration with other tools demands more time and effort than most scientists can spare.
 
-Object featurization is not exclusive of CellProfiler.
-For example, scikit-image can produce some overlapping features \citep{waltScikitimageImageProcessing2014}.
-Tools like ScaleFex or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} can generate features, but they have limitations: First, their implementation is completely independent and thus cannot reproduce existing datasets, potentially impacting the biological interpretations.
+Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} .
+However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery, potentially impacting the biological interpretations.
 Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image.
-Some alternatives, such as ScaleFex, are designed for cloud usage first, and may neglect a significant number of cases where local compute suffices.
-We thus believe that there is the need for a tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
+Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.
+
 \subsection{Engineered features in the era of deep learning}
 \label{sec:org9dc3dfa}
-The latest deep learning models can streamline image processing.
-Nevertheless, their capacity to provide biological interpretation is limited \citep{moenDeepLearningCellular2019}.
-In some cases, engineered features can even outperform learned ones \citep{tangMorphologicalProfilingDrug2024}.
+An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features.
+Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features, but not always \citep{tangMorphologicalProfilingDrug2024}.
+Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images.
+Trained models can in some cases be used on new datasets, eliminating the training step.
+Generally deep learning-based feature extraction takes less compute time than CellProfiler does; however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
 Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data.
 Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data.
-Unless a major change in neural networks occurs, they will still require more complex interpretability methods.
 Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
 
-The intended audience for such a featurization tool would be, in addition to biologists, data scientists striving to achieve reproducible data analyses within a few lines of code.
-As a whole, engineered features still have the role of, at the minimum, complementing deep learning ones.
-The bioimage software ecosystems lacks a modular tool to convert images into interpretable features with no compulsory graphical interfaces or cloud compute.
-Ideally it should also harmonize with other scientific libraries.
-With all this in mind we have developed cp\_measure.
+In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses within a few lines of code, whether locally or in the cloud.
+We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
+With all this in mind we developed cp\_measure.
+
 \section{Extracting interpretable features}
 \label{sec:org61842b5}
-In our library cp\_measure we extracted the CellProfiler measurements, removed anything that is not essential to solve the image-to-features problem and categorized the measurenents based on their input: One object and one imaging channel (e.g., intensity, texture), one object and multiple channels (e.g., Pearson correlation) and multiple objects (e.g., number of neighbours).
+The library cp\_measure is  the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
+Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to measure such metrics as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).
 
@@
 \includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
-\caption{\label{fig:overview}cp\_measure generates features from images by using the local information in every region of interest.
+\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
 It is commonly used to featurize the pairwise combination of all the available channels and objects.
-The resultant matrices represent the entire experiment and can be studied using machine and deep learning methods.}
+The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
 \end{figure}
 
-By isolating the numerical from the graphical and orchestration components we improved reusability, testability and extensibility.
-We positioned it to become a stepping stone to make the morphological profiling numerical methods accessible to the data science community.
-Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already present in numerous datasets.
+By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility.
+We aim to make the morphological profiling numerical methods accessible to the data science community.
+Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for numerous public datasets.
 
-We will now demonstrate the validity and capabilities of cp\_measure in cases that would normally require domain expertise and a considerable amount of time.
+%First, we demonstrated the validity and capabilities of cp\_measure in cases that would normally require image analysis knowledge and a considerable amount of time.
 First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}.
@@
 These use-cases demonstrate its widespread suitability for different types of problems.
-\subsection{Recapitulating CellProfiler measurements}
+\subsection{Reproducing CellProfiler measurements}
 \label{sec:org09b0cd2}
@@
 \caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
-\textbf{Left panel.} Representative examples comparing Cellprofiler feature values to cp\_measure's, generated using matching pairs of masks and images.
+\textbf{Left panel.} Representative examples comparing CellProfiler feature values to cp\_measure's, generated using matching pairs of masks and images.
 \textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
@@
 
-We first performed the numerical validation that cp\_measure features, for a set of images and masks representative of the possible values, match the original CellProfiler features.
-For this we collected 300 images corresponding to 150 perturbations from the JUMP dataset, selecting the most significant phenotypes for a given measurement each.
-To ensure that we are using identical object masks, we segmented these images to obtain the cells and nuclei using CellProfiler, providing object masks and their associated features.
-Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler.
-Lastly, we calculated a linear fit for the matched features and calculated their \(R^2\) value, indicating how well it fits a linear slope.
+We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
+These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given measurement each.
+We segmented images to obtain the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
+Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler, we calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
 
-The initial validation of our cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}.
+The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}.
 The left panel shows examples of the comparison of our features against CellProfiler's.
 The straight lines demonstrate the recapitulation of measurements from our implementation.
-A few data points fall outside the diagonals, which hint at some edge-cases are treated differently by either tool.
+A few data points fall outside the diagonals, which hint that some edge-cases are treated differently by either tool.
 The panel on the right shows the \(R^2\) value of a linear interpolation.
-Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytosols.
+Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm.
 This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
@@
 \label{sec:orge5b5c6b}
-\subsubsection{Astrocytes 3D data}
+\subsubsection{Astrocyte 3D data}
 \label{sec:org447090b}
 
-As a demonstration of its ease of use and the value of interpretable features, we used cp\_measure in a standard classification workflow.
-We processed 433 3D images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}.
-We preprocessed the data following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired.
+To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow.
+We processed 433 three-dimensional (3D) images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}.
+We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
+Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells.
 With this we identified which features distinguish cells on the later samples and distinguish subpopulations.
@@
 \includesvg[width=.9\linewidth]{./figs/example_shap}
-\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astroctyes image and masks.
+\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks.
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
@@
 
-Figure \ref{fig:astrocytes} shows an example image and object masks alongside the Shapley values of a classifier trained on our features.
+Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features.
 This indicates that the major axis length of the cell to is an indicator of phenotypic effect over the course of the experiment, implying that cells became more elongated on their minor axis.
-Though unsurprising, as we expect astrocytes to extend over time, cp\_measure was able to recover this with only a few lines of code.
+Though unsurprising, as we expect astrocytes to extend over time, cp\_measure was able to uncover this with only a few lines of code.
 \subsubsection{Spatial transcriptomics}
@@
 \label{sec:orgf37b369}
-The usage of image analysis pipelines that require manual setups hinders reproducibility and hinders our ability to compare different datasets.
-In this work we introduced our new library cp\_measure, which provides widely used engineered features and enables simpler automated analyses of microscopy data in either short scripts and complex pipelines.
-This also removes the requirement of using graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy even without cloud infrastructure.
+While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows.
+In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
+It enables simpler automated analyses of microscopy data in either short scripts and complex pipelines.
+It avoids graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
 
-The biologically interpretable features provided by cp\_measure complement deep learning ones and offer a better mechanistic understanding of the underlying biology.
-When used in tandem with generalist tools it enables more insightful pipelines that leverage machine and deep learning approaches.
+The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology.
+When used in tandem with generalist image-processing tools, such as Cellpose for segmentation (CITE Cellpose), we can streamline machine and deep learning workflows.
 
-These measurements have already been used in non-biological contexts, such as environmental monitoring \citep{ideharaExploringNileRed2025}, thus these engineered metrics also benefit other scientific fields beyond morphological profiling.
+Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
+We therefore foresee cp\_measure being useful for a broad scientific community beyond morphological profiling.
 \section{Future work}
@@
 The most obvious way to make cp\_measure more useful is to contribute it back to CellProfiler.
-This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity of formalizing the inputs and outputs of all measurements.
+This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
 
-Developing a comprehensive tests suite will guarantee mathematical correctness, which currently not even CellProfiler has.
-This test suite in turn would in turn expedite improvements in multiple ways: Firstly, optimizing the most compute-consuming features, such as granularity.
-Later on, we could add to support just-in-time compiling and GPUs.
-
-Long-term, we envision cp\_measure can be the place to develop and distribute new measurements.
-While CellProfiler's measurements are already ubiquitous in bioimaging studies, the existing palette of measurements could be further extended to cover unexplored use-cases.
-We also see adding community-contributed measurements to better match the current questions scientists pose to imaging data.
+We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking.
+A functioning test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity.
+Once tests are in place, we could add to support just-in-time compiling and GPUs.
+Finally, we envision the cp\_measure library could be the place to develop and distribute new measurements from the community.
 
@@
 The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.
+\subsubsection{Data and software}
+\label{sec:acknowledgements}
+
+This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC).
+The authors would like to thank Nodar Gogoberidze for his valuable insight into CellProfiler's code base, and Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
 \end{document}

### 2025-05-21 draft.tex
Updates from Overleaf

@@
 An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features.
-Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features, but not always \citep{tangMorphologicalProfilingDrug2024}.
+Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
 Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images.
 Trained models can in some cases be used on new datasets, eliminating the training step.
+Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data.
+Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data \cite{aza}.
 Generally deep learning-based feature extraction takes less compute time than CellProfiler does; however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
-Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data.
-Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data.
 Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.

### 2025-05-21 draft.tex

@@
 Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} .
-However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery, potentially impacting the biological interpretations.
+However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}, potentially impacting the biological interpretations.
 Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image.
@@
 Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data.
-Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution data \cite{aza}.
-Generally deep learning-based feature extraction takes less compute time than CellProfiler does; however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
+Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}.
+Generally deep learning-based feature extraction takes less wall time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
 Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
@@
 \label{sec:orge5b5c6b}
-\subsubsection{Astrocyte 3D data}
+\subsubsection{Astrocyte 3D nuclei data}
 \label{sec:org447090b}
@@
 To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow.
-We processed 433 three-dimensional (3D) images of astrocytes containing 831 cells \citep{kalinin3DCellNuclear2018}.
+We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}.
 We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
 Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells.
-With this we identified which features distinguish cells on the later samples and distinguish subpopulations.
+We achieved 87\% classification accuracy in the testing dataset.
+Using the  we identified which features distinguish cells on the later samples and distinguish subpopulations.
 Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.
@@
 \label{sec:org5711d86}
-A key advantage of providing these measurements as a standalone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
+A key advantage of providing these measurements as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
 The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality.
@@
 To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy \citep{pallaSquidpyScalableFramework2022}.
-Being standalone allowed seamless incorporation into workflows powered by the robust SpatialData \citep{marconatoSpatialDataOpenUniversal2025} framework underlying Squidpy.
+Being stand-alone allowed seamless incorporation into workflows powered by the robust SpatialData \citep{marconatoSpatialDataOpenUniversal2025} framework underlying Squidpy.
 Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts.

### 2025-05-21 draft.tex

@@
 Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} .
-However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}, potentially impacting the biological interpretations.
+However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}, potentially impacting the biological interpretations.
 Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image.
@@
 Trained models can in some cases be used on new datasets, eliminating the training step.
-Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require annotating data.
+Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset.
 Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}.
-Generally deep learning-based feature extraction takes less wall time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
+Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
 Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
@@
 We aim to make the morphological profiling numerical methods accessible to the data science community.
-Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for numerous public datasets.
+Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for multiple public datasets.
 

### 2025-05-21 draft.tex
text ready for shantanu

@@
 Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}.
-Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{moenDeepLearningCellular2019}.
+Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.
 Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
@@
 We aim to make the morphological profiling numerical methods accessible to the data science community.
-Given that the numerical internals are more transparent, we reduced the amount of time and human effort required to integrate it into pipelines, while still retaining the standard features already extracted for multiple public datasets.
+Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.
 
@@
 The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}.
+We find that most features had a $R^2$ correlation larger than 90\%.
 The left panel shows examples of the comparison of our features against CellProfiler's.
 The straight lines demonstrate the recapitulation of measurements from our implementation.
-A few data points fall outside the diagonals, which hint that some edge-cases are treated differently by either tool.
+Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool.
 The panel on the right shows the \(R^2\) value of a linear interpolation.
 Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm.
-This result provides evidence that cp\_measure can be confidently in cases where CellProfiler would be used, and the insights are unlikely to change.
+This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
 \subsection{Results}
@@
 We achieved 87\% classification accuracy in the testing dataset.
-Using the  we identified which features distinguish cells on the later samples and distinguish subpopulations.
+Using the morphological profiles, we identified which features distinguish cells based on the time at which the image was taken.
 Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.
@@
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
-\textbf{Bottom panel.} Shapley values of the most important features to classify the day in which an image was taken (out of three).
-The test data accuracy is shown in bold.}
+\textbf{Bottom panel.} Shapley values of the most important features to classify the day of differentiation based on the images (out of three).
+Each point is an image containing multiple nuclei.
+The test data accuracy is shown in bold.
+The x-axis indicates how impactful a given feature is on the classification accuracy.
+The y-axis are the most significant features, and the last entry is the rest of the features grouped.}
 \end{figure}
@@
 Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features.
-This indicates that the major axis length of the cell to is an indicator of phenotypic effect over the course of the experiment, implying that cells became more elongated on their minor axis.
-Though unsurprising, as we expect astrocytes to extend over time, cp\_measure was able to uncover this with only a few lines of code.
+This indicates that the minor axis length of the nuclei is an indicator of phenotypic effect over the course of the experiment, implying that nuclei becomes thicker (though not necessarily more round).
+Though unsurprising, as we expect astrocyte nuclei to extend over time, by using cp\_measure was able to uncover this with relatively few lines of code.
 \subsubsection{Spatial transcriptomics}
@@
 In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
-It enables simpler automated analyses of microscopy data in either short scripts and complex pipelines.
+Although there are still rough edges to polish, such as ensuring a 1:1 match, it enables simpler automated analyses of microscopy data in either short scripts and complex pipelines.
 It avoids graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
@@
 \label{sec:org5cdbb12}
-The most obvious way to make cp\_measure more useful is to contribute it back to CellProfiler.
-This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
+We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits: This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
+It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the original community for which it was originally developed.
 

### 2025-05-22 draft.tex
implement shantanu's suggestions

@@
 \twocolumn[
-\icmltitle{cp\_measure: morphological profiling for data scientists}
+\icmltitle{cp\_measure: API-first feature extraction for image-based profiling workflows}
 
@@
 \begin{abstract}
-Biological image analysis focuses on measuring a particular visual property of interest for cells or other entities. It 
-Quantifying object properties in images is a core challenge in biological imaging.
-The current tools require significant human intervention.
-Here we introduce an efficient computational library, cp\_measure, which provides programmatic access to the most commonly used metrics to convert images and objects into features.
-The features are consistent with those produced by the popular graphical user interface-based software CellProfiler, and we showcase tasks for which cp\_measure is more suitable than alternatives.
-cp\_measure opens the door to community-driven development and improvement of bioimage analysis metrics and pipelines, increasing the scaling capabilities, reproducibility, and accessibility for computational and data scientists.
+Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities.
+A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles where patterns of changes show biological properties.
+While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows.
+Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
+We demonstrate that cp\_measure features retain high fidelity with CellProfiler measurements while enabling seamless integration with the scientific Python ecosystem.
+Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
 \end{abstract}
+
 \section{Introduction}
@@
 Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility.
-Now, quantification is standard, with software often identifying regions of interest (such as cells) and calculating metrics that represent these regions, such as intensity distributions.
+Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
+We use the terms "features" and "measurements" interchangeably throughout this paper.
 
-Morphological profiling is a technique measuring an array of shape and intensity features for all biological objects, such as cells.
+Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for all biological objects, such as cells.
 These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.
 One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.
+
+\begin{figure}[htbp]
+\centering
+\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
+\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
+It is commonly used to featurize the pairwise combination of all the available channels and objects.
+The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
+\end{figure}
+
 \subsection{The current state of bioimage analysis}
@@
 From the data-science perspective, there are multiple challenges lessening the attractiveness of CellProfiler.
-First, the integration of CellProfiler with other tools can be a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
-Additionally, its usage precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
+First, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
+Its usage also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
 Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts.
@@
 The library cp\_measure is  the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
-Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to measure such metrics as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).
-
-\begin{figure}[htbp]
-\centering
-\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
-\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
-It is commonly used to featurize the pairwise combination of all the available channels and objects.
-The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
-\end{figure}
+Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).
 
 By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility.
-We aim to make the morphological profiling numerical methods accessible to the data science community.
+We aim to make the image-based profiling numerical methods accessible to the data science community.
 Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.
@@
 We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
-These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given measurement each.
+These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given feature each.
 We segmented images to obtain the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
@@
 The left panel shows examples of the comparison of our features against CellProfiler's.
-The straight lines demonstrate the recapitulation of measurements from our implementation.
+The straight lines demonstrate the recapitulation of features from our implementation.
 Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool.
 The panel on the right shows the \(R^2\) value of a linear interpolation.
-Given that this value is directly correlated to the correctness of the implementation, we can see that most of our measurements have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm.
+Given that this value is directly correlated to the correctness of the implementation, we can see that most of our features have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm.
 This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
@@
 We achieved 87\% classification accuracy in the testing dataset.
-Using the morphological profiles, we identified which features distinguish cells based on the time at which the image was taken.
-Finally, we calculated the Shapley values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.
+Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken.
+Finally, we calculated the SHAP values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.
 
@@
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
-\textbf{Bottom panel.} Shapley values of the most important features to classify the day of differentiation based on the images (out of three).
-Each point is an image containing multiple nuclei.
-The test data accuracy is shown in bold.
-The x-axis indicates how impactful a given feature is on the classification accuracy.
-The y-axis are the most significant features, and the last entry is the rest of the features grouped.}
+\textbf{Bottom panel.}
+
+SHAP values of the most important features for classifying the day of differentiation (out of three days). 
+Each point represents a single 3D image containing multiple nuclei. 
+Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value. 
+The test data accuracy is shown in bold.}
 \end{figure}
@@
 \label{sec:org5711d86}
-A key advantage of providing these measurements as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
+A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
 The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality.
 However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment.
-In contrast, classical morphological measurements yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.
+In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.
 
@@
 Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
-We therefore foresee cp\_measure being useful for a broad scientific community beyond morphological profiling.
+We therefore foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
 \section{Future work}

### 2025-05-22 src/astro3d/get_figures.py
figs: increase size of nuclei igs in shapley figure

@@ -19,12 +19,15 @@ from xgboost import XGBClassifier
 MEASUREMENTS = get_core_measurements()
 
 overwrite = False
-profiles_fpath = Path("tmp.parquet")
 figs_path = Path(".") / ".." / ".." / "figs"
+data_path = Path("/") / "datastore" / "alan" / "cp_measure"
 figs_path.parent.mkdir(parents=True, exist_ok=True)
+data_path.parent.mkdir(parents=True, exist_ok=True)
+profiles_fpath = data_path / Path("astro3d_profiles.parquet")
+
 target_feature = "day"
 
-cell_count_col = "Cell Count"
+cell_count_col = "Nuclei Count"
 seed = 42
 
 
@@ -101,44 +104,45 @@ url_hash = {
     "https://www.socr.umich.edu/data/3d-cell-morphometry-data/vpa_day7_masks.tar.gz": "63bf6d123ee9505a51fb0ef099ba77223a800babab51756d577cd7cfd8dd182d",
 }
 
-if not profiles_fpath.exists() or overwrite:
-    retrieved = list(
-        Parallel(n_jobs=-1)(
-            delayed(retrieve)(
-                url,
-                processor=Untar(extract_dir=Path(url).name.split(".")[0]),
-                known_hash=h,
-                progressbar=True,
-            )
-            for url, h in url_hash.items()
+retrieved = list(
+    Parallel(n_jobs=-1)(
+        delayed(retrieve)(
+            url,
+            processor=Untar(extract_dir=Path(url).name.split(".")[0]),
+            known_hash=h,
+            progressbar=True,
         )
+        for url, h in url_hash.items()
     )
+)
 
-    dir_files = {
-        x: y
-        for x, y in zip((Path(url).name.split(".")[0] for url in url_hash), retrieved)
-        if x.endswith("_masks")  # Optional but makes the next step faster
-    }
+dir_files = {
+    x: y
+    for x, y in zip((Path(url).name.split(".")[0] for url in url_hash), retrieved)
+    if x.endswith("_masks")  # Optional but makes the next step faster
+}
 
-    pairs = [
-        x
-        for x in Parallel(n_jobs=-1)(
-            delayed(map_mask_to_image)(x) for x in chain(*dir_files.values())
-        )
-        if x is not None  # Ignore pngs
-    ]
[…]

### 2025-05-22 src/jump/plot_cp_vs_cpm.py
scripts: add scripts to generate fig 2

@@ -1,0 +1,242 @@
+#!/usr/bin/env jupyter
+"""
+Compare Cellprofiler and cp_measure data.
+Plot the comparison.
+"""
+
+from pathlib import Path
+
+import duckdb
+import matplotlib.pyplot as plt
+import polars as pl
+import polars as pls
+import pooch
+import seaborn as sns
+from duckdb.typing import VARCHAR
+from parse_features import get_feature_groups
+from polars import selectors as cs
+from pooch import Unzip
+from util_plot import generate_label
+
+figs_dir = Path("..") / ".." / "figs"
+
+# cpmeasure_parquet = "/datastore/alan/cp_measure/profiles/first_set.parquet"
+cpmeasure_parquet = "/datastore/alan/cp_measure/profiles_via_masks/first_set.parquet"
+
+
+# Turns out it is easier to trim down the CellProfiler measurements
+# into cp_measure than anything else
+def trim_features(name: str) -> str:
+    """
+    Trim object and channel information from CellProfiler
+    """
+    replacement_obj = ("Cells_", "Image_", "Nuclei_", "Cytoplasm_")
+    replacement_ch = [f"_Orig{x}" for x in ("DNA", "AGP", "ER", "Mito", "RNA")]
+    replacement_mea = ["AreaShape_"]
+
+    for obj in replacement_obj:
+        name = name.replace(obj, "")
+    for ch in replacement_ch:
+        name = name.replace(ch, "")
+    for mea in replacement_mea:
+        name = name.replace(mea, "")
+    return name
+
+
+con = duckdb.connect()
+con.create_function("trim_features", trim_features, [VARCHAR], VARCHAR)
+
+cp_data = pooch.retrieve(
+    "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
+    known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
+    processor=Unzip(),
+)
+
+real_files = [x for x in cp_data if not Path(x).name.startswith(".")]
+
+# %%
+
+sql_files = [(Path(x).parent.parent.name, x) for x in cp_data if x.endswith("db")]
+con.sql("INSTALL sqlite;")
+con.sql("LOAD sqlite;")
+for name, sql_file in sql_files:
+    con.sql(f"ATTACH '{sql_file}' AS {name} (TYPE sqlite);")
+
[…]

### 2025-05-22 src/jump/util_names.py

@@ -1,0 +1,94 @@
+def get_cpm_to_measurement_mapper():
+    names = {
+        "Area": "SizeShape",
+        "BoundingBoxArea": "SizeShape",
+        "BoundingBoxMaximum_X": "SizeShape",
+        "BoundingBoxMaximum_Y": "SizeShape",
+        "BoundingBoxMinimum_X": "SizeShape",
+        "BoundingBoxMinimum_Y": "SizeShape",
+        "Center_X": "SizeShape",
+        "Center_Y": "SizeShape",
+        "Compactness": "SizeShape",
+        "ConvexArea": "SizeShape",
+        "Eccentricity": "SizeShape",
+        "EquivalentDiameter": "SizeShape",
+        "EulerNumber": "SizeShape",
+        "Extent": "SizeShape",
+        "FormFactor": "SizeShape",
+        "Granularity_1": "Granularity",
+        "Granularity_10": "Granularity",
+        "Granularity_11": "Granularity",
+        "Granularity_12": "Granularity",
+        "Granularity_13": "Granularity",
+        "Granularity_14": "Granularity",
+        "Granularity_15": "Granularity",
+        "Granularity_16": "Granularity",
+        "Granularity_2": "Granularity",
+        "Granularity_3": "Granularity",
+        "Granularity_4": "Granularity",
+        "Granularity_5": "Granularity",
+        "Granularity_6": "Granularity",
+        "Granularity_7": "Granularity",
+        "Granularity_8": "Granularity",
+        "Granularity_9": "Granularity",
+        "Intensity_IntegratedIntensity": "Intensity",
+        "Intensity_IntegratedIntensityEdge": "Intensity",
+        "Intensity_LowerQuartileIntensity": "Intensity",
+        "Intensity_MADIntensity": "Intensity",
+        "Intensity_MassDisplacement": "Intensity",
+        "Intensity_MaxIntensity": "Intensity",
+        "Intensity_MaxIntensityEdge": "Intensity",
+        "Intensity_MeanIntensity": "Intensity",
+        "Intensity_MeanIntensityEdge": "Intensity",
+        "Intensity_MedianIntensity": "Intensity",
+        "Intensity_MinIntensity": "Intensity",
+        "Intensity_MinIntensityEdge": "Intensity",
+        "Intensity_StdIntensity": "Intensity",
+        "Intensity_StdIntensityEdge": "Intensity",
[…]

### 2025-05-22 src/jump/utils.py

@@ -1,0 +1,75 @@
+from pathlib import Path
+
+import numpy as np
+import polars as pl
+from cp_measure.bulk import (
+    get_core_measurements,
+    # get_correlation_measurements,
+    # get_multimask_measurements,
+)
+from skimage.io import imread
+
+MEASUREMENTS = get_core_measurements()
+
+
+def read_labels(mask_path: Path):
+    if str(mask_path).endswith("npz"):
+        labels = np.load(mask_path)["arr_0"]
+    else:
+        labels = imread(mask_path)
+    return labels
+
+
+def apply_measurements(
+    mask_path: Path, img_path: Path, object_name: str = None
+) -> pl.DataFrame:
+    gene, site, channel = img_path.stem.split("_")[:3]
+    if object_name is None:
+        object_name = mask_path.stem.split("_")[2]
+
+    labels = read_labels(mask_path)
+
+    img = imread(img_path)
+    if img.ndim == 3:
+        flat_img = img.max(axis=0)
+        flat_img = flat_img / flat_img.max()
+    d = {}
+    for meas_name, meas_f in MEASUREMENTS.items():
+        measurements = meas_f(labels, img)
+        # Unpack output dictionaries
+        for k, v in measurements.items():
+            d[k] = v
+            d["object"] = object_name
+            d["gene"] = gene
+            d["site"] = site
+            d["channel"] = channel
+
+    df = pl.from_dict(d)
+
+    return df
+
+
+def get_keys(fpath: Path, n: int = 2) -> tuple[str]:
+    return tuple(fpath.stem.split("_")[:n])
+
+
+def apply_measurements_type2(
+    mask_path: Path, pixels1_path: Path, pixels2_path: Path
+) -> pl.DataFrame:
+    labels = read_labels(mask_path)
+    pixels1 = imread(pixels1_path)
+    pixels2 = imread(pixels2_path)
+
+    for meas_name, meas_f in MEASUREMENTS_TYPE2.items():
+        measurements = meas_f(labels, pixels1, pixels1)
+        # Unpack output dictionaries
+        for k, v in measurements.items():
+            d[k] = v
+            # for k, v in zip(("pert", "day", "stem"), meta):
+            #     d[k] = v
+            d["object"] = mask_path.stem.split("_")[2]
+            gene, site, channel1 = pixels1_path.stem.split("_")[:3]
+            d["gene"] = gene
+            d["site"] = site
+            channel2 = pixels2_path.stem.split("_")[2]
+            d["channel"] = (channel1, channel2)

### 2025-05-23 draft.tex
Updates from Overleaf

@@
 Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
-We use the terms "features" and "measurements" interchangeably throughout this paper.
+We use the terms "features" and "measurements" interchangeably here.
 
-Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for all biological objects, such as cells.
+Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for a population of biological objects, such as cells.
 These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.
@@
 \caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
-It is commonly used to featurize the pairwise combination of all the available channels and objects.
+It can featurize the pairwise combination of all the available channels (colours) and objects.
 The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
@@
 CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
-
-From the data-science perspective, there are multiple challenges lessening the attractiveness of CellProfiler.
-First, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
-Its usage also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
+For example, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
+The usage of CellProfiler also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
 Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts.
@@
 Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} .
-However, these have limitations: First, their implementation is completely independent from CellProfiler and thus cannot reproduce existing datasets produced using CellProfiler, such as the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}, potentially impacting the biological interpretations.
+However, these have limitations: First, their implementation is completely independent from CellProfiler and thus they do not produce the same feature set as CellProfiler, precluding matching to existing datasets such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
 Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image.
@@
 Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset.
-Foundational models, whose goal is to lower this barrier, fail to generalize well for out-of-distribution (that which comes from different phenomena from the training datasets) data \cite{azadFoundationalModelsMedical2023}.
+Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}.
 Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.
@@
 
-In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses within a few lines of code, whether locally or in the cloud.
+In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses with a few lines of code, whether locally or in the cloud.
 We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
-With all this in mind we developed cp\_measure.
+With all this in mind, we developed cp\_measure.
 
@@
 
-By isolating the feature calculations from the graphical interface and orchestration components, we improved reusability, testability and extensibility.
-We aim to make the image-based profiling numerical methods accessible to the data science community.
-Given that the numerical internals are more transparent since they are freed from all the orchestration code, we reduced the amount of time and human effort required to integrate it into existing and new pipelines, while still retaining the standard features already extracted for multiple public datasets.
+By isolating the feature calculations from the graphical interface and orchestration components, we aimed to make them accessible to the data science community while improving reusability, testability and extensibility.
+This reduces the amount of time and human effort required to integrate featurization into existing and new pipelines, while still retaining the standard set of features available for multiple public datasets.
 
@@
 We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
-These correspond to 150 perturbations from the JUMP Cell Painting dataset\citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for a given feature each.
-We segmented images to obtain the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
-Next, we applied cp\_measure on these masks with the original images and mapped the features from cp\_measure to CellProfiler, we calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
+These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature.
+We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
+Next, we applied cp\_measure on these masks and the original images.
+Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
 
-The validation of cp\_measure features is shown on Figure \ref{fig:cp_vs_cpmeasure}.
-We find that most features had a $R^2$ correlation larger than 90\%.
-The left panel shows examples of the comparison of our features against CellProfiler's.
-The straight lines demonstrate the recapitulation of features from our implementation.
-Some data points fall outside the diagonals for a small set of f, which hint that some edge-cases are treated differently by either tool.
-The panel on the right shows the \(R^2\) value of a linear interpolation.
-Given that this value is directly correlated to the correctness of the implementation, we can see that most of our features have a linear relationship, regardless of whether the masks were for nuclei or cytoplasm.
-This result provides indicates that cp\_measure can be used in some cases instead of CellProfiler, and we would not expect a massive change in results.
+We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}).
+We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation.
+For a small set of features, some data points fall off the diagonal, which hint that some edge cases are treated differently by either tool; we plan to investigate these further as part of a larger project to implement unit testing in CellProfiler.
+These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
 \subsection{Results}
@@
 Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken.
-Finally, we calculated the SHAP values to get a better understanding of the effects of the drugs on the cells \citep{sundararajanManyShapleyValues2020}.
+Finally, we calculated the SHAP values \citep{sundararajanManyShapleyValues2020} to get a better understanding of the effects of the drugs on the cells.
 
@@
 \centering
-\includesvg[width=.9\linewidth]{./figs/example_shap}
+\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
 \caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks.
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
-\textbf{Bottom panel.}
-
-SHAP values of the most important features for classifying the day of differentiation (out of three days). 
+\textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days).
 Each point represents a single 3D image containing multiple nuclei.
 Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value.
+Shape features refer to the nucleus and intensity features refer to the DNA staining.
 The test data accuracy is shown in bold.}
@@
 Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features.
-This indicates that the minor axis length of the nuclei is an indicator of phenotypic effect over the course of the experiment, implying that nuclei becomes thicker (though not necessarily more round).
-Though unsurprising, as we expect astrocyte nuclei to extend over time, by using cp\_measure was able to uncover this with relatively few lines of code.
+The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover with relatively few lines of code.
 \subsubsection{Spatial transcriptomics}

### 2025-05-23 draft.tex
Updates from Overleaf

@@
 Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.
-Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts.
+Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
 
@@
 \includesvg[width=.9\linewidth]{./figs/example_shap.svg}
-\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte image and masks.
+\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte nuclei image and masks.
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
 \textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days).
-Each point represents a single 3D image containing multiple nuclei.
+Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier.
 Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value.
@@
 In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
-Although there are still rough edges to polish, such as ensuring a 1:1 match, it enables simpler automated analyses of microscopy data in either short scripts and complex pipelines.
-It avoids graphical interfaces to process microscopy data, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
+Although there are still rough edges to polish, such as ensuring features match 1:1, it enables simpler automated profiling of microscopy data in short scripts and complex pipelines.
+It avoids graphical interfaces, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
 
 The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology.
-When used in tandem with generalist image-processing tools, such as Cellpose for segmentation (CITE Cellpose), we can streamline machine and deep learning workflows.
+When used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.
 
@@
 \label{sec:org5cdbb12}
-We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits: This would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
-It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the original community for which it was originally developed.
+We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits.
+It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
+It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.
 
@@
 A functioning test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity.
-Once tests are in place, we could add to support just-in-time compiling and GPUs.
-Finally, we envision the cp\_measure library could be the place to develop and distribute new measurements from the community.
+Once tests are in place, we could add support for just-in-time compiling and GPUs.
+We envision the cp\_measure library could also be the place to develop and distribute new measurements from the community.
 
@@
 This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC).
-The authors would like to thank Nodar Gogoberidze for his valuable insight into CellProfiler's code base, and Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
+The authors would like to thank Nodar Gogoberidze, Beth Cimini, Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
 \end{document}

### 2025-05-24 src/jump/get_profiles_from_masks.py
fix: adjust comparison, improve figure 2

@@ -3,8 +3,7 @@
 Compare CellProfiler and cp_measure metrics
 """
 
-from concurrent.futures import ThreadPoolExecutor
-from itertools import chain, combinations, groupby, product
+from itertools import chain, groupby, product
 from pathlib import Path
 
 import polars as pl
@@ -21,8 +20,8 @@ def get_mask_name(x):
 
 
 cp_data = pooch.retrieve(
-    "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
-    known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
+    "https://zenodo.org/api/records/15480187/files/cellprofiler_analysis.zip/content",
+    known_hash="42eb17d4b96863815f52764ac6cef5e03f5cd8020697ced7eecca1db7ff25482",
     processor=Unzip(),
 )
 mask_files = [Path(x) for x in cp_data if "masks" in x]
@@ -68,8 +67,8 @@ combs = list(chain(*[list(product(*xy)) for xy in pairs]))
 # Check that shapes match
 for mask, img in combs:
     assert read_labels(mask).shape == imread(img).shape, "error"
-measure_results = Parallel(n_jobs=-1)(
+measure_results_rad = Parallel(n_jobs=100)(
     delayed(apply_measurements)(m, img, get_mask_name(m)) for m, img in tqdm(combs)
 )
-first_set = pl.concat(measure_results)
-first_set.write_parquet(profiles_dir / "first_set.parquet")
+first_set_rad = pl.concat(measure_results_rad)
+first_set_rad.write_parquet(profiles_dir / "first_set.parquet")

### 2025-05-24 src/jump/parse_features.py

@@ -2,14 +2,13 @@
 """Parse feature names to divide it into its components."""
 
 import re
-from functools import cache
 
 import polars as pl
 
 
 def get_feature_groups(
     feature_fullnames: tuple[str],
-    feature_names: tuple[str] = ("Compartment", "Feature", "Channel", "Suffix"),
+    feature_names: tuple[str] = ("Feature", "Channel", "Suffix"),
 ) -> pl.DataFrame:
     """Group features in a consistent manner using a regex.
 
@@ -32,7 +31,6 @@ def get_feature_groups(
     merging channels where necessary.
 
     """
-    masks = "|".join(("Cells", "Nuclei", "Cytoplasm", "Image", "Object"))
     channels = "|".join((
         "DNA",
         "AGP",
@@ -44,7 +42,7 @@ def get_feature_groups(
     chless_feats = "|".join((
         "AreaShape",
         "Neighbors",
-        "RadialDistribution",
+        # "RadialDistribution",
         "Location",
         "Count",
         "Number",
@@ -54,16 +52,17 @@ def get_feature_groups(
         "Threshold",
     ))
 
-    std = re.compile(rf"({masks})_(\S+)_(Orig)?({channels})(_.*)?")
-    chless = re.compile(f"({masks})_?({chless_feats})_?([a-zA-Z]+)?(.*)?")
-
-    results = [(std.findall(x) or chless.findall(x))[0] for x in feature_fullnames]
+    std = re.compile(rf"(\S+)_(Orig)?({channels})(_.*)?")
+    chless = re.compile(f"({chless_feats})_?([a-zA-Z]+)?(.*)?")
+    results = []
+    for x in feature_fullnames:
+        try:
+            results.append((std.findall(x) or chless.findall(x))[0])
+        except:
+            print(x)
+            raise Exception("failed")
     results = [
-        (
-            (x[0], "".join(x[1:3]), "", x[3])
-            if len(x) < 5
-            else (*x[:2], "".join(x[2:4]), x[4])
-        )
+        (("".join(x[:2]), "", x[2]) if len(x) < 4 else (x[0], "".join(x[1:3]), x[3]))
         for x in results
     ]
 

### 2025-05-24 src/jump/plot_cp_vs_cpm.py

@@ -9,18 +9,18 @@ from pathlib import Path
 import duckdb
 import matplotlib.pyplot as plt
 import polars as pl
-import polars as pls
+import polars_ds as pds
 import pooch
 import seaborn as sns
 from duckdb.typing import VARCHAR
 from parse_features import get_feature_groups
 from polars import selectors as cs
 from pooch import Unzip
+from util_names import get_cpm_to_measurement_mapper
 from util_plot import generate_label
 
 figs_dir = Path("..") / ".." / "figs"
 
-# cpmeasure_parquet = "/datastore/alan/cp_measure/profiles/first_set.parquet"
 cpmeasure_parquet = "/datastore/alan/cp_measure/profiles_via_masks/first_set.parquet"
 
 
@@ -30,12 +30,9 @@ def trim_features(name: str) -> str:
     """
     Trim object and channel information from CellProfiler
     """
-    replacement_obj = ("Cells_", "Image_", "Nuclei_", "Cytoplasm_")
     replacement_ch = [f"_Orig{x}" for x in ("DNA", "AGP", "ER", "Mito", "RNA")]
-    replacement_mea = ["AreaShape_"]
+    replacement_mea = ["AreaShape_", "Texture_"]
 
-    for obj in replacement_obj:
-        name = name.replace(obj, "")
     for ch in replacement_ch:
         name = name.replace(ch, "")
     for mea in replacement_mea:
@@ -47,62 +44,63 @@ con = duckdb.connect()
 con.create_function("trim_features", trim_features, [VARCHAR], VARCHAR)
 
 cp_data = pooch.retrieve(
-    "https://zenodo.org/api/records/15426610/files/cellprofiler_analysis.zip/content",
-    known_hash="aef261cf87e5f138ef2dd91b9ed57add1a4d6f997ab14f139c17024f13a610d9",
+    "https://zenodo.org/api/records/15498193/files/cellprofiler_analysis.zip/content",
+    known_hash="9ffbde3814cad15b143036a370ada7521a30ba9818baba94e4ff60545b2ced3e",
     processor=Unzip(),
 )
 
-real_files = [x for x in cp_data if not Path(x).name.startswith(".")]
+all_csv_files = [x for x in cp_data if x.endswith("csv")]
+csv_files = {
[…]

### 2025-05-24 src/jump/util_names.py

@@ -49,8 +49,10 @@ def get_cpm_to_measurement_mapper():
         "K": "Correlaton",
         "Location_CenterMassIntensity_X": "Intensity",
         "Location_CenterMassIntensity_Y": "Intensity",
+        "Location_CenterMassIntensity_Z": "Intensity",
         "Location_MaxIntensity_X": "Intensity",
         "Location_MaxIntensity_Y": "Intensity",
+        "Location_MaxIntensity_Z": "Intensity",
         "MajorAxisLength": "SizeShape",
         "Manders": "Correlaton",
         "MaxFeretDiameter": "SizeShape",

### 2025-05-24 src/jump/utils.py

@@ -31,11 +31,16 @@ def apply_measurements(
 
     img = imread(img_path)
     if img.ndim == 3:
-        flat_img = img.max(axis=0)
-        flat_img = flat_img / flat_img.max()
+        img = img.max(axis=0)
+    # We know that the input data is int16
+    img = (img / 65535).astype(np.float32)
     d = {}
     for meas_name, meas_f in MEASUREMENTS.items():
-        measurements = meas_f(labels, img)
+        if meas_name == "texture":
+            # This has a specific parameter in the data
+            measurements = meas_f(labels, img, scale=15)
+        else:
+            measurements = meas_f(labels, img)
         # Unpack output dictionaries
         for k, v in measurements.items():
             d[k] = v

### 2025-05-24 draft.tex
Updates from Overleaf

@@
 CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
-For example, coupling CellProfiler with other is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
+For example, mix and matching CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
 The usage of CellProfiler also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
@@
 
-Already, cp\_measure  has been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
-We therefore foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
+Already, Cellprofiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
+By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
 \section{Future work}

### 2025-05-24 src/jump/get_profiles_via_segmentation.py
fix: use latest dataset, add texture and granularity, drop nans

@@ -23,7 +23,7 @@ output_dir = Path("/datastore/alan/cp_measure/")
 img_dir = output_dir / "jump_subset"
 mask_dir = output_dir / "jump_masks"
 profiles_dir = output_dir / "profiles"
-for folder in (img_dir, mask_dir, results_dir):
+for folder in (img_dir, mask_dir, profiles_dir):
     folder.mkdir(exist_ok=True, parents=True)
 
 

### 2025-05-24 src/jump/plot_cp_vs_cpm.py

@@ -44,14 +44,16 @@ con = duckdb.connect()
 con.create_function("trim_features", trim_features, [VARCHAR], VARCHAR)
 
 cp_data = pooch.retrieve(
-    "https://zenodo.org/api/records/15498193/files/cellprofiler_analysis.zip/content",
-    known_hash="9ffbde3814cad15b143036a370ada7521a30ba9818baba94e4ff60545b2ced3e",
+    "https://zenodo.org/api/records/15505477/files/cellprofiler_analysis.zip/content",
+    known_hash="1ca6c08955336d15832fc6dc5c15000990f4dd4733e47a06030a416e7ac7a3e9",
     processor=Unzip(),
 )
 
 all_csv_files = [x for x in cp_data if x.endswith("csv")]
 csv_files = {
-    Path(x).stem: x for x in cp_data if Path(x).stem in ("Cells", "Nuclei", "Image")
+    Path(x).stem: x
+    for x in cp_data
+    if Path(x).stem.split("_")[-1] in ("Cells", "Nuclei", "Image")
 }
 
 original_tables_d = {
@@ -59,11 +61,14 @@ original_tables_d = {
     for k, v in csv_files.items()
 }
 # %%
-feature_tables = [v for k, v in original_tables_d.items() if k != "Image"]
+feature_tables = [v for k, v in original_tables_d.items() if not k.endswith("Image")]
 common = set(feature_tables[0].columns)
 for table in feature_tables[1:]:
     common &= set(table.columns)
-orig_profiles = pl.concat([x.select(common) for x in feature_tables])
+orig_profiles = pl.concat([
+    x.select(common).with_columns(pl.col("object").str.strip_prefix("cp_measure_"))
+    for x in feature_tables
+])
 orig_consensus = (
     orig_profiles.group_by("ImageNumber", "object")
     .median()
@@ -84,7 +89,7 @@ orig_unpivot = orig_consensus.unpivot(
 with_parsed = con.sql("SELECT * FROM orig_unpivot NATURAL JOIN parsed")
 
 with_trimmed = con.sql("SELECT *,trim_features(fullname) AS cpm_id FROM with_parsed")
-image_table = original_tables_d["Image"]
+image_table = original_tables_d["cp_measure_Image"]
 imageid_mapper = con.sql(
[…]

### 2025-05-24 src/jump/util_names.py

@@ -1,3 +1,6 @@
+from itertools import product
+
+
 def get_cpm_to_measurement_mapper():
     names = {
         "Area": "SizeShape",
@@ -65,32 +68,147 @@ def get_cpm_to_measurement_mapper():
         "Overlap": "Correlaton",
         "Perimeter": "SizeShape",
         "RWC": "Correlaton",
-        "RadialDistribution_FracAtD_1of4": "IntensityDistribution",
-        "RadialDistribution_FracAtD_2of4": "IntensityDistribution",
-        "RadialDistribution_FracAtD_3of4": "IntensityDistribution",
-        "RadialDistribution_FracAtD_4of4": "IntensityDistribution",
-        "RadialDistribution_MeanFrac_1of4": "IntensityDistribution",
-        "RadialDistribution_MeanFrac_2of4": "IntensityDistribution",
-        "RadialDistribution_MeanFrac_3of4": "IntensityDistribution",
-        "RadialDistribution_MeanFrac_4of4": "IntensityDistribution",
-        "RadialDistribution_RadialCV_1of4": "IntensityDistribution",
-        "RadialDistribution_RadialCV_2of4": "IntensityDistribution",
-        "RadialDistribution_RadialCV_3of4": "IntensityDistribution",
-        "RadialDistribution_RadialCV_4of4": "IntensityDistribution",
         "Solidity": "SizeShape",
-        "Texture_AngularSecondMoment": "Texture",
-        "Texture_Contrast": "Texture",
-        "Texture_Correlation": "Texture",
-        "Texture_DifferenceEntropy": "Texture",
-        "Texture_DifferenceVariance": "Texture",
-        "Texture_Entropy": "Texture",
-        "Texture_InfoMeas1": "Texture",
-        "Texture_InfoMeas2": "Texture",
-        "Texture_InverseDifferenceMoment": "Texture",
-        "Texture_SumAverage": "Texture",
-        "Texture_SumEntropy": "Texture",
-        "Texture_SumVariance": "Texture",
-        "Texture_Variance": "Texture",
         "Zernike": "AreaShape",
+        **{
+            "CentralMoment_0_0": "SizeShape",
+            "CentralMoment_0_1": "SizeShape",
+            "CentralMoment_0_2": "SizeShape",
+            "CentralMoment_0_3": "SizeShape",
+            "CentralMoment_1_0": "SizeShape",
+            "CentralMoment_1_1": "SizeShape",
+            "CentralMoment_1_2": "SizeShape",
[…]

### 2025-05-25 draft.tex
Updates from Overleaf

@@
 Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
-We demonstrate that cp\_measure features retain high fidelity with CellProfiler measurements while enabling seamless integration with the scientific Python ecosystem.
+We demonstrate that cp\_measure features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem.
 Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
@@
 \label{sec:org4c9ba67}
-Modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components by microscopy.
-
-Historically, the human eyes and brain drew conclusions from biological images but this approach is limited in scale, objectivity, and reproducibility.
-Now, quantification is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
-We use the terms "features" and "measurements" interchangeably here.
+High throughput screening of biological phenomena via complex modalities, such as RNA sequencing, is prohibitively expensive, therefore microscopy is an efficient first step.
+Through microscopy, modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components.
+Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
 
@@
 \centering
-\includesvg[width=.9\linewidth]{./figs/cpmeasure_overview}
+\includegraphics[width=.99\linewidth]{./figs/fig_1_tmp.png}
 \caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
@@
 It can be used by experimental biologists with scant programming experience, for whom it has been a boon.
-By contrast, for computer-savvy scientists, its graphical user interface and lack of programmatic access means friction in pipelines comprised of multiple tools.
+By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools.
 CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
-For example, mix and matching CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
-The usage of CellProfiler also precludes the development of custom-made image-processing steps, such as applying an image filter before feature extraction.
+For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
 Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts.
@@
 Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} .
-However, these have limitations: First, their implementation is completely independent from CellProfiler and thus they do not produce the same feature set as CellProfiler, precluding matching to existing datasets such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
-Moreover, a common problem is that these tools enforce a structure for the input data and require wrangling with the data before processing a single image.
+However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
+Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image.
 Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.
@@
 \label{sec:org9dc3dfa}
-An alternative to extracting a defined set of engineered features is to instead use a trained deep learning model to extract features.
+An alternative to extracting a defined set of engineered features is to instead use a pretrained deep-learning model to extract features.
 Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
 Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images.
-Trained models can in some cases be used on new datasets, eliminating the training step.
-Compared to engineered features, deep learning approaches have a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset.
+Compared to engineered features, deep learning approaches may pose a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset, in addition to having access to a GPU.
 Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}.
 Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.
-Unlike learned features, whose interpretation requires complex methods like counterfactual methods or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
+Unlike learned features, whose interpretation requires complex methods like counterfactual techniques or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts, such as the colocalisation of two proteins \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
 
-In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract engineered features while readily integrating with computational workflows to achieve reproducible data analyses with a few lines of code, whether locally or in the cloud.
+In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract interpretable features while readily integrating with computational workflows to achieve reproducible data analyses, whether locally or in the cloud.
 We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
@@
 \label{sec:org61842b5}
-The library cp\_measure is  the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
-Measurements are categorized based on three kinds of input: One object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Pearson correlations between intensity values across channels) and multiple objects (e.g., number of neighbours).
+The library cp\_measure branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
+Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels).
 
@@
 \caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
-\textbf{Left panel.} Representative examples comparing CellProfiler feature values to cp\_measure's, generated using matching pairs of masks and images.
+\textbf{Left panel.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
 \textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
@@
 Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features.
-The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover with relatively few lines of code.
+The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
 \subsubsection{Spatial transcriptomics}
@@
 \label{sec:orgf37b369}
-While point-and-click interfaces open the world of image analysis to many researchers, they can present barriers for smooth adoption into computational workflows.
+While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop.
 In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
-Although there are still rough edges to polish, such as ensuring features match 1:1, it enables simpler automated profiling of microscopy data in short scripts and complex pipelines.
-It avoids graphical interfaces, resulting in better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
+It enables simpler automated profiling of microscopy data in short scripts and complex pipelines.
+The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
 
 The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology.
-When used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.
+A potential workflow would be to use deep learning features to cluster and then use cp\_measure features from these clusters.
+As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.
 
@@
 \label{sec:org5cdbb12}
-We propose in future work to contribute cp\_measure back to CellProfiler, which offers several benefits.
+We propose in future work to make it possible for CellProfiler to import cp\_measure, which offers several benefits.
 It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
-It would also mean that any changes made in the leaner and more approachable cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.
+It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.
 
 We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking.
-A functioning test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity.
+A comprehensive test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity, by providing rapid iteration capabilities.
 Once tests are in place, we could add support for just-in-time compiling and GPUs.
-We envision the cp\_measure library could also be the place to develop and distribute new measurements from the community.
+We envision that cp\_measure could also be the place to develop and distribute new measurements from the community.
 

### 2025-05-25 src/jump/get_profiles_from_masks.py
[WIP]feat: add colocalization measurements

@@ -3,7 +3,7 @@
 Compare CellProfiler and cp_measure metrics
 """
 
-from itertools import chain, groupby, product
+from itertools import chain, combinations, groupby, product
 from pathlib import Path
 
 import polars as pl
@@ -12,7 +12,7 @@ from joblib import Parallel, delayed
 from pooch import Unzip
 from skimage.io import imread
 from tqdm import tqdm
-from utils import apply_measurements, get_keys, read_labels
+from utils import apply_measurements, apply_measurements_2, get_keys, read_labels
 
 
 def get_mask_name(x):
@@ -67,8 +67,28 @@ combs = list(chain(*[list(product(*xy)) for xy in pairs]))
 # Check that shapes match
 for mask, img in combs:
     assert read_labels(mask).shape == imread(img).shape, "error"
-measure_results_rad = Parallel(n_jobs=100)(
-    delayed(apply_measurements)(m, img, get_mask_name(m)) for m, img in tqdm(combs)
-)
-first_set_rad = pl.concat(measure_results_rad)
-first_set_rad.write_parquet(profiles_dir / "first_set.parquet")
+# measure_results = Parallel(n_jobs=100)(
+#     delayed(apply_measurements)(m, img, get_mask_name(m)) for m, img in tqdm(combs)
+# )
+# first_set_rad = pl.concat(measure_results)
+# first_set_rad.write_parquet(profiles_dir / "first_set.parquet")
+
+# %% Type 2
+triads = [  # Triad with mask, img1, img2
+    (mask_ind, *pair_)
+    for masks, channels in pairs
+    for mask_ind, pair_ in product(masks, combinations(channels, 2))
+]
+for m, img1, img2 in triads:
+    shape = try_imread(m).shape
+    assert shape == try_imread(m).shape and shape == try_imread(img2).shape, "error"
+# measure_results = Parallel(n_jobs=100)(
+#     delayed(apply_measurements_2)(m, img1, img2, get_mask_name(m))
+#     for m, img1, img2 in tqdm(triads[:5])
+# )
+# %%
+for m, img1, img2 in tqdm(triads[:5]):
+    tmp = apply_measurements_2(m, img1, img2, get_mask_name(m))
+# measure_results = Parallel(n_jobs=100)(
+#     delayed(apply_measurements_2)(m, img1, img2, get_mask_name(m))
+# )

### 2025-05-25 src/jump/utils.py

@@ -4,12 +4,13 @@ import numpy as np
 import polars as pl
 from cp_measure.bulk import (
     get_core_measurements,
-    # get_correlation_measurements,
+    get_correlation_measurements,
     # get_multimask_measurements,
 )
 from skimage.io import imread
 
 MEASUREMENTS = get_core_measurements()
+MEASUREMENTS_2 = get_correlation_measurements()
 
 
 def read_labels(mask_path: Path):
@@ -58,23 +59,27 @@ def get_keys(fpath: Path, n: int = 2) -> tuple[str]:
     return tuple(fpath.stem.split("_")[:n])
 
 
-def apply_measurements_type2(
-    mask_path: Path, pixels1_path: Path, pixels2_path: Path
+def apply_measurements_2(
+    mask_path: Path, pixels1_path: Path, pixels2_path: Path, object_name: str
 ) -> pl.DataFrame:
     labels = read_labels(mask_path)
     pixels1 = imread(pixels1_path)
     pixels2 = imread(pixels2_path)
 
-    for meas_name, meas_f in MEASUREMENTS_TYPE2.items():
-        measurements = meas_f(labels, pixels1, pixels1)
+    d = {}
+    for meas_name, meas_f in MEASUREMENTS_2.items():
+        measurements = meas_f(labels, pixels1, pixels2)
         # Unpack output dictionaries
+        breakpoint()
         for k, v in measurements.items():
             d[k] = v
-            # for k, v in zip(("pert", "day", "stem"), meta):
-            #     d[k] = v
-            d["object"] = mask_path.stem.split("_")[2]
+            d["object"] = object_name
             gene, site, channel1 = pixels1_path.stem.split("_")[:3]
             d["gene"] = gene
             d["site"] = site
             channel2 = pixels2_path.stem.split("_")[2]
             d["channel"] = (channel1, channel2)
+
+    df = pl.from_dict(d)
+
+    return df

### 2025-05-25 src/astro3d/plot_figures.py
fig: add labels to fig3

@@ -14,6 +14,7 @@ from joblib import Parallel, delayed
 from pooch import Untar, retrieve
 from skimage.io import imread
 from sklearn.model_selection import train_test_split
+from util_plot import generate_label
 from xgboost import XGBClassifier
 
 MEASUREMENTS = get_core_measurements()
@@ -210,8 +211,8 @@ plt.close()
 # %% example figure
 axd = plt.figure(layout="constrained").subplot_mosaic(
     """
-    AC
-    BB
+    AB
+    CC
     """,
 )
 i = 1
@@ -221,13 +222,13 @@ labels = imread(pairs[i][0]).max(axis=0)
 shap.plots.beeswarm(
     explanation,
     max_display=6,
-    ax=axd["B"],
+    ax=axd["C"],
     plot_size=None,
     show=False,
     # color_bar=False,
 )
-axd["B"].set_yticklabels(
-    axd["B"].get_yticklabels(),
+axd["C"].set_yticklabels(
+    axd["C"].get_yticklabels(),
     # rotation=-30,
     ha="right",
     rotation_mode="anchor",
@@ -235,9 +236,9 @@ axd["B"].set_yticklabels(
 axd["A"].imshow(img)
 axd["A"].axis("off")
 axd["A"].set_title("Z-projected image")
-axd["C"].imshow(labels)
-axd["C"].axis("off")
-axd["C"].set_title("Labels")
+axd["B"].imshow(labels)
+axd["B"].axis("off")
+axd["B"].set_title("Labels")
 # plt.text(0.15, -0.5, acc, fontweight="bold")
 from matplotlib.offsetbox import AnchoredText
 
@@ -249,7 +250,19 @@ text_box = AnchoredText(
     borderpad=0,
 )
 
-axd["B"].add_artist(text_box)
-axd["B"].set_aspect(0.15)
+axd["C"].add_artist(text_box)
+
+axd["C"].set_aspect(0.15)
+for ax_id in "AB":
+    axd[ax_id].add_artist(
+        generate_label(
+            ax_id,
+            bbox_to_anchor=(-0.2, 1),
+            bbox_transform=axd[ax_id].transAxes,
+        )
+    )
+axd["C"].add_artist(
+    generate_label("C", bbox_to_anchor=(-0.98, 1.04), bbox_transform=axd["C"].transAxes)
+)
 plt.savefig(figs_path / "example_shap.svg")
 plt.close()

### 2025-05-25 src/jump/plot_cp_vs_cpm.py
figs: update fig2

@@ -144,6 +144,7 @@ merged = con.sql(
 ).pl()
 # %% Plot
 
+# Temporarily ignore granularity
 mpd = merged
 plt.close()
 g = sns.FacetGrid(
@@ -169,7 +170,7 @@ partitioned = {
 }
 dfs = []
 for feat_name, x in partitioned.items():
-    if len(x):
+    if x["cp_measure"].n_unique() > 1 and x["cp_measure"].abs().mean() > 1e-12:
         dfs.append(
             x.select(
                 pds.lin_reg_report(
@@ -199,7 +200,9 @@ axd = plt.figure(layout="constrained").subplot_mosaic(
     """
 )
 g = sns.swarmplot(
-    data=res.sort("Measurement").to_pandas(),
+    data=res.filter(pl.col("Measurement") != "Granularity")
+    .sort("Measurement")
+    .to_pandas(),
     x="Measurement",
     y="r2",
     ax=axd["D"],
@@ -207,10 +210,9 @@ g = sns.swarmplot(
     hue="Measurement",
 )
 
-# g.set_ylim(0.989, 1.002)
-g.set_ylim(0, 1)
+g.set_ylim(0.925, 1.009)
+# g.set_ylim(0, 1)
 g.set_title("Linear fit")
-# g.set_xlabel("Individual features")
 g.set_ylabel("R squared")
 
 pad = 0.25

### 2025-05-25 draft.tex

@@
 
-Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025} .
+Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}.
 However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
@@
 \label{sec:org5711d86}
-A key advantage of providing these features as a stand-alone Python package is their ease of integration into diverse analytical workflows, which otherwise would require substantial adaptation to the standard CellProfiler environment.
-The recent proliferation of black-box foundation models trained solely on morphological data highlights morphology as a highly informative and predictive modality.
+A key advantage of providing these widely used features with proven usefulness as a stand-alone Python package is how easy this makes integration into diverse analytical workflows.
+To demonstrate this, we have developed a module based on cp\_measure for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
+Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E stained or fluorescently labeled.
+
+While not the typical application domain of CellProfiler, the recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality.
 However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment.
-In contrast, classical morphological features yield explicit, interpretable readouts -- for instance, the co-localization of fluorescent markers -- facilitating clear biological interpretations.
+In contrast, using cp\_measure to featurize such images will yield more interpretable features.
 
-To demonstrate this utility, we integrated our cp\_measure-based feature extraction into the widely used spatial analysis library Squidpy \citep{pallaSquidpyScalableFramework2022}.
-Being stand-alone allowed seamless incorporation into workflows powered by the robust SpatialData \citep{marconatoSpatialDataOpenUniversal2025} framework underlying Squidpy.
-Because spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screenings -- up to approximately 100,000 cells-traditional software typically cannot process these large images without cropping, which introduces boundary artefacts.
-Leveraging the modular design of cp\_measure, we parallelized feature extraction at the single-cell level, streaming batches of cells across computational cores.
-This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler software.
+However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing.
+Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, we used cp\_measure to develop a workflow tailored to such data.
+For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects.
+This allows us to parallelize the feature extraction with cp\_measure at the mask level, for example individual cells, streaming batches of cells across computational cores.
+This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler.
+Furthermore, this establishes morphological profiling within the scverse ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.
 
-To further illustrate the value of morphological features, we evaluated their impact on cell-type prediction tasks using spatial transcriptomics data.
-This application is particularly compelling, as current spatial transcriptomics technologies typically produce matched histological images that remain largely underutilized beyond visualization.
-We analysed two mouse brain datasets generated by Bruker Spatial's CosMx platform \citep{CosMxSMIMouse2025}.
-Each dataset comprises expression profiles for 960 genes and immunofluorescence images captured via five distinct fluorescent probes ('Histone', 'DNA', 'GFAP', 'G', 'rRNA').
-Morphological features were extracted from these 5-channel images for both datasets.
-Subsequently, both gene expression and morphological data were preprocessed according to best practices established by Scanpy \citep{wolfSCANPYLargescaleSinglecell2018} and Pycytominer \citep{serranoReproducibleImagebasedProfiling2025} respectively.
-We trained an XGBoost model to predict cell types on the larger dataset (48,556 cells; see Fig.
-XXX, panel XXX), comparing models using either gene expression alone or combined gene expression and morphological data.
-Model performance was assessed by predicting cell types in a smaller independent dataset (38,996 cells), using the F1-score metric stratified by cell type.
-Figure XXX (panel XXX) highlights the improved predictive accuracy obtained when morphological features are included.
-Importantly, this performance enhancement required no additional experimental effort, underscoring the benefit of employing cp\_measure beyond its traditional scope.
+To illustrate the value of this workflow, we evaluated its usefulness in identifying morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium by 10x Genomics \citep{10x-Genomics2023-el}.
+We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue.
+These masks were then featurised with the described cp\_measure-based workflow.
+We first preprocessed the features as described in
+\citep{serranoReproducibleImagebasedProfiling2025}, then integrated both samples using Harmony \citep{Korsunsky2019-Harmony}, and finally clustered them with the Leiden algorithm through Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.
+
+For each cluster, we extracted the cells within its hexagons and calculated the proportions of cell types predicted by the spatial gene-expression data.
+A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture.
+Panel C further reveals that some clusters have distinct cell-type compositions: for example, Cluster XXX is dominated by fibroblasts, whereas adipocytes appear almost exclusively in Cluster XXX.
 
@@
 \centering
-\includegraphics[width=.9\linewidth]{./figs/spatial.png}
+\includegraphics[width=.99\linewidth]{./figs/fig_4_placeholder.png}
 \caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
 \end{figure}
+
+These results demonstrate that our cp\_measure–powered workflow not only scales to the high cell counts typical of spatial datasets but also produces biologically meaningful, interpretable morphological clusters that align with - and complement - transcriptomic phenotypes.
+By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.
+
 \section{Discussion}

### 2025-05-25 src/jump/plot_cp_vs_cpm.py
fig(2): transpose plot

@@ -194,9 +194,8 @@ feats_to_show = [
 ]
 axd = plt.figure(layout="constrained").subplot_mosaic(
     """
-    AD
-    BD
-    CD
+    ABC
+    DDD
     """
 )
 g = sns.swarmplot(
@@ -206,13 +205,13 @@ g = sns.swarmplot(
     x="Measurement",
     y="r2",
     ax=axd["D"],
-    alpha=0.8,
+    alpha=0.7,
     hue="Measurement",
 )
 
-g.set_ylim(0.925, 1.009)
+g.set_ylim(0.925, 1.019)
 # g.set_ylim(0, 1)
-g.set_title("Linear fit")
+# g.set_title("Linear fit")
 g.set_ylabel("R squared")
 
 pad = 0.25
@@ -243,6 +242,6 @@ for ax_id, featname in zip("ABC", feats_to_show):
         for lh in ax.get_legend().legend_handles:
             lh.set_alpha(1)
 
-        sns.move_legend(ax, loc="lower right", bbox_to_anchor=(1.12, 0))
+        sns.move_legend(ax, loc="lower right", bbox_to_anchor=(1.2, 0), frameon=False)
 plt.savefig(figs_dir / "jump_r2_examples.svg")
 plt.savefig(figs_dir / "jump_r2_examples.png")

### 2025-05-25 src/jump/plot_cp_vs_cpm.py
fig(2): improve colour palettes

@@ -206,6 +206,7 @@ g = sns.swarmplot(
     y="r2",
     ax=axd["D"],
     alpha=0.7,
+    palette=sns.color_palette("husl", 8),
     hue="Measurement",
 )
 
@@ -225,8 +226,7 @@ for ax_id, featname in zip("ABC", feats_to_show):
         y="cp_measure",
         hue="object",
         ax=ax,
-        alpha=0.1,
-        palette=sns.color_palette("husl", 8),
+        alpha=0.05,
         legend=None if ax_id != "C" else True,
     )
     # ax.text(0, 0, ax_id, fontsize=9)

### 2025-05-25 draft.tex

@@
 \caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
-\textbf{Left panel.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
-\textbf{Right panel.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
+\textbf{A,B,C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
+\textbf{D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
 \end{figure}
@@
 
-Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the Shapley values of a classifier trained on our features.
+Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features.
 The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.

### 2025-05-25 src/jump/plot_cp_vs_cpm.py
figs(2): increase font size

@@ -7,6 +7,7 @@ Plot the comparison.
 from pathlib import Path
 
 import duckdb
+import matplotlib
 import matplotlib.pyplot as plt
 import polars as pl
 import polars_ds as pds
@@ -20,8 +21,10 @@ from util_names import get_cpm_to_measurement_mapper
 from util_plot import generate_label
 
 figs_dir = Path("..") / ".." / "figs"
+profiles_dir = Path("/") / "datastore" / "alan" / "cp_measure" / "profiles_via_masks"
 
-cpmeasure_parquet = "/datastore/alan/cp_measure/profiles_via_masks/first_set.parquet"
+cpmeasure_parquet = profiles_dir / "first_set.parquet"
+correlation_parquet = profiles_dir / "correlation.parquet"
 
 
 # Turns out it is easier to trim down the CellProfiler measurements
@@ -187,6 +190,11 @@ for feat_name, x in partitioned.items():
 res = pl.concat(dfs)
 # %%
 plt.close()
+# font = {"family": "serif", "size": 14}
+font = {"family": "sans-serif", "size": 13}
+
+matplotlib.rc("font", **font)
+
 feats_to_show = [
     "Area",
     "Intensity_MedianIntensity",
@@ -210,10 +218,10 @@ g = sns.swarmplot(
     hue="Measurement",
 )
 
-g.set_ylim(0.925, 1.019)
+g.set_ylim(0.925, 1.021)
 # g.set_ylim(0, 1)
 # g.set_title("Linear fit")
-g.set_ylabel("R squared")
+g.set_ylabel(r"$R^2$")
 
 pad = 0.25
 axd["D"].add_artist(generate_label("D", pad=pad))
@@ -233,6 +241,11 @@ for ax_id, featname in zip("ABC", feats_to_show):
     h.set_yticklabels(
         h.get_yticklabels(), rotation=30, ha="right", rotation_mode="anchor"
     )
+    h.set_xticklabels(
+        h.get_xticklabels(), rotation=30, ha="right", rotation_mode="anchor"
+    )
+    if ax_id != "A":
+        ax.set_ylabel("")
     featname = featname.removeprefix("RadialDistribution_")
     featname = featname.removeprefix("Intensity_")
     ax.set_title(featname.replace("_", " "))
@@ -242,6 +255,13 @@ for ax_id, featname in zip("ABC", feats_to_show):
         for lh in ax.get_legend().legend_handles:
             lh.set_alpha(1)
 
[…]

### 2025-05-25 draft.tex
Updates from Overleaf

@@
 It can be used by experimental biologists with scant programming experience, for whom it has been a boon.
-By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools.
+By contrast, it may not suit the needs of computer-satvvy scientists building high throughput pipelines comprised of multiple tools.
 CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
@@
 \label{sec:org09b0cd2}
-
-\begin{figure}[htbp]
-\centering
-\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
-\textbf{A,B,C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
-\textbf{D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
-\end{figure}
-
 We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
@@
 These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
+
+\begin{figure}[htbp]
+\centering
+\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
+\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
+\textbf{A,B,C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
+\textbf{D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
+\end{figure}
+
 \subsection{Results}
@@
 
+Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features.
+The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
+
 \begin{figure}[htbp]
@@
 \includesvg[width=.9\linewidth]{./figs/example_shap.svg}
-\caption{\label{fig:astrocytes}\textbf{Top panel.} Example pair of astrocyte nuclei image and masks.
+\caption{\label{fig:astrocytes}\textbf{A,B.} Example pair of astrocyte nuclei image and masks.
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
@@
 
-Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features.
-The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
 \subsubsection{Spatial transcriptomics}
 \label{sec:org5711d86}
-A key advantage of providing these widely used features with proven usefulness as a stand-alone Python package is how easy this makes integration into diverse analytical workflows.
-To demonstrate this, we have developed a module based on cp\_measure for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
-Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E stained or fluorescently labeled.
+To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
+Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labeled.
 
-While not the typical application domain of CellProfiler, the recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality.
-However, the feature vectors produced by these models are typically not interpretable, preventing direct biological assessment.
+The recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality.
+However, the features these models are typically not interpretable, preventing direct biological assessment.
 In contrast, using cp\_measure to featurize such images will yield more interpretable features.
@@
 However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing.
-Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, we used cp\_measure to develop a workflow tailored to such data.
+Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, the standalone nature of cp\_measure allowed us to tailor a workflow to such data.
 For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects.
@@
 We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue.
-These masks were then featurised with the described cp\_measure-based workflow.
+These masks were then featurized with the described cp\_measure-based workflow.
 We first preprocessed the features as described in
@@
 A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture.
-Panel C further reveals that some clusters have distinct cell-type compositions: for example, Cluster XXX is dominated by fibroblasts, whereas adipocytes appear almost exclusively in Cluster XXX.
+Panel C further reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in cluster 5-7.
 
@@
 \centering
-\includegraphics[width=.99\linewidth]{./figs/fig_4_placeholder.png}
-\caption{\label{fig:spatial_omics}{[}PLACEHOLDER] Spatial omics analysis.}
+\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
+\caption{\label{fig:spatial_omics}{}Morphology-driven tumor subregion discovery.
+\textbf{Panel A}: Spatial map of cell types inferred from transcriptomics.
+\textbf{Panel B}: Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings..
+\textbf{Panel C}: Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (color legend as shown).
+Of note is that the cell segmentation in the top-left quarter of the image seems to have performed poorly, since no cells were annotated, whereas the morphological clusters cover these regions.}
 \end{figure}
 
-These results demonstrate that our cp\_measure–powered workflow not only scales to the high cell counts typical of spatial datasets but also produces biologically meaningful, interpretable morphological clusters that align with - and complement - transcriptomic phenotypes.
+These results demonstrate how cp\_measure can now easily be used outside of its traditional domain.
+Using it to featurize extremely large images, we show that it produces biologically meaningful morphological clusters that align with - and complement - transcriptomic phenotypes.
 By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.
@@
 \label{sec:org5cdbb12}
-We propose in future work to make it possible for CellProfiler to import cp\_measure, which offers several benefits.
+We propose to make cp\_measure an imported dependency for CellProfiler, offering several benefits.
 It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
@@
 All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}.
-The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest}.
-\subsubsection{Data and software}
+The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest} and \url{https://zenodo.org/records/15505477/latest}.
+\subsubsection{Acknowledgements}
 \label{sec:acknowledgements}

### 2025-05-25 src/jump/plot_cp_vs_cpm.py
fig2) : increase fontsize further

@@ -191,13 +191,13 @@ res = pl.concat(dfs)
 # %%
 plt.close()
 # font = {"family": "serif", "size": 14}
-font = {"family": "sans-serif", "size": 13}
+font = {"family": "sans-serif", "size": 14}
 
 matplotlib.rc("font", **font)
 
 feats_to_show = [
-    "Area",
     "Intensity_MedianIntensity",
+    "Area",
     "RadialDistribution_RadialCV_1of4",
 ]
 axd = plt.figure(layout="constrained").subplot_mosaic(

### 2025-05-25 src/astro3d/plot_figures.py
fig(3): increase font size

@@ -2,6 +2,7 @@
 from itertools import chain
 from pathlib import Path
 
+import matplotlib
 import matplotlib.pyplot as plt
 import numpy as np
 import polars as pl
@@ -208,7 +209,11 @@ shap_values = explanation.values
 # make sure the SHAP values add up to marginal predictions
 # np.abs(shap_values.sum(axis=1) + explanation.base_values - pred).max)
 plt.close()
-# %% example figure
+# %% Figure 2: Examples and SHAP values
+font = {"family": "sans-serif", "size": 14}
+
+matplotlib.rc("font", **font)
+
 axd = plt.figure(layout="constrained").subplot_mosaic(
     """
     AB
@@ -264,5 +269,7 @@ for ax_id in "AB":
 axd["C"].add_artist(
     generate_label("C", bbox_to_anchor=(-0.98, 1.04), bbox_transform=axd["C"].transAxes)
 )
+axd["C"].set_yticklabels(axd["C"].get_yticklabels(), fontdict=font)
+axd["C"].set_xlabel(axd["C"].get_xlabel(), fontdict=font)
 plt.savefig(figs_path / "example_shap.svg")
 plt.close()

### 2025-05-25 src/jump/plot_cp_vs_cpm.py

@@ -190,7 +190,6 @@ for feat_name, x in partitioned.items():
 res = pl.concat(dfs)
 # %%
 plt.close()
-# font = {"family": "serif", "size": 14}
 font = {"family": "sans-serif", "size": 14}
 
 matplotlib.rc("font", **font)

### 2025-05-25 src/astro3d/plot_figures.py
fig(3): further adjust text size and details

@@ -240,7 +240,7 @@ axd["C"].set_yticklabels(
 )
 axd["A"].imshow(img)
 axd["A"].axis("off")
-axd["A"].set_title("Z-projected image")
+axd["A"].set_title("Image")
 axd["B"].imshow(labels)
 axd["B"].axis("off")
 axd["B"].set_title("Labels")
@@ -269,7 +269,9 @@ for ax_id in "AB":
 axd["C"].add_artist(
     generate_label("C", bbox_to_anchor=(-0.98, 1.04), bbox_transform=axd["C"].transAxes)
 )
+font = {"family": "sans-serif", "size": 15}
 axd["C"].set_yticklabels(axd["C"].get_yticklabels(), fontdict=font)
+axd["C"].tick_params(axis="y", which="major", pad=-16)
 axd["C"].set_xlabel(axd["C"].get_xlabel(), fontdict=font)
 plt.savefig(figs_path / "example_shap.svg")
 plt.close()

### 2025-05-26 draft.tex
Updates from Overleaf

@@
 \label{sec:org9dc3dfa}
-An alternative to extracting a defined set of engineered features is to instead use a pretrained deep-learning model to extract features.
+An alternative to extracting a defined set of engineered features is to instead use a pre-trained deep-learning model to extract features.
 Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
@@
 \includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogs.
-\textbf{A,B,C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
-\textbf{D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
+\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogues.
+\textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
+\textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
 \end{figure}
@@
 \includesvg[width=.9\linewidth]{./figs/example_shap.svg}
-\caption{\label{fig:astrocytes}\textbf{A,B.} Example pair of astrocyte nuclei image and masks.
+\caption{\label{fig:astrocytes}\textbf{Panels A and B.} Example pair of astrocyte nuclei image and masks.
 The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
-\textbf{Bottom panel.} SHAP values of the most important features for classifying the day of differentiation (out of three days).
+\textbf{Panel C.} SHAP values of the most important features for classifying the day of differentiation (out of three days).
 Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier.
@@
 To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
-Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labeled.
+Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.
 
-The recent proliferation of black-box foundation models trained solely on morphological data \citep{Chen2024-ix, Xu2024-mn} highlights morphology as a highly informative and predictive modality.
-However, the features these models are typically not interpretable, preventing direct biological assessment.
-In contrast, using cp\_measure to featurize such images will yield more interpretable features.
+Using CellProfiler on spatial datasets is not convenient due to its enforcement of a nested structure on input data and, since it requires loading entire images into RAM, struggles.
+Since images from spatial datasets contain thousands of times more cells than conventional microscopy data, usually reaching 500.000 cells per image, the stand-alone nature of cp\_measure allowed us to tailor a workflow to match this data throughput.
+We used Squidpy to leverage the SpatialData data format \citep{Marconato2024-SpatialData}, enabling us to parallelize the feature extraction with cp\_measure to streaming batches of individual cells across multiple processing cores.
+This approach enables efficient computation on large-scale datasets, which would be intractable using CellProfiler.
+Furthermore, this brings morphological profiling into the single-cell -- scverse -- ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.
 
-However, directly applying CellProfiler to spatial datasets is not convenient since it imposes a nested structure on input data and also cannot handle extremely large images since the full image is loaded into RAM for processing.
-Since spatial datasets often comprise significantly more cells per field-of-view (FOV) than conventional microscopy screening data, easily 500.000 cells per image, the standalone nature of cp\_measure allowed us to tailor a workflow to such data.
-For this, Squidpy leverages the powerful data format SpatialData \citep{Marconato2024-SpatialData} in which images are stored as zarr objects.
-This allows us to parallelize the feature extraction with cp\_measure at the mask level, for example individual cells, streaming batches of cells across computational cores.
-This approach enables efficient computation even on large-scale datasets, a feat not achievable with standard CellProfiler.
-Furthermore, this establishes morphological profiling within the scverse ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.
+To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}.
+We tessellated hexagonal masks across both images, of which XXXXXX were found to overlay the tissue.
+These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following the standard method
+\citep{serranoReproducibleImagebasedProfiling2025}, then integrated our two samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.
 
-To illustrate the value of this workflow, we evaluated its usefulness in identifying morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium by 10x Genomics \citep{10x-Genomics2023-el}.
-We generated a tessellating pattern of hexagonal masks across both images, of which XXXXXX were found to overlay the tissue.
-These masks were then featurized with the described cp\_measure-based workflow.
-We first preprocessed the features as described in
-\citep{serranoReproducibleImagebasedProfiling2025}, then integrated both samples using Harmony \citep{Korsunsky2019-Harmony}, and finally clustered them with the Leiden algorithm through Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.
-
-For each cluster, we extracted the cells within its hexagons and calculated the proportions of cell types predicted by the spatial gene-expression data.
+For each cluster, we extracted the cells within its hexagons and calculated the cell type proportions predicted by gene-expression.
 A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture.
-Panel C further reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in cluster 5-7.
+Panel C reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.
 
@@
 \includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
-\caption{\label{fig:spatial_omics}{}Morphology-driven tumor subregion discovery.
-\textbf{Panel A}: Spatial map of cell types inferred from transcriptomics.
-\textbf{Panel B}: Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings..
-\textbf{Panel C}: Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (color legend as shown).
-Of note is that the cell segmentation in the top-left quarter of the image seems to have performed poorly, since no cells were annotated, whereas the morphological clusters cover these regions.}
+\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions.
+\textbf{Panel A.} Spatial map of cell types inferred from transcriptomics.
+\textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings.
+\textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown).
+The top left quarter of the image shows that the morphological approach worked to predict regions while the expression annotations did not.}
 \end{figure}
 
-These results demonstrate how cp\_measure can now easily be used outside of its traditional domain.
-Using it to featurize extremely large images, we show that it produces biologically meaningful morphological clusters that align with - and complement - transcriptomic phenotypes.
-By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly integrate morphology into their spatial omics analyses, unlocking new insights into tissue organization and cell–cell interactions.
+These results demonstrate how cp\_measure can can be used outside traditional microscopy.
+When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes.
+By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization and cell–cell interactions.
 
@@
 \label{sec:orgbda0ae2}
-The code for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
+The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
 All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}.
-The datasets we produced for this work are available on Zenodo, and the latest version can be found on \url{https://zenodo.org/records/15390631/latest} and \url{https://zenodo.org/records/15505477/latest}.
+The datasets we produced for this work are available on Zenodo.
+The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}.
+The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
+
 \subsubsection{Acknowledgements}

### 2025-05-26 draft.tex

@@
 \centering
-\includegraphics[width=.99\linewidth]{./figs/fig_1_tmp.png}
+\includesvg[width=.99\linewidth]{./figs/cpmeasure_overview.svg}
 \caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
@@
 The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
-All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025\_cpmeasure/}.
-The datasets we produced for this work are available on Zenodo.
+All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_cpmeasure-3FBF}.
+%The datasets we produced for this work are available on Zenodo.
 The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}.
 The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
-
+% 
 \subsubsection{Acknowledgements}

### 2025-05-26 draft.tex

@@
 Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. 
-A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles where patterns of changes show biological properties.
+A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
 While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows.
@@
 
-Image-based profiling--also termed morphological profiling--is a technique of measuring an array of shape and intensity features for a population of biological objects, such as cells.
+Image-based profiling--also termed morphological profiling--is a technique of measuring an extensive suite of morphological features for a population of biological objects, such as cells. 
 These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.
+
 One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.
@@
 It can be used by experimental biologists with scant programming experience, for whom it has been a boon.
-By contrast, it may not suit the needs of computer-satvvy scientists building high throughput pipelines comprised of multiple tools.
+By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools.
 CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
@@
 Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}.
-However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
+%However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
 Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image.
@@
 
+ However, these approaches present trade-offs for different use cases. 
+ Their independent implementation means they generate different feature sets than CellProfiler, which can be advantageous for novel
+ applications but complicates direct comparison with established datasets like those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}. 
+Additionally, while all tools necessarily impose input requirements, some enforce rigid organizational constraints—demanding specific directory structures and naming conventions—that require extensive data reorganization before analysis, beyond the standard data type and format specifications.
+Some alternatives also target specific deployment scenarios, with ScaleFex designed primarily for cloud environments, limiting utility for researchers preferring local computation.
+
 \subsection{Engineered features in the era of deep learning}
 \label{sec:org9dc3dfa}
-An alternative to extracting a defined set of engineered features is to instead use a pre-trained deep-learning model to extract features.
+%An alternative to extracting a defined set of engineered features is to instead use a pre-trained deep-learning model to extract features.
 Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
@@
 
-In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract interpretable features while readily integrating with computational workflows to achieve reproducible data analyses, whether locally or in the cloud.
+%In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract interpretable features while readily integrating with computational workflows to achieve reproducible data analyses, whether locally or in the cloud.
 We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
+With all this in mind, we developed cp\_measure.
+
+While computer vision has largely transitioned to deep learning-based feature extraction, biological imaging presents unique challenges that maintain the relevance of engineered features.
+These images differ substantially from natural images in their acquisition methods, visual properties, and analytical objectives, often limiting the effectiveness of standard pre-trained models, even those trained on biological data.
+
+The application of deep learning to feature extraction for biological image analysis shows mixed results.
+When appropriately trained on domain-specific data, deep learning networks can outperform engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not universally \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
+Effective deep learning feature extractors for biological applications typically require substantial domain-specific datasets and careful model architecture choices, rather than simple transfer from general computer vision models.
+
+Several practical considerations favor engineered features in biological imaging. 
+First, interpretability remains crucial for biological discovery—researchers need to understand which morphological characteristics drive observed phenotypes. 
+While deep learning features can be powerful predictors, their interpretation requires sophisticated techniques that may not provide the mechanistic insights biologists seek \citep{liChallengesOpportunitiesBioimage2023}.
+Engineered features have clear mathematical definitions that directly translate to biological concepts, such as protein colocalization or nuclear morphology changes \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
+
+Second, many biological imaging laboratories lack the GPU resources, large annotated datasets, or machine learning expertise required for custom model development.
+While foundational models are emerging, they often struggle with the diversity of imaging modalities and experimental conditions \cite{azadFoundationalModelsMedical2023}.
+
+Third, engineered features offer consistency and reproducibility advantages valuable for comparative studies and multi-site collaborations, producing identical results across different computing environments.
+
+Rather than viewing these approaches as competing alternatives, the field increasingly recognizes their complementary roles. 
+This motivates the need for robust, accessible tools that efficiently extract interpretable features while integrating seamlessly with modern computational workflows.
 With all this in mind, we developed cp\_measure.
@@
 Next, we applied cp\_measure on these masks and the original images.
-Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
-
-We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}).
+To validate cp\_measure, we compared its output directly with CellProfiler's on identical images and masks. 
+For each feature, we computed the $R^2$ value between the two tools' measurements. 
+Over 90\% of features showed $R^2$ > 0.9 (Figure \ref{fig:cp_vs_cpmeasure}), indicating near-perfect agreement. 
+The few outliers with lower correlations reveal edge cases handled differently by each tool; these will be addressed through comprehensive unit testing.
+%Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
+%We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}).
 We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation.
@@
 
-To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow.
-We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalinin3DCellNuclear2018}.
+%To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow.
+We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalininValproicAcidinducedChanges2021}.
 We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
@@
 Finally, we calculated the SHAP values \citep{sundararajanManyShapleyValues2020} to get a better understanding of the effects of the drugs on the cells.
+% Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features.
+The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
 
-Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features.
-The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
+We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
+From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
+We then traind a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
+SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
 
@@
 
-Using CellProfiler on spatial datasets is not convenient due to its enforcement of a nested structure on input data and, since it requires loading entire images into RAM, struggles.
+% Using CellProfiler on spatial datasets is not convenient due to its enforcement of a nested structure on input data and, since it requires loading entire images into RAM, struggles.
 Since images from spatial datasets contain thousands of times more cells than conventional microscopy data, usually reaching 500.000 cells per image, the stand-alone nature of cp\_measure allowed us to tailor a workflow to match this data throughput.
@@
 
+Spatial transcriptomics poses unique computational challenges: images routinely contain 500,000+ cells, far exceeding CellProfiler's memory constraints and rigid data structure requirements. 
+cp\_measure's modular design enabled us to build a custom workflow integrated with Squidpy \citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches.
+This architecture handles massive spatial datasets efficiently while bringing morphological profiling directly into the scverse ecosystem \citep{Virshup2023-scverse}, making these features immediately accessible to the single-cell community.
+
 To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}.
-We tessellated hexagonal masks across both images, of which XXXXXX were found to overlay the tissue.
-These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following the standard method
+We tessellated hexagonal masks across both images, of which 18469 were found to overlay the tissue.
+These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following standard workflows
 \citep{serranoReproducibleImagebasedProfiling2025}, then integrated our two samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.
@@
 
-Already, Cellprofiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
+% Already, CellProfiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
 By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
+While developed for biology, engineered morphological features have proven valuable far beyond their original domain—from analyzing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING.
+cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
+
 \section{Future work}

### 2025-05-26 draft.tex

@@
 From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-We then traind a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
+We then trained a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
 SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
@@
 Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value.
-Shape features refer to the nucleus and intensity features refer to the DNA staining.
+Shape features, such as the minor axis length, refer to the nucleus, while intensity features refer to the DNA staining.
 The test data accuracy is shown in bold.}
@@
 \label{sec:org5711d86}
-To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer with it at its core for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
+To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
 Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.
@@
 \textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown).
-The top left quarter of the image shows that the morphological approach worked to predict regions while the expression annotations did not.}
+The top left quarter of Panel A shows that the morphological approach worked to predict regions while the expression annotations did not.}
 \end{figure}
@@
 
-% Already, CellProfiler features have been used in multiple experiments, including in non-biological contexts such as environmental monitoring \citep{ideharaExploringNileRed2025}.
-By association, we foresee cp\_measure being useful for a broad scientific community beyond image-based profiling.
-While developed for biology, engineered morphological features have proven valuable far beyond their original domain—from analyzing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING.
+While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING and agricultural automation \citep{xuDesignNonDestructiveSeed2024}.
 cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
@@
 \bibliography{bibliography}
+
+\onecolumn
 \section{Appendix}
+
 \label{sec:orgdd18dd8}
@@
 %
+
+\subsubsection{Details on the spatial use case}
+\label{sec:spatial_extra}
+
 \subsubsection{Acknowledgements}

### 2025-05-26 draft.tex
sent and anonymized version

@@
 \twocolumn[
-\icmltitle{cp\_measure: API-first feature extraction for image-based profiling workflows}
+\icmltitle{LIB\_NAME: API-first feature extraction for image-based profiling workflows}
 
@@
 A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
-While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering ML workflows.
-Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
-We demonstrate that cp\_measure features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem.
-Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
+While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering machine learning workflows.
+Here we introduce LIB\_NAME, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
+We demonstrate that LIB\_NAME features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem.
+Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how LIB\_NAME enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
 \end{abstract}
@@
 Through microscopy, modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components.
-Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity distributions.
+Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity.
 
@@
 These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.
-
-One of its biggest applications is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{sealDecadeSystematicReview2024}.
+One of its biggest applications of morphological profiling is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{chandrasekaranImagebasedProfilingDrug2021}.
 
@@
 \includesvg[width=.99\linewidth]{./figs/cpmeasure_overview.svg}
-\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
+\caption{\label{fig:overview}LIB\_NAME generates features from images by using information in every region of interest ("object").
 It can featurize the pairwise combination of all the available channels (colours) and objects.
@@
 \label{sec:org8f5b33d}
-The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}.
-It can be used by experimental biologists with scant programming experience, for whom it has been a boon.
-By contrast, it may not suit the needs of computer-savvy scientists building high throughput pipelines comprised of multiple tools.
-CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles, however, certain aspects become a hindrance.
+The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}, providing experimental biologists with limited programming expertise an accessible yet powerful toolset.
+By contrast, it may not suit the needs of computational biologists building high throughput pipelines comprised of multiple tools.
+CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles or spatial transcriptomics data, however, certain aspects become a hindrance.
 For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
@@
 
-Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}.
-%However, these come with constraints: Given that their implementation is completely independent from CellProfiler, they do not produce the same feature set, precluding matching to existing datasets, such as those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024,fayRxRx3PhenomicsMap2023}.
-Moreover, a common problem is that these tools enforce a structure for the input data, such as determined hierarchical file structures, which require significant wrangling with the data before processing a single image.
-Some alternatives, such as ScaleFex, are designed for cloud usage, neglecting a significant number of cases where local compute suffices.
-
- However, these approaches present trade-offs for different use cases. 
+Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}, however, these approaches present trade-offs for different use cases.
 Their independent implementation means they generate different feature sets than CellProfiler, which can be advantageous for novel
@@
 \label{sec:org9dc3dfa}
-%An alternative to extracting a defined set of engineered features is to instead use a pre-trained deep-learning model to extract features.
-Results are mixed, with suitably-trained deep learning networks usually outperforming engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not always \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
-Training a deep learning model to be a good feature extractor takes significant hands-on and compute time, and often requires appropriate training images.
-Compared to engineered features, deep learning approaches may pose a higher barrier of entry, as most models require either annotated data or a large and high-quality dataset, in addition to having access to a GPU.
-Foundational models aim to lower this barrier, but can fail to generalize well for out-of-distribution data (data which differs sufficiently from the training datasets) \cite{azadFoundationalModelsMedical2023}.
-Generally deep learning-based feature extraction takes less wall-clock time than CellProfiler does (provided GPU compute is available); however, the features are not obviously interpretable and require specialized techniques for biologists to attempt to understand patterns of changes in image-based profiles, which is not always successful \citep{liChallengesOpportunitiesBioimage2023}.
-Unlike learned features, whose interpretation requires complex methods like counterfactual techniques or attention maps, most engineered features have a mathematical definition that is faster to calculate and easier to translate into biological concepts, such as the colocalisation of two proteins \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
-
-%In summary, while deep learning-based features have some advantages in some contexts, and while tools to extract engineered features exist, there remains a need for a modular tool that can extract interpretable features while readily integrating with computational workflows to achieve reproducible data analyses, whether locally or in the cloud.
-We thus believe that there is the need for a readily-integrated feature extraction tool aimed at generalist and high throughput problems not being solved by the current software ecosystem.
-With all this in mind, we developed cp\_measure.
 
@@
 
-Several practical considerations favor engineered features in biological imaging. 
+Several practical considerations favour engineered features in biological imaging. 
 First, interpretability remains crucial for biological discovery—researchers need to understand which morphological characteristics drive observed phenotypes. 
 While deep learning features can be powerful predictors, their interpretation requires sophisticated techniques that may not provide the mechanistic insights biologists seek \citep{liChallengesOpportunitiesBioimage2023}.
-Engineered features have clear mathematical definitions that directly translate to biological concepts, such as protein colocalization or nuclear morphology changes \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
+Engineered features have clear mathematical definitions that directly translate to biological concepts, such as protein co-localisation or nuclear morphology changes \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
 
 Second, many biological imaging laboratories lack the GPU resources, large annotated datasets, or machine learning expertise required for custom model development.
-While foundational models are emerging, they often struggle with the diversity of imaging modalities and experimental conditions \cite{azadFoundationalModelsMedical2023}.
+While foundational models are emerging, visual foundational models can be challenged by the diversity of imaging modalities and experimental conditions \cite{azadFoundationalModelsMedical2023}.
 
@@
 This motivates the need for robust, accessible tools that efficiently extract interpretable features while integrating seamlessly with modern computational workflows.
-With all this in mind, we developed cp\_measure.
+With all this in mind, we developed LIB\_NAME.
 
@@
 \label{sec:org61842b5}
-The library cp\_measure branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
+The library LIB\_NAME branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
 Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels).
+A visual representation of this can be found in Figure~\ref{fig:overview}.
 
@@
 
-%First, we demonstrated the validity and capabilities of cp\_measure in cases that would normally require image analysis knowledge and a considerable amount of time.
-First, we validate cp\_measure features versus CellProfiler results with a subset of the JUMP dataset \citep{chandrasekaranJUMPCellPainting2023}.
-Then we showcase cases in which cp\_measure is a more practical choice to process microscopy data: First 3D images of astrocytes and then spatial transcriptomics.
-These use-cases demonstrate its widespread suitability for different types of problems.
 \subsection{Reproducing CellProfiler measurements}
 \label{sec:org09b0cd2}
-We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
+We first tested whether LIB\_NAME features match the original CellProfiler features, for a representative set of 300 images.
 These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature.
-We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
-Next, we applied cp\_measure on these masks and the original images.
-To validate cp\_measure, we compared its output directly with CellProfiler's on identical images and masks. 
+We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to LIB\_NAME for feature extraction.
+Next, we applied LIB\_NAME on these masks and the original images.
+To validate LIB\_NAME, we compared its output directly with CellProfiler's on identical images and masks. 
 For each feature, we computed the $R^2$ value between the two tools' measurements. 
-Over 90\% of features showed $R^2$ > 0.9 (Figure \ref{fig:cp_vs_cpmeasure}), indicating near-perfect agreement. 
+Over 95\% of features showed $R^2$ > 0.9 (Figure \ref{fig:cp_vs_cpmeasure}), indicating near-perfect agreement. 
 The few outliers with lower correlations reveal edge cases handled differently by each tool; these will be addressed through comprehensive unit testing.
-%Finally, we mapped the features from cp\_measure to CellProfiler and calculated a linear fit for each matched feature, plus the \(R^2\) value, indicating how well the correlation fits a linear slope.
-%We find that most features had an $R^2$ correlation larger than 90\% (Figure \ref{fig:cp_vs_cpmeasure}).
-We examined several individual features, typically finding straight lines demonstrating the recapitulation of features from our implementation.
-For a small set of features, some data points fall off the diagonal, which hint that some edge cases are treated differently by either tool; we plan to investigate these further as part of a larger project to implement unit testing in CellProfiler.
-These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
+These results indicate that LIB\_NAME is not likely to produce massively different profiles than those calculated by CellProfiler.
 
@@
 \includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogues.
-\textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
-\textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
+\caption{\label{fig:cp_vs_cpmeasure}LIB\_NAME features match their CellProfiler analogues.
+\textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to LIB\_NAME's (y-axis), generated using matching pairs of masks and images.
+\textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing LIB\_NAME to CellProfiler.}
 \end{figure}
@@
 
-%To demonstrate its ease of use and the value of interpretable features, we next tested cp\_measure in a classification workflow.
-We processed 433 three-dimensional (3D) images of astrocyte nuclei containing 831 nuclei \citep{kalininValproicAcidinducedChanges2021}.
-We preprocessed the profiles following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-Then, we trained a Gradient Boosting classifier to identify the day in which the image of any given cell was acquired, which corresponds to the extent of maturation of the cells.
-We achieved 87\% classification accuracy in the testing dataset.
-Using the image-based profiles, we identified which features distinguish cells based on the time at which the image was taken.
-Finally, we calculated the SHAP values \citep{sundararajanManyShapleyValues2020} to get a better understanding of the effects of the drugs on the cells.
-% Figure \ref{fig:astrocytes} shows an example image and its corresponding object masks alongside the SHAP values of a classifier trained on our features.
-The minor axis length of the nuclei increases over the course of the experiment, implying that nuclei become wider - an expected phenotypic change that cp\_measure was able to uncover.
+We demonstrate LIB\_NAME's utility through a biologically relevant task: tracking astrocyte maturation over time. 
+From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, LIB\_NAME extracted morphological features that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
+We then trained a gradient-boosting classifier to predict the day of cell differentiation which it was able to do with 86\% accuracy.
 
-We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
-From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-We then trained a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
-SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
+SHAP analysis \citep{sundararajanManyShapleyValues2020} (Figure\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
 
@@
 \label{sec:org5711d86}
-To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy \citep{pallaSquidpyScalableFramework2022}.
+To demonstrate the broad applicability of LIB\_NAME, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy~\citep{pallaSquidpyScalableFramework2022}.
 Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.
 
-% Using CellProfiler on spatial datasets is not convenient due to its enforcement of a nested structure on input data and, since it requires loading entire images into RAM, struggles.
-Since images from spatial datasets contain thousands of times more cells than conventional microscopy data, usually reaching 500.000 cells per image, the stand-alone nature of cp\_measure allowed us to tailor a workflow to match this data throughput.
-We used Squidpy to leverage the SpatialData data format \citep{Marconato2024-SpatialData}, enabling us to parallelize the feature extraction with cp\_measure to streaming batches of individual cells across multiple processing cores.
-This approach enables efficient computation on large-scale datasets, which would be intractable using CellProfiler.
-Furthermore, this brings morphological profiling into the single-cell -- scverse -- ecosystem \citep{Virshup2023-scverse}, exposing it to the wider single-cell community.
-
 Spatial transcriptomics poses unique computational challenges: images routinely contain 500,000+ cells, far exceeding CellProfiler's memory constraints and rigid data structure requirements. 
-cp\_measure's modular design enabled us to build a custom workflow integrated with Squidpy \citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches.
+LIB\_NAME's modular design enabled us to build a custom workflow integrated with Squidpy~\citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches.
 This architecture handles massive spatial datasets efficiently while bringing morphological profiling directly into the scverse ecosystem \citep{Virshup2023-scverse}, making these features immediately accessible to the single-cell community.
 
+\\
+
 To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}.
-We tessellated hexagonal masks across both images, of which 18469 were found to overlay the tissue.
-These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following standard workflows
-\citep{serranoReproducibleImagebasedProfiling2025}, then integrated our two samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018}.
+We generated a pattern of tessellating hexagonal masks across both images, of which 18469 were found to overlay the tissue.
+These masks were then featurized with a LIB\_NAME-based workflow: We first preprocessed the features following standard workflows
+\citep{serranoReproducibleImagebasedProfiling2025}, then removed batch effects between the samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018, Traag2019-fr}.
 
-For each cluster, we extracted the cells within its hexagons and calculated the cell type proportions predicted by gene-expression.
-A visual comparison of panels A and B in Figure \ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture.
-Panel C reveals that some clusters have distinct cell-type compositions: for example, cluster 2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.
+For each cluster, we extracted the cells within its hexagons and calculated the respective cell type proportions.
+A visual comparison of panels A and B in Figure~\ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture.
+Panel C reveals that some clusters have distinct cell-type compositions: For example, cluster~2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.
+Furthermore, comparing clusters~0 and 1 suggests intra-tumour differences, which could be correlated to immune-hotness (see Figure \ref{fig:spatial_immunehot}).
 
@@
 \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics.
-\textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings.
-\textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown).
-The top left quarter of Panel A shows that the morphological approach worked to predict regions while the expression annotations did not.}
+\textbf{Panel B.} Morphological subregions obtained by Leiden clustering the image-derived feature embeddings.
+\textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type.
+Clusters 5-7 were merged due to similarity, cluster~10 removed due to low cell count (see Figure \ref{fig:spatial_c10}B).}
 \end{figure}
 
-These results demonstrate how cp\_measure can can be used outside traditional microscopy.
+These results demonstrate how LIB\_NAME can can be used outside traditional microscopy.
 When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes.
-By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization and cell–cell interactions.
+By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization.
 
@@
 While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop.
-In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
+In this work we introduced our new library LIB\_NAME, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
 It enables simpler automated profiling of microscopy data in short scripts and complex pipelines.
@@
 
-The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology.
-A potential workflow would be to use deep learning features to cluster and then use cp\_measure features from these clusters.
+The biologically interpretable features provided by LIB\_NAME complement deep learning features and offer a better mechanistic understanding of the underlying biology.
+A potential workflow would be to use deep learning features to cluster and then use LIB\_NAME features from these clusters.
 As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.
 
-While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing CITE SOMETHING and agricultural automation \citep{xuDesignNonDestructiveSeed2024}.
-cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
+While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing \citep{ilhanEffectsProcessParameters2021} and agriculture automation \citep{xuDesignNonDestructiveSeed2024}.
+LIB\_NAME's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
 
@@
 \label{sec:org5cdbb12}
-We propose to make cp\_measure an imported dependency for CellProfiler, offering several benefits.
+We propose to make LIB\_NAME an imported dependency for CellProfiler, offering several benefits.
 It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
-It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.
+It would also mean that any changes made in LIB\_NAME propagate to CellProfiler, benefiting the community for which it was originally developed.
 
@@
 Once tests are in place, we could add support for just-in-time compiling and GPUs.
-We envision that cp\_measure could also be the place to develop and distribute new measurements from the community.
+We envision that LIB\_NAME could also be the place to develop and distribute new measurements from the community.
 
@@
 
+\appendix
+\setcounter{figure}{0}
+% redefine \thefigure to prefix “A” and then use arabic
+\renewcommand{\thefigure}{A\arabic{figure}}
+
+% fix appendix section numbering
+\setcounter{section}{5}
+\renewcommand{\thesection}{\arabic{section}}
+\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
+\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
+
 \label{sec:orgdd18dd8}
@@
 \label{sec:orgbda0ae2}
-The codebase for cp\_measure is available on \url{https://anonymous.4open.science/r/cp\_measure-B0DA}.
-All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_cpmeasure-3FBF}.
+The codebase for LIB\_NAME is available on \url{https://anonymous.4open.science/anonymize/feature_extraction_image_profiling}.
+All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_paper_feature_extraction_image_profiling/readme.org}.
 %The datasets we produced for this work are available on Zenodo.
@@
 
+The following figures show the processing steps for the spatial data used to generate Figure~\ref{fig:spatial_omics} in more detail.
+Raw data from two Xenium Spatial Transcriptomics available from the 10x Genomics Datasets homepage (name: "FFPE Human Breast with Custom Add-on Panel", release date: 2023-01-22) were downloaded and converted to the `SpatialData` format.
+The data comprises two samples, one infiltrating ductal carcinoma ("sample 1") and one invasive lobular carcinoma ("sample 2").
+H\&E images from the same dataset were also downloaded, aligned to the respective reference images, and saved to the data object.
+These can be seen in Figure~\ref{fig:spatial_hne}A and Figure~\ref{fig:spatial_hne}B.
+A pattern of tessellating hexagons was generated for each sample, spanning the entire tissue section.
+These hexagons were subsetted to only those overlaying the specimens as visible in the H\&E image, resulting in 10321 hexagons for sample~1 and 8148 for sample~2 respectively.
+These hexagonal masks were used to automatically extract the underlying H\&E crops which were then featurized using `Squidpy`, resulting in 945 features each.
+These features were then pre-processed using `Pycytominer`, resulting in 252 features.
+We then removed batch effect using Harmony.
+UMAP projections of these features are shown in Figure~\ref{fig:spatial_umap}.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=.75\linewidth]{./figs/fig_S1_spatial_umap.png}
+\caption{\label{fig:spatial_umap}{}UMAP projection of the processed morphology features.
+\textbf{Panel A.} Colored by the sample of origin, we see good mixing after the Harmony integration.
+\textbf{Panel B.} Colored by Leiden clustering.}
+\end{figure}
+
+Due to the low cell count in Leiden cluster 10 for sample 2, as shown in Figure~\ref{fig:spatial_c10}B, the cluster was excluded for some analysis steps, and Figure~\ref{fig:spatial_omics}.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=.7\linewidth]{./figs/fig_S4_spatial_c10.png}
+\caption{\label{fig:spatial_c10}{}Barplots showing the total number of cells for each Leiden cluster in sample~1 (\textbf{A}) and sample~2 (\textbf{B}).}
+\end{figure}
+
+\newpage
+
+Then, the gene expression data from the Xenium experiment was used to annotate cell types for both samples.
+For every Leiden cluster, as defined by the Leiden clustering of the joint morphology features from both samples, the cells covered by the respective hexagons were extracted.
+Figure~\ref{fig:spatial_hne} shows the raw H\&E images, the distribution of cell types across these, and the hexagons colored by their respective Leiden cluster.
+Of note is that the H\&E images both seemed to be partially corrupted, resulting in blacked-out sections.
+While the data contains cell segmentation masks for these sections, we filtered out hexagons overlaying these corrupted regions for the morphology featurization.
+
+\begin{figure}[htbp]
+\centering
+\includegraphics[width=.65\linewidth]{./figs/fig_S2_spatial_hne.png}
+\caption{\label{fig:spatial_hne}{}Overview about the spatial distribution of cell types and morphology clusters.
+\textbf{Panel A and B.} H\&E images of both samples.
+\textbf{Panel C and D.} Cell segmentation masks colored by cell type.
+\textbf{Panel E and F.} Hexagons colored by their morphology Leiden cluster.}
+\end{figure}
+
+For every morphology Leiden cluster, we counted all cells whose centroids fell into the respective hexagons and calculated each cell type's share of the total cells per cluster.
+This data is shown in Figure~\ref{fig:spatial_composition}.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=.95\linewidth]{./figs/fig_S3_spatial_composition.png}
+\caption{\label{fig:spatial_composition}{}Barplots showing the per-cluster cell type composition for both samples. Additionally, the average per sample is shown.}
+\end{figure}
+
+\newpage
+
+Based on \citep{Wu2024-zp} we define an immune-hot signature comprising B~cells, dendritic~cells, NK~cells, and T~cells.
+For every Leiden cluster, we aggregated the respective cell type counts of this signature for both samples.
+These are shown in Figure~\ref{fig:spatial_immunehot}B and provide initial evidence for the difference in clusters 0 and 1, despite both being visually mixed as can be seen in Figure~\ref{fig:spatial_immunehot}A.
+
+\begin{figure}[htbp!]
+\centering
+\includegraphics[width=.85\linewidth]{./figs/fig_S5_immunehot.png}
+\caption{\label{fig:spatial_immunehot}{}Potential explanation of the differences between cluster~0 and cluster~1.
+\textbf{Panel A} Hexagons in sample~2 colored by Leiden cluster.
+\textbf{Panel B} Cumulative counts of B~cells, dendritic~cells, NK~cells, and T~cells.}
+\end{figure}
+
 \subsubsection{Acknowledgements}

### 2025-06-10 draft.tex
submission version

@@
 \twocolumn[
-\icmltitle{LIB\_NAME: API-first feature extraction for image-based profiling workflows}
+\icmltitle{cp\_measure: API-first feature extraction for image-based profiling workflows}
 
@@
 While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering machine learning workflows.
-Here we introduce LIB\_NAME, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
-We demonstrate that LIB\_NAME features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem.
-Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how LIB\_NAME enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
+Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
+We demonstrate that cp\_measure features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem.
+Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
 \end{abstract}
@@
 \includesvg[width=.99\linewidth]{./figs/cpmeasure_overview.svg}
-\caption{\label{fig:overview}LIB\_NAME generates features from images by using information in every region of interest ("object").
+\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
 It can featurize the pairwise combination of all the available channels (colours) and objects.
@@
 This motivates the need for robust, accessible tools that efficiently extract interpretable features while integrating seamlessly with modern computational workflows.
-With all this in mind, we developed LIB\_NAME.
+With all this in mind, we developed cp\_measure.
 
@@
 \label{sec:org61842b5}
-The library LIB\_NAME branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
+The library cp\_measure branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
 Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels).
@@
 \label{sec:org09b0cd2}
-We first tested whether LIB\_NAME features match the original CellProfiler features, for a representative set of 300 images.
+We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
 These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature.
-We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to LIB\_NAME for feature extraction.
-Next, we applied LIB\_NAME on these masks and the original images.
-To validate LIB\_NAME, we compared its output directly with CellProfiler's on identical images and masks. 
+We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
+Next, we applied cp\_measure on these masks and the original images.
+To validate cp\_measure, we compared its output directly with CellProfiler's on identical images and masks. 
 For each feature, we computed the $R^2$ value between the two tools' measurements. 
@@
 The few outliers with lower correlations reveal edge cases handled differently by each tool; these will be addressed through comprehensive unit testing.
-These results indicate that LIB\_NAME is not likely to produce massively different profiles than those calculated by CellProfiler.
+These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
 
@@
 \includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:cp_vs_cpmeasure}LIB\_NAME features match their CellProfiler analogues.
-\textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to LIB\_NAME's (y-axis), generated using matching pairs of masks and images.
-\textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing LIB\_NAME to CellProfiler.}
+\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogues.
+\textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
+\textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
 \end{figure}
@@
 
-We demonstrate LIB\_NAME's utility through a biologically relevant task: tracking astrocyte maturation over time. 
-From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, LIB\_NAME extracted morphological features that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-We then trained a gradient-boosting classifier to predict the day of cell differentiation which it was able to do with 86\% accuracy.
-
-SHAP analysis \citep{sundararajanManyShapleyValues2020} (Figure\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
+We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
+From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
+We then trained a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
+SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
 
@@
 \label{sec:org5711d86}
-To demonstrate the broad applicability of LIB\_NAME, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy~\citep{pallaSquidpyScalableFramework2022}.
+To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy~\citep{pallaSquidpyScalableFramework2022}.
 Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.
@@
 Spatial transcriptomics poses unique computational challenges: images routinely contain 500,000+ cells, far exceeding CellProfiler's memory constraints and rigid data structure requirements. 
-LIB\_NAME's modular design enabled us to build a custom workflow integrated with Squidpy~\citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches.
+cp\_measure's modular design enabled us to build a custom workflow integrated with Squidpy~\citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches.
 This architecture handles massive spatial datasets efficiently while bringing morphological profiling directly into the scverse ecosystem \citep{Virshup2023-scverse}, making these features immediately accessible to the single-cell community.
-
-\\
 
@@
 We generated a pattern of tessellating hexagonal masks across both images, of which 18469 were found to overlay the tissue.
-These masks were then featurized with a LIB\_NAME-based workflow: We first preprocessed the features following standard workflows
+These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following standard workflows
 \citep{serranoReproducibleImagebasedProfiling2025}, then removed batch effects between the samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018, Traag2019-fr}.
@@
 Panel C reveals that some clusters have distinct cell-type compositions: For example, cluster~2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.
-Furthermore, comparing clusters~0 and 1 suggests intra-tumour differences, which could be correlated to immune-hotness (see Figure \ref{fig:spatial_immunehot}).
+Furthermore, comparing clusters~0 and 1 suggest intra-tumour differences, which could be correlated to immune-hotness (see Figure \ref{fig:spatial_immunehot}).
 
@@
 
-These results demonstrate how LIB\_NAME can can be used outside traditional microscopy.
+These results demonstrate how cp\_measure can can be used outside traditional microscopy.
 When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes.
@@
 While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop.
-In this work we introduced our new library LIB\_NAME, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
+In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
 It enables simpler automated profiling of microscopy data in short scripts and complex pipelines.
@@
 
-The biologically interpretable features provided by LIB\_NAME complement deep learning features and offer a better mechanistic understanding of the underlying biology.
-A potential workflow would be to use deep learning features to cluster and then use LIB\_NAME features from these clusters.
+The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology.
+A potential workflow would be to use deep learning features to cluster and then use cp\_measure features from these clusters.
 As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.
@@
 While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing \citep{ilhanEffectsProcessParameters2021} and agriculture automation \citep{xuDesignNonDestructiveSeed2024}.
-LIB\_NAME's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
+cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
 
@@
 \label{sec:org5cdbb12}
-We propose to make LIB\_NAME an imported dependency for CellProfiler, offering several benefits.
+We propose to make cp\_measure an imported dependency for CellProfiler, offering several benefits.
 It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
-It would also mean that any changes made in LIB\_NAME propagate to CellProfiler, benefiting the community for which it was originally developed.
+It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.
 
@@
 Once tests are in place, we could add support for just-in-time compiling and GPUs.
-We envision that LIB\_NAME could also be the place to develop and distribute new measurements from the community.
+We envision that cp\_measure could also be the place to develop and distribute new measurements from the community.
 
@@
 \onecolumn
+\section{Acknowledgements}
+\label{sec:acknowledgements}
+
+This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC).
+The authors would like to thank Nodar Gogoberidze, Beth Cimini, Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
+
 \section{Appendix}
@@
 \label{sec:orgbda0ae2}
-The codebase for LIB\_NAME is available on \url{https://anonymous.4open.science/anonymize/feature_extraction_image_profiling}.
+The codebase for cp\_measure is available on \url{https://anonymous.4open.science/anonymize/feature_extraction_image_profiling}.
 All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_paper_feature_extraction_image_profiling/readme.org}.
@@
 A pattern of tessellating hexagons was generated for each sample, spanning the entire tissue section.
-These hexagons were subsetted to only those overlaying the specimens as visible in the H\&E image, resulting in 10321 hexagons for sample~1 and 8148 for sample~2 respectively.
+These hexagons were subselected to only those overlaying the specimens as visible in the H\&E image, resulting in 10321 hexagons for sample~1 and 8148 for sample~2 respectively.
 These hexagonal masks were used to automatically extract the underlying H\&E crops which were then featurized using `Squidpy`, resulting in 945 features each.
@@
 \centering
-\includegraphics[width=.75\linewidth]{./figs/fig_S1_spatial_umap.png}
+\includegraphics[width=.8\linewidth]{./figs/fig_S1_spatial_umap.png}
 \caption{\label{fig:spatial_umap}{}UMAP projection of the processed morphology features.
-\textbf{Panel A.} Colored by the sample of origin, we see good mixing after the Harmony integration.
-\textbf{Panel B.} Colored by Leiden clustering.}
+\textbf{Panel A.} Coloured by sample of origin, we see good mixing after the Harmony integration.
+\textbf{Panel B.} Coloured by Leiden clustering.}
 \end{figure}
 
-Due to the low cell count in Leiden cluster 10 for sample 2, as shown in Figure~\ref{fig:spatial_c10}B, the cluster was excluded for some analysis steps, and Figure~\ref{fig:spatial_omics}.
+Due to the low cell count in Leiden cluster 10 for sample 2, as shown in Figure~\ref{fig:spatial_c10}B the cluster was excluded for some analysis steps.
 
@@
 \includegraphics[width=.7\linewidth]{./figs/fig_S4_spatial_c10.png}
-\caption{\label{fig:spatial_c10}{}Barplots showing the total number of cells for each Leiden cluster in sample~1 (\textbf{A}) and sample~2 (\textbf{B}).}
+\caption{\label{fig:spatial_c10}{}Morphology reveals the composition of tumour subregions.
+\textbf{Panel A.} Spatial map of cell types inferred from transcriptomics.
+\textbf{Panel B.} Morphological subregions obtained by applying Leiden clustering to image-derived feature embeddings.
+\textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type (colour legend as shown).
+The top left quarter of Panel A shows that the morphological approach worked to predict regions while the expression annotations did not.}
 \end{figure}
@@
 For every Leiden cluster, as defined by the Leiden clustering of the joint morphology features from both samples, the cells covered by the respective hexagons were extracted.
-Figure~\ref{fig:spatial_hne} shows the raw H\&E images, the distribution of cell types across these, and the hexagons colored by their respective Leiden cluster.
+Figure~\ref{fig:spatial_hne} shows the raw H\&E images, the distribution of cell types across these, and the hexagons coloured by their respective Leiden cluster.
 Of note is that the H\&E images both seemed to be partially corrupted, resulting in blacked-out sections.
@@
 \caption{\label{fig:spatial_immunehot}{}Potential explanation of the differences between cluster~0 and cluster~1.
-\textbf{Panel A} Hexagons in sample~2 colored by Leiden cluster.
+\textbf{Panel A} Hexagons in sample~2 coloured by Leiden cluster.
 \textbf{Panel B} Cumulative counts of B~cells, dendritic~cells, NK~cells, and T~cells.}
@@
 
-\subsubsection{Acknowledgements}
-\label{sec:acknowledgements}
-
-This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC).
-The authors would like to thank Nodar Gogoberidze, Beth Cimini, Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
 \end{document}

### 2025-06-10 src/astro3d/plot_figures.py
refactor: replace mich urls with zenodo's

@@ -92,19 +92,33 @@ def map_mask_to_image(fpath: Path) -> tuple[str, str] or None:
 
 
 url_hash = {
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/nha_day3_deconv.tar.gz": "6d1fd24d7f9381e1aa4f95dbc9c2c351c7633a3a4500d95b4a6736d40b4100af",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/nha_day3_masks.tar.gz": "d428491b3bc8207bedf77f33afdde5e6542ea6c4f9f63e0cf920f46b6f68f43b",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/nha_day5_deconv.tar.gz": "f7c7f93d1a173441f8b116ca8411df524678d926ed30d490d580d01edd782503",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/nha_day5_masks.tar.gz": "ddbf48a883f03ab89c3adf11cbf4b1e07905f8deb734502f38b684bb009e900b",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/nha_day7_deconv.tar.gz": "9dd69934f306cab31e94df427ee7618d21c32038a62a14ccab15b7d3517eadf7",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/nha_day7_masks.tar.gz": "55aafc7b49f904f711187a85ef0bf5e410d3e6aef4b0b7bf69ef17fce51ee8b6",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/vpa_day3_deconv.tar.gz": "2e1637fd0f240fda6b19ea7d57668ce6b82ae35f697fa2af9b30fc6915d609b5",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/vpa_day3_masks.tar.gz": "e47c7b476f1328f79f4d6825c74706fc7e5504ca3eed7d41caa7199b3d001690",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/vpa_day5_deconv.tar.gz": "08c7278e4c1a61fa77b8c196e79bcd5e222241ff05b492d13a18a9f976c81cf3",
-    "https://www.socr.umich.edu/data/3d-cell-morphometry-data/vpa_day5_masks.tar.gz": "366293f5e68247e927628b6b2e8a8c33338de9a56db9f269e7fb6a0500dbb55b",
[…]

### 2025-06-10 src/astro3d/plot_figures.py
fix(astro3d): remove /content suffix before untarring

@@ -124,7 +124,9 @@ retrieved = list(
     Parallel(n_jobs=-1)(
         delayed(retrieve)(
             url,
-            processor=Untar(extract_dir=Path(url).name.split(".")[0]),
+            processor=Untar(
+                extract_dir=Path(url.removesuffix("/content")).name.split(".")[0]
+            ),
             known_hash=h,
             progressbar=True,
         )

### 2025-06-30 draft.tex
fix(draft): replace — -> --
* draft.tex (subsection{Engineered features in the era of deep learning}):

@@
 Several practical considerations favour engineered features in biological imaging. 
-First, interpretability remains crucial for biological discovery—researchers need to understand which morphological characteristics drive observed phenotypes. 
+First, interpretability remains crucial for biological discovery--researchers need to understand which morphological characteristics drive observed phenotypes. 
 While deep learning features can be powerful predictors, their interpretation requires sophisticated techniques that may not provide the mechanistic insights biologists seek \citep{liChallengesOpportunitiesBioimage2023}.

### 2025-06-30 draft.tex
tweak(draft): replace "-" -> "--"

@@
 Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. 
-A complementary paradigm gaining increasing traction is image-based profiling - quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
+A complementary paradigm gaining increasing traction is image-based profiling -- quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
 While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering machine learning workflows.

### 2025-06-30 draft.tex
fix(draft): — -> --

@@
  applications but complicates direct comparison with established datasets like those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}. 
-Additionally, while all tools necessarily impose input requirements, some enforce rigid organizational constraints—demanding specific directory structures and naming conventions—that require extensive data reorganization before analysis, beyond the standard data type and format specifications.
+Additionally, while all tools necessarily impose input requirements, some enforce rigid organizational constraints--demanding specific directory structures and naming conventions--that require extensive data reorganization before analysis, beyond the standard data type and format specifications.
 Some alternatives also target specific deployment scenarios, with ScaleFex designed primarily for cloud environments, limiting utility for researchers preferring local computation.

### 2025-06-30 draft.tex
draft: remove anonymised info

@@
 \label{sec:orgbda0ae2}
-The codebase for cp\_measure is available on \url{https://anonymous.4open.science/anonymize/feature_extraction_image_profiling}.
-All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://anonymous.4open.science/r/2025_paper_feature_extraction_image_profiling/readme.org}.
+The codebase for cp\_measure is available on \url{https://github.com/afermg/cp_measure}.
+All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025_cpmeasure}.
 %The datasets we produced for this work are available on Zenodo.

### 2025-06-30 appendix.tex
refactor: isolate appendix

--- 2025-06-30 draft.tex
@@
-% Created 2025-05-20 Tue 13:14
-% Intended LaTeX compiler: pdflatex
 \documentclass{article}
+\usepackage[inkscapelatex=false]{svg}
+\usepackage{graphicx}
+\usepackage{hyperref}
+\usepackage{natbib}
+\usepackage{xr}
+\externaldocument{draft}
 
-\input{style/header.tex}
-\usepackage[inkscapelatex=false]{svg}
 \date{}
 \title{}
-\hypersetup{
- pdfauthor={Alan Munoz},
- pdftitle={},
- pdfkeywords={},
- pdfsubject={},
- pdfcreator={Emacs 30.1 (Org mode 9.7.11)}, 
- pdflang={English}}
-\usepackage{natbib}
 \begin{document}
-
-\twocolumn[
-\icmltitle{cp\_measure: API-first feature extraction for image-based profiling workflows}
-
-% It is OKAY to include author information, even for blind
-% submissions: the style file will automatically remove it for you
-% unless you've provided the [accepted] option to the icml2025
-% package.
-
-% List of affiliations: The first argument should be a (short)
-% identifier you will use later to specify author affiliations
-% Academic affiliations should list Department, University, City, Region, Country
-% Industry affiliations should list Company, City, Region, Country
-
-% You can specify symbols, otherwise they are numbered in order.
-% Ideally, you should not use this facility. Affiliations will be numbered
-% in order of appearance and this is the preferred way.
-\icmlsetsymbol{equal}{*}
-
-\begin{icmlauthorlist}
-\icmlauthor{Al\'an F. Mu\~{n}oz}{broad}
-\icmlauthor{Tim Treis}{hh,broad}
-\icmlauthor{Alexandr A. Kalinin}{broad}
-\icmlauthor{Shatavisha Dasgupta}{broad}
-\icmlauthor{Fabian Theis}{hh}
-\icmlauthor{Anne E. Carpenter}{broad}
-\icmlauthor{Shantanu Singh}{broad}
-\end{icmlauthorlist}
-
-\icmlaffiliation{broad}{Broad Institute of MIT and Harvard, United States}
-\icmlaffiliation{hh}{Institute of Computational Biology, Helmholtz Zentrum München, Germany}
-
-\icmlcorrespondingauthor{Al\'an F. Mu\~{n}oz}{amunozgo@broadinstitute.org}
-\icmlcorrespondingauthor{Shantanu Singh}{shantanu@broadinstitute.org}
-
-% You may provide any keywords that you
-% find helpful for describing your paper; these are used to populate
-% the "keywords" metadata in the PDF but will not be shown in the document
-\icmlkeywords{Machine Learning, ICML}
-
-\vskip 0.3in
-]
-
-% this must go after the closing bracket ] following \twocolumn[ ...
-
-% This command actually creates the footnote in the first column
-% listing the affiliations and the copyright notice.
-% The command takes one argument, which is text to display at the start of the footnote.
-% The \icmlEqualContribution command is standard text for equal contribution.
-% Remove it (just {}) if you do not need this facility.
-
-\printAffiliationsAndNotice{}  % leave blank if no need to mention equal contribution
-% \printAffiliationsAndNotice{\icmlEqualContribution} % otherwise use the standard text.
-
-\begin{abstract}
-Biological image analysis has traditionally focused on measuring specific visual properties of interest for cells or other entities. 
-A complementary paradigm gaining increasing traction is image-based profiling -- quantifying many distinct visual features to form comprehensive profiles which may reveal hidden patterns in cellular states, drug responses, and disease mechanisms.
-While current tools like CellProfiler can generate these feature sets, they pose significant barriers to automated and reproducible analyses, hindering machine learning workflows.
-Here we introduce cp\_measure, a Python library that extracts CellProfiler's core measurement capabilities into a modular, API-first tool designed for programmatic feature extraction.
-We demonstrate that cp\_measure features retain high fidelity with CellProfiler features while enabling seamless integration with the scientific Python ecosystem.
-Through applications to 3D astrocyte imaging and spatial transcriptomics, we showcase how cp\_measure enables reproducible, automated image-based profiling pipelines that scale effectively for machine learning applications in computational biology.
-\end{abstract}
-
-\section{Introduction}
-\label{sec:org4c9ba67}
-High throughput screening of biological phenomena via complex modalities, such as RNA sequencing, is prohibitively expensive, therefore microscopy is an efficient first step.
-Through microscopy, modern biologists use a wide array of fluorescence dyes or proteins to observe the location and distribution of cells, organelles, and other components.
-Nowadays, quantification of biological images is standard, with software often identifying regions of interest (such as cells) and extracting features that represent these regions, such as intensity.
-
-Image-based profiling--also termed morphological profiling--is a technique of measuring an extensive suite of morphological features for a population of biological objects, such as cells. 
-These features are fed into statistical or machine learning methods to identify biologically meaningful patterns.
-One of its biggest applications of morphological profiling is drug discovery, where scientists leverage microscopy's low acquisition cost and high throughput to accomplish many goals, such as grouping genes by function, identifying chemical compounds that target a protein, and predicting toxicity of drug candidates \citep{chandrasekaranImagebasedProfilingDrug2021}.
-
-\begin{figure}[htbp]
-\centering
-\includesvg[width=.99\linewidth]{./figs/cpmeasure_overview.svg}
-\caption{\label{fig:overview}cp\_measure generates features from images by using information in every region of interest ("object").
-It can featurize the pairwise combination of all the available channels (colours) and objects.
-The resultant matrices represent the entire experiment and can be studied using statistical, machine, and deep learning methods.}
-\end{figure}
-
-\subsection{The current state of bioimage analysis}
-\label{sec:org8f5b33d}
-The most widely used software for processing high-throughput biological images is CellProfiler \citep{stirlingCellProfiler4Improvements2021}, providing experimental biologists with limited programming expertise an accessible yet powerful toolset.
-By contrast, it may not suit the needs of computational biologists building high throughput pipelines comprised of multiple tools.
-CellProfiler is ideal for creating and iteratively adapting manually-defined workflows; for low-customization high-throughput analyses such as extracting features for image-based profiles or spatial transcriptomics data, however, certain aspects become a hindrance.
-For example, complementing CellProfiler with other image analysis tools is a struggle, as building plug-ins to add functionality is a time and effort-consuming challenge that requires an understanding of its interfaces.
-Lastly, CellProfiler depends on many Python and Java packages, increasing the likelihood of dependency conflicts.
-Containers mitigate this problem, but not without increased complexity and caveats.
-
-Alternative tools can extract features for image-based profiles, such as scikit-image \citep{waltScikitimageImageProcessing2014}, ScaleFex, or SpaCR \citep{comoletHighlyEfficientScalable2024,einarolafssonSpaCr2025}, however, these approaches present trade-offs for different use cases.
-Their independent implementation means they generate different feature sets than CellProfiler, which can be advantageous for novel
- applications but complicates direct comparison with established datasets like those in the Cell Painting Gallery \cite{weisbartCellPaintingGallery2024}. 
-Additionally, while all tools necessarily impose input requirements, some enforce rigid organizational constraints--demanding specific directory structures and naming conventions--that require extensive data reorganization before analysis, beyond the standard data type and format specifications.
-Some alternatives also target specific deployment scenarios, with ScaleFex designed primarily for cloud environments, limiting utility for researchers preferring local computation.
-
-\subsection{Engineered features in the era of deep learning}
-\label{sec:org9dc3dfa}
-
-While computer vision has largely transitioned to deep learning-based feature extraction, biological imaging presents unique challenges that maintain the relevance of engineered features.
-These images differ substantially from natural images in their acquisition methods, visual properties, and analytical objectives, often limiting the effectiveness of standard pre-trained models, even those trained on biological data.
-
-The application of deep learning to feature extraction for biological image analysis shows mixed results.
-When appropriately trained on domain-specific data, deep learning networks can outperform engineered features \cite{lafargeCapturingSingleCellPhenotypic2019,moshkovLearningRepresentationsImagebased2022,chowPredictingDrugPolypharmacology2022,wolfSCANPYLargescaleSinglecell2018}, but not universally \cite{tangMorphologicalProfilingDrug2024,kimSelfsupervisionAdvancesMorphological2023}.
-Effective deep learning feature extractors for biological applications typically require substantial domain-specific datasets and careful model architecture choices, rather than simple transfer from general computer vision models.
-
-Several practical considerations favour engineered features in biological imaging. 
-First, interpretability remains crucial for biological discovery--researchers need to understand which morphological characteristics drive observed phenotypes. 
-While deep learning features can be powerful predictors, their interpretation requires sophisticated techniques that may not provide the mechanistic insights biologists seek \citep{liChallengesOpportunitiesBioimage2023}.
-Engineered features have clear mathematical definitions that directly translate to biological concepts, such as protein co-localisation or nuclear morphology changes \cite{garcia-fossaInterpretingImagebasedProfiles2023}.
-
-Second, many biological imaging laboratories lack the GPU resources, large annotated datasets, or machine learning expertise required for custom model development.
-While foundational models are emerging, visual foundational models can be challenged by the diversity of imaging modalities and experimental conditions \cite{azadFoundationalModelsMedical2023}.
-
-Third, engineered features offer consistency and reproducibility advantages valuable for comparative studies and multi-site collaborations, producing identical results across different computing environments.
-
-Rather than viewing these approaches as competing alternatives, the field increasingly recognizes their complementary roles. 
-This motivates the need for robust, accessible tools that efficiently extract interpretable features while integrating seamlessly with modern computational workflows.
-With all this in mind, we developed cp\_measure.
-
-\section{Extracting interpretable features}
-\label{sec:org61842b5}
-The library cp\_measure branches off the CellProfiler codebase, adapted to include calculations for all features of an image-based profiling pipeline, while removing the user interface and anything else non-essential to the task.
-Measurements are defined as a collection of related features, following CellProfiler's nomenclature, and are categorized based on three kinds of input: One object (producing features such as area and eccentricity), one object + one imaging channel (these can be used to extract such measurements as intensity, texture), one object + multiple channels (e.g., for Manders correlations between intensity values across channels).
-A visual representation of this can be found in Figure~\ref{fig:overview}.
-
-By isolating the feature calculations from the graphical interface and orchestration components, we aimed to make them accessible to the data science community while improving reusability, testability and extensibility.
-This reduces the amount of time and human effort required to integrate featurization into existing and new pipelines, while still retaining the standard set of features available for multiple public datasets.
-
-\subsection{Reproducing CellProfiler measurements}
-\label{sec:org09b0cd2}
-We first tested whether cp\_measure features match the original CellProfiler features, for a representative set of 300 images.
-These correspond to 150 perturbations from the JUMP Cell Painting dataset \citep{chandrasekaranJUMPCellPainting2023}, selecting a representative subset of the most significant phenotypes for each feature.
-We segmented images to delineate the cells and nuclei using CellProfiler, providing object masks (regions of interest) to CellProfiler and to cp\_measure for feature extraction.
-Next, we applied cp\_measure on these masks and the original images.
-To validate cp\_measure, we compared its output directly with CellProfiler's on identical images and masks. 
-For each feature, we computed the $R^2$ value between the two tools' measurements. 
-Over 95\% of features showed $R^2$ > 0.9 (Figure \ref{fig:cp_vs_cpmeasure}), indicating near-perfect agreement. 
-The few outliers with lower correlations reveal edge cases handled differently by each tool; these will be addressed through comprehensive unit testing.
-These results indicate that cp\_measure is not likely to produce massively different profiles than those calculated by CellProfiler.
-
-\begin{figure}[htbp]
-\centering
-\includesvg[width=.9\linewidth]{./figs/jump_r2_examples}
-\caption{\label{fig:cp_vs_cpmeasure}cp\_measure features match their CellProfiler analogues.
-\textbf{Panels A-C.} Representative examples comparing CellProfiler feature values (x-axis) to cp\_measure's (y-axis), generated using matching pairs of masks and images.
-\textbf{Panel D.} \(R^2\) value of a linear fit for each individual feature, comparing cp\_measure to CellProfiler.}
-\end{figure}
-
-\subsection{Results}
-\label{sec:orge5b5c6b}
-\subsubsection{Astrocyte 3D nuclei data}
-\label{sec:org447090b}
-
-We demonstrate cp\_measure's utility through a biologically relevant task: tracking astrocyte maturation over time. 
-From 433 3D images containing 831 astrocyte nuclei \citep{kalininValproicAcidinducedChanges2021}, cp\_measure extracted morphological features  that we preprocessed following standard procedures \citep{caicedoDataanalysisStrategiesImagebased2017}.
-We then trained a Gradient Boosting classifier to predict the day of cell differentiation with 87\% accuracy. 
-SHAP analysis \citep{sundararajanManyShapleyValues2020} (\ref{fig:astrocytes}) revealed that nuclear minor axis length was the most predictive feature: nuclei became progressively wider during maturation, a phenotype consistent with known astrocyte biology.
-
-\begin{figure}[htbp]
-\centering
-\includesvg[width=.9\linewidth]{./figs/example_shap.svg}
-\caption{\label{fig:astrocytes}\textbf{Panels A and B.} Example pair of astrocyte nuclei image and masks.
-The 3D images were projected over the z-axis, taking the maximum value across the z-stack.
-\textbf{Panel C.} SHAP values of the most important features for classifying the day of differentiation (out of three days).
-Each point represents a single 3D image containing multiple nuclei, the magnitude of SHAP values indicate how important those features are for the classifier.
-Features are ranked by importance, with individual SHAP values shown for the top features and the summed impact shown for the remaining 267 features shown as a single aggregated value.
-Shape features, such as the minor axis length, refer to the nucleus, while intensity features refer to the DNA staining.
-The test data accuracy is shown in bold.}
-\end{figure}
-
-\subsubsection{Spatial transcriptomics}
-\label{sec:org5711d86}
-To demonstrate the broad applicability of cp\_measure, we developed a morphology featurizer for the popular spatial omics analysis toolbox Squidpy~\citep{pallaSquidpyScalableFramework2022}.
-Many of the recently developed spatial omics analysis technologies generate, in addition to their target omics layer, imaging data, usually H\&E (hematoxylin and eosin) stained or fluorescently labelled.
-
-Spatial transcriptomics poses unique computational challenges: images routinely contain 500,000+ cells, far exceeding CellProfiler's memory constraints and rigid data structure requirements. 
-cp\_measure's modular design enabled us to build a custom workflow integrated with Squidpy~\citep{pallaSquidpyScalableFramework2022} that leverages the SpatialData format \citep{Marconato2024-SpatialData} to process cells in streaming parallel batches.
-This architecture handles massive spatial datasets efficiently while bringing morphological profiling directly into the scverse ecosystem \citep{Virshup2023-scverse}, making these features immediately accessible to the single-cell community.
-
-To illustrate the value of this workflow, we used it to identify morphologically similar regions within two human breast cancer sections profiled with the spatial transcriptomics technology Xenium \citep{10x-Genomics2023-el}.
-We generated a pattern of tessellating hexagonal masks across both images, of which 18469 were found to overlay the tissue.
-These masks were then featurized with a cp\_measure-based workflow: We first preprocessed the features following standard workflows
-\citep{serranoReproducibleImagebasedProfiling2025}, then removed batch effects between the samples using Harmony \citep{Korsunsky2019-Harmony}, and finally obtained the Leiden clusters via Scanpy \citep{wolfSCANPYLargescaleSinglecell2018, Traag2019-fr}.
-
-For each cluster, we extracted the cells within its hexagons and calculated the respective cell type proportions.
-A visual comparison of panels A and B in Figure~\ref{fig:spatial_omics} shows that morphology-only clustering largely recapitulates the underlying tissue architecture.
-Panel C reveals that some clusters have distinct cell-type compositions: For example, cluster~2 is dominated by fibroblasts, whereas adipocytes appear almost exclusively in clusters 5-7.
-Furthermore, comparing clusters~0 and 1 suggest intra-tumour differences, which could be correlated to immune-hotness (see Figure \ref{fig:spatial_immunehot}).
-
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.99\linewidth]{./figs/fig_4_spatial.png}
-\caption{\label{fig:spatial_omics}{}Morphology reveals the composition of tumour subregions.
-\textbf{Panel A.} Spatial map of cell types inferred from transcriptomics.
-\textbf{Panel B.} Morphological subregions obtained by Leiden clustering the image-derived feature embeddings.
-\textbf{Panel C.} Bar chart showing, for each morphological cluster, the fraction of cells assigned to each cell type.
-Clusters 5-7 were merged due to similarity, cluster~10 removed due to low cell count (see Figure \ref{fig:spatial_c10}B).}
-\end{figure}
-
-These results demonstrate how cp\_measure can can be used outside traditional microscopy.
-When applied on extremely large images it produces biologically meaningful morphological clusters that validate and complement transcriptomic phenotypes.
-By embedding this capability within the scverse ecosystem via Squidpy, we enable researchers to seamlessly add morphology to their spatial omics workflows, unlocking new insights into tissue organization.
-
-\section{Discussion}
-\label{sec:orgf37b369}
-While point-and-click interfaces open the world of image analysis to many researchers, they are not as effective for computational workflows with no human-in-the-loop.
-In this work we introduced our new library cp\_measure, which calculates a set of widely used engineered features relevant to whole images and to regions of interest (object masks).
-It enables simpler automated profiling of microscopy data in short scripts and complex pipelines.
-The modularity it provides facilitates pipelines with better scaling capabilities for high-content microscopy, with or without cloud infrastructure.
-
-The biologically interpretable features provided by cp\_measure complement deep learning features and offer a better mechanistic understanding of the underlying biology.
-A potential workflow would be to use deep learning features to cluster and then use cp\_measure features from these clusters.
-As a whole, when used in tandem with generalist image-processing tools, such as Cellpose for segmentation \cite{stringerCellposeGeneralistAlgorithm2021}, machine and deep learning workflows can be streamlined.
-
-While developed for biology, engineered morphological features have proven valuable far beyond their original domain--from analysing microplastics in environmental samples \citep{ideharaExploringNileRed2025} to quality control in manufacturing \citep{ilhanEffectsProcessParameters2021} and agriculture automation \citep{xuDesignNonDestructiveSeed2024}.
-cp\_measure's lightweight, API-first design makes these powerful features accessible to any field working with image data, removing the barriers of biological-specific interfaces.
-
-\section{Future work}
-\label{sec:org5cdbb12}
-We propose to make cp\_measure an imported dependency for CellProfiler, offering several benefits.
-It would ensure that the results from pipelines built with either tool will always be comparable, while also providing the opportunity to formalize the inputs and outputs of all measurements.
-It would also mean that any changes made in cp\_measure propagate to CellProfiler, benefiting the community for which it was originally developed.
-
-We also plan to develop a comprehensive test suite to guarantee mathematical correctness, which currently CellProfiler itself is lacking.
-A comprehensive test suite would enable more confident and expedient optimization for the most compute-consuming features, such as granularity, by providing rapid iteration capabilities.
-Once tests are in place, we could add support for just-in-time compiling and GPUs.
-We envision that cp\_measure could also be the place to develop and distribute new measurements from the community.
-
-\bibliographystyle{icml2025}
-\bibliography{bibliography}
-
-\onecolumn
-\section{Acknowledgements}
-\label{sec:acknowledgements}
-
-This work was supported by GSK and the National Institutes of Health, grant (R35 GM122547 to AEC).
-The authors would like to thank Nodar Gogoberidze, Beth Cimini, Minh Doan and Eliot McKinley for their valuable feedback and discussions during the course of this project.
-
 \section{Appendix}
 
-\appendix
 \setcounter{figure}{0}
@@
 
+\bibliographystyle{icml2025}
+\bibliography{bibliography}
+
 \end{document}

### 2025-06-30 draft.tex
refactor: isolate appendix

@@
 \documentclass{article}
+\usepackage{xr}
+\externaldocument{appendix}
 
@@
 
-\section{Appendix}
+%% \section{Appendix}
 
-\appendix
-\setcounter{figure}{0}
-% redefine \thefigure to prefix “A” and then use arabic
-\renewcommand{\thefigure}{A\arabic{figure}}
+%% \appendix
+%% \setcounter{figure}{0}
+%% % redefine \thefigure to prefix “A” and then use arabic
+%% \renewcommand{\thefigure}{A\arabic{figure}}
 
-% fix appendix section numbering
-\setcounter{section}{5}
-\renewcommand{\thesection}{\arabic{section}}
-\renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
-\renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
+%% % fix appendix section numbering
+%% \setcounter{section}{5}
+%% \renewcommand{\thesection}{\arabic{section}}
+%% \renewcommand{\thesubsection}{\thesection.\arabic{subsection}}
+%% \renewcommand{\thesubsubsection}{\thesubsection.\arabic{subsubsection}}
 
-\label{sec:orgdd18dd8}
-\subsection{Methods}
-\label{sec:orgb3e9382}
-\subsubsection{Data and software}
-\label{sec:orgbda0ae2}
-The codebase for cp\_measure is available on \url{https://github.com/afermg/cp_measure}.
+%% \label{sec:orgdd18dd8}
+%% \subsection{Methods}
+%% \label{sec:orgb3e9382}
+%% \subsubsection{Data and software}
+%% \label{sec:orgbda0ae2}
+%% The codebase for cp\_measure is available on \url{https://github.com/afermg/cp_measure}.
 All code to reproduce the analyses and figures, alongside links to the original data, is available on the GitHub repository \url{https://github.com/afermg/2025_cpmeasure}.
-%The datasets we produced for this work are available on Zenodo.
+%% %The datasets we produced for this work are available on Zenodo.
 The selection of samples from JUMP and the images can be found on \url{https://zenodo.org/records/15390631/latest}.
 The image-based profiles produced with Cellprofiler and used for the comparisons in Figure 2 can be on \url{https://zenodo.org/records/15505477/latest}.
-%
+%% %
 
-\subsubsection{Details on the spatial use case}
-\label{sec:spatial_extra}
+%% \subsubsection{Details on the spatial use case}
+%% \label{sec:spatial_extra}
 
-The following figures show the processing steps for the spatial data used to generate Figure~\ref{fig:spatial_omics} in more detail.
+%% The following figures show the processing steps for the spatial data used to generate Figure~\ref{fig:spatial_omics} in more detail.
 Raw data from two Xenium Spatial Transcriptomics available from the 10x Genomics Datasets homepage (name: "FFPE Human Breast with Custom Add-on Panel", release date: 2023-01-22) were downloaded and converted to the `SpatialData` format.
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.8\linewidth]{./figs/fig_S1_spatial_umap.png}
-\caption{\label{fig:spatial_umap}{}UMAP projection of the processed morphology features.
+%% \begin{figure}[htbp]
+%% \centering
+%% \includegraphics[width=.8\linewidth]{./figs/fig_S1_spatial_umap.png}
+%% \caption{\label{fig:spatial_umap}{}UMAP projection of the processed morphology features.
 \textbf{Panel A.} Coloured by sample of origin, we see good mixing after the Harmony integration.
 \textbf{Panel B.} Coloured by Leiden clustering.}
-\end{figure}
+%% \end{figure}
 
-Due to the low cell count in Leiden cluster 10 for sample 2, as shown in Figure~\ref{fig:spatial_c10}B the cluster was excluded for some analysis steps.
+%% Due to the low cell count in Leiden cluster 10 for sample 2, as shown in Figure~\ref{fig:spatial_c10}B the cluster was excluded for some analysis steps.
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.7\linewidth]{./figs/fig_S4_spatial_c10.png}
-\caption{\label{fig:spatial_c10}{}Morphology reveals the composition of tumour subregions.
+%% \begin{figure}[htbp]
+%% \centering
+%% \includegraphics[width=.7\linewidth]{./figs/fig_S4_spatial_c10.png}
+%% \caption{\label{fig:spatial_c10}{}Morphology reveals the composition of tumour subregions.
 \textbf{Panel A.} Spatial map of cell types inferred from transcriptomics.
@@
 The top left quarter of Panel A shows that the morphological approach worked to predict regions while the expression annotations did not.}
-\end{figure}
+%% \end{figure}
 
-\newpage
+%% \newpage
 
-Then, the gene expression data from the Xenium experiment was used to annotate cell types for both samples.
+%% Then, the gene expression data from the Xenium experiment was used to annotate cell types for both samples.
 For every Leiden cluster, as defined by the Leiden clustering of the joint morphology features from both samples, the cells covered by the respective hexagons were extracted.
@@
 
-\begin{figure}[htbp]
-\centering
-\includegraphics[width=.65\linewidth]{./figs/fig_S2_spatial_hne.png}
-\caption{\label{fig:spatial_hne}{}Overview about the spatial distribution of cell types and morphology clusters.
+%% \begin{figure}[htbp]
+%% \centering
+%% \includegraphics[width=.65\linewidth]{./figs/fig_S2_spatial_hne.png}
+%% \caption{\label{fig:spatial_hne}{}Overview about the spatial distribution of cell types and morphology clusters.
 \textbf{Panel A and B.} H\&E images of both samples.
@@
 \textbf{Panel E and F.} Hexagons colored by their morphology Leiden cluster.}
-\end{figure}
+%% \end{figure}
 
-For every morphology Leiden cluster, we counted all cells whose centroids fell into the respective hexagons and calculated each cell type's share of the total cells per cluster.
+%% For every morphology Leiden cluster, we counted all cells whose centroids fell into the respective hexagons and calculated each cell type's share of the total cells per cluster.
 This data is shown in Figure~\ref{fig:spatial_composition}.
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=.95\linewidth]{./figs/fig_S3_spatial_composition.png}
-\caption{\label{fig:spatial_composition}{}Barplots showing the per-cluster cell type composition for both samples. Additionally, the average per sample is shown.}
-\end{figure}
+%% \begin{figure}[htbp!]
+%% \centering
+%% \includegraphics[width=.95\linewidth]{./figs/fig_S3_spatial_composition.png}
+%% \caption{\label{fig:spatial_composition}{}Barplots showing the per-cluster cell type composition for both samples. Additionally, the average per sample is shown.}
+%% \end{figure}
 
-\newpage
+%% \newpage
 
-Based on \citep{Wu2024-zp} we define an immune-hot signature comprising B~cells, dendritic~cells, NK~cells, and T~cells.
+%% Based on \citep{Wu2024-zp} we define an immune-hot signature comprising B~cells, dendritic~cells, NK~cells, and T~cells.
 For every Leiden cluster, we aggregated the respective cell type counts of this signature for both samples.
@@
 
-\begin{figure}[htbp!]
-\centering
-\includegraphics[width=.85\linewidth]{./figs/fig_S5_immunehot.png}
-\caption{\label{fig:spatial_immunehot}{}Potential explanation of the differences between cluster~0 and cluster~1.
+%% \begin{figure}[htbp!]
+%% \centering
+%% \includegraphics[width=.85\linewidth]{./figs/fig_S5_immunehot.png}
+%% \caption{\label{fig:spatial_immunehot}{}Potential explanation of the differences between cluster~0 and cluster~1.
 \textbf{Panel A} Hexagons in sample~2 coloured by Leiden cluster.
 \textbf{Panel B} Cumulative counts of B~cells, dendritic~cells, NK~cells, and T~cells.}
-\end{figure}
+%% \end{figure}
 
